### 2014-06-01 ijlyttle/first_pass.Rmd
adds Rmd file where the data are loaded

# First pass through the data

This will be a stream-of-concoiusness walk through the data, a first pass to see what I can see.

## Libraries

At some point, we should make sure we agree on which versions of which packages we propose to use. As a starting point, I can show what I am using:

```{r load_libraries, message=FALSE}
library(stringr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggvis)
```

Here's my versions:

```{r session_info}
sessionInfo()
```

I suspect that for publication, `ggplot2` will be more capable than `ggvis`.

# Load the data

```{r load_data}
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){

  template <- c("2012.rda", "2012dict.rda")

  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path("..", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
rm(fn_build, sets)
```

The dictionaries are stored as named vectors - I would prefer data frames so that I can use the RStudio GUI to examine them.

```{r modify_dictionaries}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict)

# clean
rm(fn_make_df)
rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

Looking at the number of variables in each of the data frames, I wish I knew how to use ggobi...

### 2014-06-09 ijlyttle/first_pass.Rmd
gets as far as generating Gini coefficients

@@
 library(dplyr)
+library(reshape2)
 library(lubridate)
@@
 Looking at the number of variables in each of the data frames, I wish I knew how to use ggobi...
+
+One of the tracks to investigate is inequality. I see there is a package, `ineq` on CRAN, that claims to have a Gini-coefficient function, so that seems like a good place to start.
+
+```{r ineq}
+library(ineq)
+```
+
+Let's get an idea, for each country and subject, an estimate of the inequality of outcomes.
+
+NOTE: `dplyr` seems to choke on grouping of data.frames with NA, hence the `filter()`.
+
+Also: is there a problem the help text for `dplyr::grouped_df`? The description of `drop` seems backwards to me.
+
+```{r twiddle}
+df_twiddle <- student2012 %>%
+  select(country = CNT, age = ST06Q01,
+         math = PV1MATH, reading = PV1READ, science = PV1SCIE) %>%
+  melt(id.vars = c("country", "age"), 
+       variable.name = "subject", 
+       value.name = "score")  %>%
+  do(filter(., complete.cases(.))) %>%
+  dplyr::group_by(country, subject) %>%
+  summarize(n = n(), gini_coeff = ineq(score, type="Gini")) %>%
+  arrange(as.character(country))
+```

### 2014-06-10 dicook/Look.Rmd
start on looking at data

PISA
========================================================

Working from Ian's code to read in the data - it is beautiful!

```{r load_libraries, message=FALSE}
library(stringr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
```

```{r load_data}
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){

  template <- c("2012.rda", "2012dict.rda")

  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path(".", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
#rm(fn_build, sets)
```

```{r modify_dictionaries}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict)

# clean
#rm(fn_make_df)
#rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

Do some basic summaries, to see what's in the data.

```{r whatsthere}
# Schools
dim(dict_school2012)
table(school2012$CNT)
dim(dict_student2012)
length(table(student2012$STIDSTD))
subset(student2012, STIDSTD=="09631")
dict_student2012[,2]
table(student2012$ST04Q01)
length(scoredItem2012dict)
length(item2012dict)
length(parent2012dict)
```

Dizzyingly many variables! (GGobi won't help) Need to map out a plan of attack, questions of interest, and variables to create.

- It will be good to include map information, so we can make maps

- Variables need to be re-organized, many columns represent one level of a categorical variable

- What questions would we have about the data?

- Which variable has the score?

Now create a few variables, to provide examples, and make a few pictures.

```{r usa}
school.usa <- subset(school2012, CNT=="United States of America")
dim(school.usa)
school.usa$GenderPropBoys <- school.usa[,15]/(school.usa[,15]+school.usa[,16])
summary(school.usa$GenderPropBoys)
qplot(GenderPropBoys, data=school.usa)
table(school.usa$SC03Q01)
qplot(GenderPropBoys, SC03Q01, data=school.usa)
summarise(group_by(school.usa, SC03Q01), median(GenderPropBoys, na.rm=T))
qplot(GenderPropBoys, PROPMATH, data=school.usa, ylim=c(0,0.3), xlim=c(0.25, 0.75)) # Ignore extremese
table(school.usa$SC01Q01)
summary(school.usa[,290])
```

Merge data with maps, so it can be plotted on a map. Interesting set of countries in the study - or more what countries are not included! This indicates some inequities.

```{r maps}
student2012.sub <- student2012[,c(1,12)]
student2012.gender <- summarise(group_by(student2012.sub, CNT), PropBoys=length(ST04Q01[ST04Q01=="Male"])/length(ST04Q01))
student2012.gender$CNT <- factor(student2012.gender$CNT, levels=levels(student2012.gender$CNT)[order(student2012.gender$PropBoys)])
qplot(CNT, PropBoys, data=student2012.gender) + coord_flip()
map.CNT <- map()$names
world <- map_data("world")
student2012.gender$CNT <- as.character(student2012.gender$CNT)
student2012.gender$CNT[student2012.gender$CNT=="United States of America"] <- "USA"
student2012.gender.map <- merge(student2012.gender, world, by.x="CNT", by.y="region")
qplot(long, lat, data=student2012.gender.map, group=group, order=order, geom="polygon", fill=PropBoys) + coord_map()
```

### 2014-06-11 alex/initial.Rmd
average hours studied and average scores per country

OECD
========================================================
```{r load, cache=TRUE}
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)
```

Average test scores: each student has 5 scores for each area.  
```{r avg, cache=TRUE}
avg_test = function(subj) {
  apply(student2012[, paste0("PV", 1:5, subj)], 1, mean)
}
subj = c("MATH", "MACC", "MACQ", "MACS", "MACU", "MAPE", "MAPF", "MAPI", "READ", "SCIE")

#rank avg math scores by country
student2012$math = avg_test("MATH")

math_scores = student2012 %>% 
    select(CNT, math) %>% 
    group_by(CNT) %>% 
    summarise(m_math = mean(math)) %>% 
    arrange(m_math)

#rank avg reading scores by country
student2012$read = avg_test("READ")

read_scores = student2012 %>% 
    select(CNT, read) %>% 
    group_by(CNT) %>% 
    summarise(m_read = mean(read)) %>% 
    arrange(m_read)

#rank avg science scores by country
student2012$sci = avg_test("SCIE")

sci_scores = student2012 %>% 
    select(CNT, sci) %>% 
    group_by(CNT) %>% 
    summarise(m_sci = mean(sci)) %>% 
    arrange(m_sci)

print(math_scores)
print(read_scores)
print(sci_scores)
```

I've heard that Chinese students on average spend the most time studying and Finnish students spend the least amount of time.  We can see that Finnish students study very little outside of school.
```{r hours, cache=TRUE}
#removing the stupid attributes from atomic vectors
student2012$MMINS = as.numeric(student2012$MMINS)
student2012$LMINS = as.numeric(student2012$LMINS)
student2012$SMINS = as.numeric(student2012$SMINS)
student2012$ST57Q01 = as.numeric(student2012$ST57Q01)
student2012$ST57Q02 = as.numeric(student2012$ST57Q02)
student2012$ST57Q03 = as.numeric(student2012$ST57Q03)
student2012$ST57Q04 = as.numeric(student2012$ST57Q04)
student2012$ST57Q05 = as.numeric(student2012$ST57Q05)

#all time related vars
dict_student2012[grep("time", dict_student2012[,2], ignore.case=TRUE), ]

OUTHOURS = student2012 %>%
    select(CNT, LMINS, MMINS, OUTHOURS, SMINS, ST57Q01, ST57Q02, ST57Q03, ST57Q04, ST57Q05) %>%
    group_by(CNT) %>%
    summarise(m_OUTHOURS = mean(OUTHOURS, na.rm=T)) %>%
    arrange(m_OUTHOURS)
print(OUTHOURS)
```

We can calculate score per hour to normalize test scores against how much students actually study.
```{r norm, cache=TRUE}
OUTHOURS = OUTHOURS %>% 
  group_by(CNT) %>% 
  left_join(math_scores) %>% 
  left_join(read_scores) %>% 
  left_join(sci_scores) %>%
  mutate(math_hour = m_math / m_OUTHOURS) %>%
  mutate(read_hour = m_read / m_OUTHOURS) %>%
  mutate(sci_hour = m_sci / m_OUTHOURS)

print(OUTHOURS[, c(1,6:8)])
```

### 2014-06-12 alex/initial.Rmd
found out student ids are not unique

@@
 ```
+
+Let's see if parent income has any effect on student performance
+```{r parents, cache=TRUE}
+library(ggvis)
+students = student2012 %>% select(CNT, STIDSTD, math, read, sci)
+parents = parent2012 %>% select(CNT, STIDSTD, PA07Q01,
+                                PA03Q01, PA03Q02, PA03Q03, PA03Q04,
+                                PA05Q01, PA05Q02, PA05Q03, PA05Q04)
+student_parent = students %>% left_join(parents, by = "STIDSTD") %>% filter(!is.na(PA07Q01)) %>% arrange(desc(math))
+head(student_parent, 50)
+```
+
+Apparently student IDs are not unique!  Students from different countries can share IDs.

### 2014-06-12 ijlyttle/first_pass.Rmd
ijl first set of plots

@@
 library(dplyr)
-library(reshape2)
+library(tidyr)
 library(lubridate)
@@
          math = PV1MATH, reading = PV1READ, science = PV1SCIE) %>%
-  melt(id.vars = c("country", "age"), 
-       variable.name = "subject", 
-       value.name = "score")  %>%
+  gather(key = subject, value = score, -country, -age) %>%
   do(filter(., complete.cases(.))) %>%
   dplyr::group_by(country, subject) %>%
-  summarize(n = n(), gini_coeff = ineq(score, type="Gini")) %>%
+  summarize(
+    n = n(), 
+    gini_coeff = ineq(score, type="Gini"), 
+    mean = mean(score), 
+    median = median(score)) %>%
   arrange(as.character(country))
 ```
+
+"Alabama first" works to look for a specific country, but lets arrange by something more meaningful.
+
+Let's get a mean of coefficents over all subjects, to help us order things.
+
+```{r by_country}
+df_twiddle_mean <- df_twiddle %>%
+  group_by(country) %>%
+  summarize(
+    gini_coeff = mean(gini_coeff),
+    mean = mean(mean),
+    median = mean(median)
+  )
+```
+
+```{r arrange}
+my_levels <- 
+  df_twiddle_mean$country[rev(order(df_twiddle_mean$gini_coeff))] %>%
+  as.character
+
+df_twiddle$country <- factor(
+  df_twiddle$country,
+  levels = my_levels
+)
+
+head(my_levels)
+
+tail(my_levels)
+```
+
+I know I should figure out the `ggmap` package (I'll look at Di's code), but in the meantinme:
+
+```{r vis_ineq_median}
+ggplot(
+  df_twiddle %>% filter(subject != "mean_subject"), 
+  aes(x = gini_coeff, y = median, color = subject, size = n)
+) + 
+geom_point(alpha = 0.75) + 
+scale_size_area()
+```
+
+This is print, so using ggvis and toolips is out of the question...
+
+There appear two clusters - it will be interesting to see if there is a thread to bind them.
+
+```{r vis_countries}
+ggplot(
+  df_twiddle,
+  aes(x = gini_coeff, y = country, color = subject, size = n)
+) + 
+geom_point() + 
+scale_size_area()
+
+ggplot(
+  df_twiddle,
+  aes(x = median, y = country, color = subject, size = n)
+) + 
+geom_point() + 
+scale_size_area()
+```
+
+Hopefully enough to get started.

### 2014-06-19 alex/2020.Rmd
looking a top20 and bottom20

OUTHOURS vs Scores and 20-20 ratio
========================================================

```{r load, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

```{r avg, cache=TRUE, echo=FALSE}
avg_test = function(subj) {
  apply(student2012[, paste0("PV", 1:5, subj)], 1, mean)
}

student2012$math = avg_test("MATH")
student2012$read = avg_test("READ")
student2012$sci = avg_test("SCIE")
```

While looking at my favorite country with a city named Tirana, I noticed that something strange was happening.  No other country appears to have this.
```{r albania, cache=TRUE}
#peak at 54 hours
qplot(OUTHOURS, data = subset(student2012, CNT == "Albania"))
```

Not much correlation between scores and out of school study hours
```{r outhours, cache=TRUE, tidy=FALSE}
qplot(x = OUTHOURS, y = math, 
      data = student2012 %>% 
        select(OUTHOURS, math) %>% 
        filter(!is.na(OUTHOURS)), 
      alpha = I(0.1)) + stat_smooth()
qplot(x = OUTHOURS, y = read, 
      data = student2012 %>% 
        select(OUTHOURS, read) %>% 
        filter(!is.na(OUTHOURS)), 
      alpha = I(0.1) ) + stat_smooth()
qplot(x = OUTHOURS, y = sci, 
      data = student2012 %>% 
        select(OUTHOURS, sci) %>% 
        filter(!is.na(OUTHOURS)), 
      alpha = I(0.1) ) + stat_smooth()
```

```{r 2020, cache=TRUE, echo=FALSE, eval=FALSE}
##UNUSED
#cutoffs for top 20/bottom 20 for each country
best_worst = student2012 %>% 
  select(CNT, math, read, sci, OUTHOURS) %>% 
  group_by(CNT) %>% 
  summarise(math_low = quantile(math, probs = seq(0, 1, 0.2))[2],
            math_high = quantile(math, probs = seq(0, 1, 0.2))[5],
            read_low = quantile(read, probs = seq(0, 1, 0.2))[2],
            read_high = quantile(read, probs = seq(0, 1, 0.2))[5],
            sci_low = quantile(sci, probs = seq(0, 1, 0.2))[2],
            sci_high = quantile(sci, probs = seq(0, 1, 0.2))[5])
head(best_worst)
```

Mean scores in the top 20 and bottom 20 percent for each nation
```{r 2020_1, cache=TRUE, tidy=FALSE}
library(reshape2)
best_worst = student2012 %>% 
  select(CNT, math, read, sci) %>% 
  group_by(CNT) %>% 
  summarise(math_high = mean(math[math > quantile(math, probs = seq(0, 1, 0.2))[5]]),
            math_low = mean(math[math < quantile(math, probs = seq(0, 1, 0.2))[2]]),
            sci_high = mean(sci[sci > quantile(sci, probs = seq(0, 1, 0.2))[5]]),
            sci_low = mean(sci[sci < quantile(sci, probs = seq(0, 1, 0.2))[2]]),
            read_high = mean(read[read > quantile(read, probs = seq(0, 1, 0.2))[5]]),
            read_low = mean(read[read < quantile(read, probs = seq(0, 1, 0.2))[2]])) %>%
  mutate(math_gap = math_high - math_low,
         sci_gap = sci_high - sci_low,
         read_gap = read_high - read_low) %>%
  mutate(avg_gap = (math_gap + sci_gap + read_gap)/3 ) %>%
  arrange(desc(avg_gap))
head(best_worst)

best_worst$CNT = factor(best_worst$CNT, 
                        levels = best_worst$CNT[order(best_worst$avg_gap)])
qplot(x = avg_gap, y = CNT, data = best_worst)

qplot(x = value, y = CNT, data = best_worst %>% 
        select(CNT, math_gap, read_gap, sci_gap) %>% 
        melt(id.vars = "CNT"), color = variable)
```

Compare differences in the top 20 and bottom 20 math scores for each nation
```{r math, cache=TRUE}
student2012 %>%
  select(CNT, math, OUTHOURS) %>%
  group_by(CNT) %>%

  filter(math < quantile(math, probs = seq(0, 1, 0.2))[2] | 
         math > quantile(math, probs = seq(0, 1, 0.2))[5])
```

### 2014-06-19 alex/initial.Rmd
looking a top20 and bottom20

--- 2014-06-11 alex/initial.Rmd
@@
 ```
+
+Let's see if parent income has any effect on student performance.  No surprise here.
+```{r parents, cache=TRUE}
+library(ggvis)
+students = student2012 %>% select(CNT, STIDSTD, math, read, sci)
+parents = parent2012 %>% select(CNT, STIDSTD, PA07Q01,
+                                PA03Q01, PA03Q02, PA03Q03, PA03Q04,
+                                PA05Q01, PA05Q02, PA05Q03, PA05Q04)
+student_parent = students %>% left_join(parents, by = c("STIDSTD", "CNT"))
+student_parent = student_parent[!is.na(student_parent$PA07Q01), ]
+qplot(x = PA07Q01, y = math, geom="boxplot", data = student_parent)
+```
+
+```{r time, cache=TRUE}
+time_vars = c(paste0("ST57Q0", 1:6), "LMINS", "MMINS", "SMINS", "OUTHOURS", paste0("ST69Q0", 1:3), paste0("ST70Q0", 1:3), "ST71Q01")
+
+#  ST57Q01                                       Out-of-School Study Time - Homework
+#  ST57Q02                                Out-of-School Study Time - Guided Homework
+#  ST57Q03                                 Out-of-School Study Time - Personal Tutor
+#  ST57Q04                             Out-of-School Study Time - Commercial Company
+#  ST57Q05                                    Out-of-School Study Time - With Parent
+#  ST57Q06                                       Out-of-School Study Time - Computer
+#172 ST69Q01	Min in <class period> - <test lang>
+#173	ST69Q02	Min in <class period> - <Maths>
+#174	ST69Q03	Min in <class period> - <Science>
+#175	ST70Q01	No of <class period> p/wk - <test lang>
+#176	ST70Q02	No of <class period> p/wk - <Maths>
+#177	ST70Q03	No of <class period> p/wk - <Science>
+#178	ST71Q01	No of ALL <class period> a week
+
+```

### 2014-06-19 alex/2020.Rmd
some stuff

@@
 While looking at my favorite country with a city named Tirana, I noticed that something strange was happening.  No other country appears to have this.
-```{r albania, cache=TRUE}
+```{r albania, echo=FALSE, cache=TRUE}
+library(ggplot2)
 #peak at 54 hours
@@
 
-Compare differences in the top 20 and bottom 20 math scores for each nation
+Compare differences in the top 20 and bottom 20 math scores for each nation.
 ```{r math, cache=TRUE}
-student2012 %>%
-  select(CNT, math, OUTHOURS) %>%
+student2012$ST69Q02 = as.numeric(student2012$ST69Q02)
+student2012$ST70Q02 = as.numeric(student2012$ST70Q02)
+
+math2020 = student2012 %>%
+            select(CNT, math, OUTHOURS, ST69Q02, ST70Q02, #minutes studying, class periods
+                   ST44Q01, ST44Q03, ST44Q04,ST44Q05, ST44Q07, ST44Q08, #attributions to failure
+                   ST15Q01, ST19Q01) %>% #job status for parents
             group_by(CNT) %>%
-
+            mutate(top20 = math > quantile(math, probs = seq(0, 1, 0.2))[5]) %>%
             filter(math < quantile(math, probs = seq(0, 1, 0.2))[2] | 
@@
 ```
+
+```{r math2, eval=TRUE}
+library(ggplot2)
+
+#Attributions to Failure - Not Good at Maths Problems
+qplot(x = ST44Q01, fill = top20, data = subset(math2020, !is.na(ST44Q01)), geom = "bar", position = "fill")
+
+#Attributions to Failure - Teacher Did Not Explain Well
+qplot(x = ST44Q03, fill = top20, data = subset(math2020, !is.na(ST44Q03)), geom = "bar", position = "fill")
+
+#Attributions to Failure - Bad Guesses
+qplot(x = ST44Q04, fill = top20, data = subset(math2020, !is.na(ST44Q04)), geom = "bar", position = "fill")
+
+#Attributions to Failure - Material Too Hard
+qplot(x = ST44Q05, fill = top20, data = subset(math2020, !is.na(ST44Q05)), geom = "bar", position = "fill")
+
+#Attributions to Failure - Teacher Didnt Get Students Interested
+qplot(x = ST44Q07, fill = top20, data = subset(math2020, !is.na(ST44Q07)), geom = "bar", position = "fill")
+
+#Attributions to Failure - Unlucky
+qplot(x = ST44Q08, fill = top20, data = subset(math2020, !is.na(ST44Q08)), geom = "bar", position = "fill")
+
+#Father Current Job Status
+qplot(x = ST19Q01, fill = top20, data = subset(math2020, !is.na(ST19Q01)), position = "fill")
+
+#mother current job status
+qplot(x = ST15Q01, fill = top20, data = subset(math2020, !is.na(ST15Q01)), position = "fill")
+
+#hours studied
+qplot(x = OUTHOURS, fill = top20, data = subset(math2020, !is.na(OUTHOURS)), position = "fill")
+```

### 2014-06-20 ijlyttle/factor_explorer.Rmd
a possibly-ill-fated experiment to make a data-exploration tool

--- 2014-06-01 ijlyttle/first_pass.Rmd
@@
-# First pass through the data
+---
+title: "factor_explorer"
+runtime: shiny
+output: html_document
+---
 
-This will be a stream-of-concoiusness walk through the data, a first pass to see what I can see.
+## Intro
+
+ggvis is amlmost there, but not quite, so I'll use ggplot.
 
@@
 library(dplyr)
+library(tidyr)
 library(lubridate)
-library(ggvis)
+library(shiny)
 ```
@@
 ```
-
-I suspect that for publication, `ggplot2` will be more capable than `ggvis`.
-
-# Load the data
 
@@
 
-Looking at the number of variables in each of the data frames, I wish I knew how to use ggobi...
+## Shiny app
+
+We are told that we can put a shiny app into an Rmarkdown document - let's try that.
+
+The thought is that we could use this (I wish we had another couple weeks) to discover the interesting relationships, then turn those into posters.
+
+```{r shiny_variable_lists}
+var_names_student <- dict_student2012$variable
+names(var_names_student) <- 
+  dict_student2012$description %>%
+  str_replace_all("\x92", "'") # gets rid of funny characters"
+```
+
+```{r shiny_inputs}
+selectInput(
+  "test",
+  "variable",
+  choices = var_names_student
+)
+```

### 2014-06-20 ijlyttle/factor_explorer.Rmd
adds comments on packages

@@
 library(shiny)
+
+# 2014-06-20
+# for what it's worth, I'm running the latest versions of RStudio (0.98.932 )and 
+# github versions of "rstudio/shiny", "rstudio/rmarkdown", "yihui/knitr"
 ```

### 2014-06-20 alex/2020.Rmd
fixed to not average PV

@@
 
-```{r avg, cache=TRUE, echo=FALSE}
+```{r avg, cache=TRUE, echo=FALSE, eval=FALSE}
 avg_test = function(subj) {
@@
 student2012$sci = avg_test("SCIE")
+```
+
+```{r pv, cache=TRUE, echo=FALSE}
+#choose which of the 5 PVs to use for each subject
+student2012$math = as.numeric(student2012[, "PV1MATH"])
+student2012$read = as.numeric(student2012[, "PV1READ"])
+student2012$sci = as.numeric(student2012[, "PV1SCIE"])
 ```

### 2014-06-20 dicook/Look.Rmd
getting cleaner map code

@@
 qplot(long, lat, data=student2012.gender.map, group=group, order=order, geom="polygon", fill=PropBoys) + coord_map()
+
+# Use dplyr fo rcheck for mismatches
+library(dplyr)
+library(rworldmap)
+world <- getMap(resolution = "low")
+world$NAME
+colnames(student2012.gender)[1] <- "NAME"
+sort(unique(world$region))
+anti_join(student2012.gender, world@data)
+student2012.gender$NAME[student2012.gender$NAME=="United States of America"] <- "United States"
+student2012.gender$NAME[student2012.gender$NAME=="USA"] <- "United States"
+student2012.gender$NAME[student2012.gender$NAME=="Czech Republic"] <- "Czech Rep."
+student2012.gender$NAME[student2012.gender$NAME=="Chinese Taipei"] <- "Taiwan"
+student2012.gender$NAME[student2012.gender$NAME=="Slovak Republic"] <- "Slovakia"
+student2012.gender$NAME[student2012.gender$NAME=="Russian Federation"] <- "Russia"
+student2012.gender$NAME[student2012.gender$NAME=="Korea"] <- "S. Korea"
+student2012.gender$NAME[student2012.gender$NAME=="Jordan"] <- "Jordan"
+student2012.gender$NAME[student2012.gender$NAME=="Hong Kong-China"] <- "Hong Kong"
+student2012.gender$NAME[student2012.gender$NAME=="China-Shanghai"] <- "China"
+student2012.gender.map <- inner_join(student2012.gender, world@data)
+new_theme_empty <- theme_bw()
+new_theme_empty$line <- element_blank()
+new_theme_empty$rect <- element_blank()
+new_theme_empty$strip.text <- element_blank()
+new_theme_empty$axis.text <- element_blank()
+new_theme_empty$plot.title <- element_blank()
+new_theme_empty$axis.title <- element_blank()
+new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
+
+ggplot(world) + aes(long, lat, group=group) + 
+  geom_polygon(fill="grey80") +
+  geom_path(color="white") +
+  coord_equal() + new_theme_empty
+student2012.gender.map <- joinData2Map(student2012.gender, world, "NAME", "NAME")
+ggplot(student2012.gender.map) + aes(long, lat, group=group, fill=PropBoys) + 
+  geom_polygon() +
+  #geom_path(color="white") +
+  coord_equal() + new_theme_empty
+
+extractPolygons <- function(shapes) {
+
+  dframe <- ldply(1:length(shapes@polygons), function(i) {
+    ob <- shapes@polygons[[i]]@Polygons
+    dframe <- ldply(1:length(ob), function(j) {
+      x <- ob[[j]]
+      co <- x@coords
+      data.frame(co, order=1:nrow(co), group=j)
+    })
+    dframe$region <- i
+    dframe$name <- shapes@polygons[[i]]@ID
+    dframe
+  })
+  # construct a group variable from both group and polygon:
+  dframe$group <- interaction(dframe$region, dframe$group)
+
+  dframe
+}
+
+world.polys <- extractPolygons(world)
+colnames(student2012.sub)[1] <- "name"
+student2012.sub[,1] <- as.character(student2012.sub[,1])
+unique(anti_join(student2012.sub, world.polys)[1])
+student2012.sub$name[student2012.sub$name=="United States"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Czech Republic"] <- "Czech Rep."
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "S. Korea"
+student2012.sub$name[student2012.sub$name=="Jordan"] <- "Jordan"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="S. Korea"] <- "Korea"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub.map <- join(student2012.sub, world.polys)
+unique(student2012.sub$name)
+qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=PropBoys) + new_theme_empty
+
 ```

### 2014-06-20 dicook/map.Rmd
getting cleaner map code

--- 2014-06-10 dicook/Look.Rmd
@@
-PISA
+Exploring maps
 ========================================================
-
-Working from Ian's code to read in the data - it is beautiful!
 
@@
 library(ggmap)
+library(rworldmap)
 ```
@@
 
-Do some basic summaries, to see what's in the data.
+```{r usefulfn}
+library(plyr)
+extractPolygons <- function(shapes) {
 
-```{r whatsthere}
-# Schools
-dim(dict_school2012)
-table(school2012$CNT)
-dim(dict_student2012)
-length(table(student2012$STIDSTD))
-subset(student2012, STIDSTD=="09631")
-dict_student2012[,2]
-table(student2012$ST04Q01)
-length(scoredItem2012dict)
-length(item2012dict)
-length(parent2012dict)
+  dframe <- ldply(1:length(shapes@polygons), function(i) {
+    ob <- shapes@polygons[[i]]@Polygons
+    dframe <- ldply(1:length(ob), function(j) {
+      x <- ob[[j]]
+      co <- x@coords
+      data.frame(co, order=1:nrow(co), group=j)
+    })
+    dframe$region <- i
+    dframe$name <- shapes@polygons[[i]]@ID
+    dframe
+  })
+  # construct a group variable from both group and polygon:
+  dframe$group <- interaction(dframe$region, dframe$group)
+
+  dframe
+}
+
+# To get a blank background on map
+new_theme_empty <- theme_bw()
+new_theme_empty$line <- element_blank()
+new_theme_empty$rect <- element_blank()
+new_theme_empty$strip.text <- element_blank()
+new_theme_empty$axis.text <- element_blank()
+new_theme_empty$plot.title <- element_blank()
+new_theme_empty$axis.title <- element_blank()
+new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
+
 ```
 
-Dizzyingly many variables! (GGobi won't help) Need to map out a plan of attack, questions of interest, and variables to create.
+```{r maps}
+# Extract map polygons for modern world
+world <- getMap(resolution = "low")
+world.polys <- extractPolygons(world)
+detach("package:plyr")
+# Subset data
+student2012.sub <- student2012[, c(1:7, 500:550)]
+colnames(student2012.sub)[1] <- "name"
+student2012.sub$name <- as.character(student2012.sub$name)
+# Check mismatches of names
+unique(anti_join(student2012.sub, world.polys)[1])
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
 
-- It will be good to include map information, so we can make maps
+student2012.sub.map <- left_join(student2012.sub, world.polys)
+unique(student2012.sub$name)
+qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=PropBoys) + new_theme_empty
 
-- Variables need to be re-organized, many columns represent one level of a categorical variable
-
-- What questions would we have about the data?
-
-- Which variable has the score?
-
-Now create a few variables, to provide examples, and make a few pictures.
-
-```{r usa}
-school.usa <- subset(school2012, CNT=="United States of America")
-dim(school.usa)
-school.usa$GenderPropBoys <- school.usa[,15]/(school.usa[,15]+school.usa[,16])
-summary(school.usa$GenderPropBoys)
-qplot(GenderPropBoys, data=school.usa)
-table(school.usa$SC03Q01)
-qplot(GenderPropBoys, SC03Q01, data=school.usa)
-summarise(group_by(school.usa, SC03Q01), median(GenderPropBoys, na.rm=T))
-qplot(GenderPropBoys, PROPMATH, data=school.usa, ylim=c(0,0.3), xlim=c(0.25, 0.75)) # Ignore extremese
-table(school.usa$SC01Q01)
-summary(school.usa[,290])
 ```
-
-Merge data with maps, so it can be plotted on a map. Interesting set of countries in the study - or more what countries are not included! This indicates some inequities.
-
-```{r maps}
-student2012.sub <- student2012[,c(1,12)]
-student2012.gender <- summarise(group_by(student2012.sub, CNT), PropBoys=length(ST04Q01[ST04Q01=="Male"])/length(ST04Q01))
-student2012.gender$CNT <- factor(student2012.gender$CNT, levels=levels(student2012.gender$CNT)[order(student2012.gender$PropBoys)])
-qplot(CNT, PropBoys, data=student2012.gender) + coord_flip()
-map.CNT <- map()$names
-world <- map_data("world")
-student2012.gender$CNT <- as.character(student2012.gender$CNT)
-student2012.gender$CNT[student2012.gender$CNT=="United States of America"] <- "USA"
-student2012.gender.map <- merge(student2012.gender, world, by.x="CNT", by.y="region")
-qplot(long, lat, data=student2012.gender.map, group=group, order=order, geom="polygon", fill=PropBoys) + coord_map()
-```

### 2014-06-20 alex/2020.Rmd

@@
 ```
+
+For each country, find the percentage of top 20% students in each school.
+```{r math-sci-read}
+school2020 = student2012 %>% 
+              select(CNT, math, SCHOOLID) %>%
+              group_by(CNT) %>%
+              mutate(top20 = math > quantile(math, probs = seq(0, 1, 0.2))[5]) %>%
+              group_by(CNT, SCHOOLID) %>%
+              summarise(num_top20 = sum(top20), total_students = length(top20)) %>%
+              mutate(pct_top20 = num_top20 / total_students)
+
+```

### 2014-06-20 ijlyttle/factor_explorer.Rmd

@@
 library(shiny)
+library(magrittr)
 
@@
 rm(fn_make_df)
-rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
+#rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
 ```
@@
   str_replace_all("\x92", "'") # gets rid of funny characters"
+
+# using pv1 for now
+var_names_student_subject <- c(
+  math = "PV1MATH",
+  science = "PV1SCIE",
+  reading = "PV1READ"
+)
+
+var_names_student_type <- 
+  lapply(student2012, class)
+
+var_names_student_factor <- 
+  var_names_student_type[var_names_student_type == "factor"] %>%
+  names
+
+tmp <- var_names_student_factor[var_names_student_factor == var_names_student] 
 ```
 
-```{r shiny_inputs}
+```{r shiny_inputs, echo = FALSE}
 selectInput(
-  "test",
-  "variable",
-  choices = var_names_student
+  "subject",
+  label = "Subject",
+  choices = var_names_student_subject
 )
+
+selectInput(
+  "country",
+  label = "Countries",
+  choices = levels(student2012$CNT),
+  selected = "Albania",
+  multiple = TRUE
+)
+
+# selectInput(
+#   "factor"
+#   label = "Factor",
+#   choices = 
+# )
 ```
+
+```{r shiny_transform, echo = FALSE}
+rct_data <- reactive({
+  df <- 
+    student2012 %>%
+    filter(CNT %in% input$country)
+
+})
+
+renderPlot({
+  ggplot(
+    data = rct_data()
+  ) + 
+  geom_violin(
+    aes_string(x = "factor(ST06Q01)", 
+               y = input$subject), 
+    scale = "width"
+  )
+})
+
+```

### 2014-06-20 dicook/map.Rmd
first draft of a map

@@
 ```{r usefulfn}
-library(plyr)
 extractPolygons <- function(shapes) {
@@
 new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
-
 ```
 
-```{r maps}
+```{r maps, fig.width=12, fig.height=7}
 # Extract map polygons for modern world
 world <- getMap(resolution = "low")
+library(plyr)
 world.polys <- extractPolygons(world)
@@
 
-student2012.sub.map <- left_join(student2012.sub, world.polys)
-unique(student2012.sub$name)
-qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=PropBoys) + new_theme_empty
+# Only need one of the plausible values, checked they are effectively identical
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1MACC <- as.numeric(student2012.sub$PV1MACC)
+student2012.sub$PV1MACQ <- as.numeric(student2012.sub$PV1MACQ)
+student2012.sub$PV1MACS <- as.numeric(student2012.sub$PV1MACS)
+student2012.sub$PV1MACU <- as.numeric(student2012.sub$PV1MACU)
+student2012.sub$PV1MAPE <- as.numeric(student2012.sub$PV1MAPE)
+student2012.sub$PV1MAPF <- as.numeric(student2012.sub$PV1MAPF)
+student2012.sub$PV1MAPI <- as.numeric(student2012.sub$PV1MAPI)
+student2012.sub.math <- summarise(group_by(student2012.sub[,c(1,seq(9,48,5))], name), math=mean(PV1MATH), mCC=mean(PV1MACC), 
+                                  mCQ=mean(PV1MACQ), mCS=mean(PV1MACS), mCU=mean(PV1MACU), mPE=mean(PV1MAPE), 
+                                  mPF=mean(PV1MAPF), mPI=mean(PV1MAPI))
+#pairs(student2012.sub.math[,-1])
 
+# Left join to only get countries that are measured
+student2012.sub.map <- left_join(student2012.sub.math, world.polys)
+#qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=math) + coord_map() + new_theme_empty
+
+# Really need all boundaries, doesn't quite work to have boundaries with one data: coord_map is the problem
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=math)) +
+#  coord_map() + 
+  new_theme_empty + theme(legend.position="none")
+
+# Join to get missings for countries that are not measured
+#student2012.sub.map <- merge(student2012.sub.math, world.polys, all.y=TRUE)
+#qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=math, colour=I("grey70")) + 
+  #coord_map(xlim=c(-195, 185), ylim=c(-150, 95)) + # Having trouble getting coord_map to not cut off polygons weirdly
+#  new_theme_empty + theme(legend.position="none")
 ```

### 2014-06-20 dicook/map.Rmd
got nice map, even with inserts

@@
 library(rworldmap)
+library (grid)  	
+library (scales)	
 ```
@@
 # Check mismatches of names
-unique(anti_join(student2012.sub, world.polys)[1])
+# unique(anti_join(student2012.sub, world.polys)[1])
 student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
@@
                                   mPF=mean(PV1MAPF), mPI=mean(PV1MAPI))
-#pairs(student2012.sub.math[,-1])
+colnames(student2012.sub.math)[-c(1:2)] <- c("Change", "Quantity", "Spatial", "Data", "Employ", "Formulate", "Interpret")
 
@@
 
-# Join to get missings for countries that are not measured
-#student2012.sub.map <- merge(student2012.sub.math, world.polys, all.y=TRUE)
-#qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=math, colour=I("grey70")) + 
-  #coord_map(xlim=c(-195, 185), ylim=c(-150, 95)) + # Having trouble getting coord_map to not cut off polygons weirdly
-#  new_theme_empty + theme(legend.position="none")
+# Now try to do an insert
+# Note China is represented just by Shanghai
+student2012.sub.math$name <- factor(student2012.sub.math$name, 
+      levels=student2012.sub.math$name[order(student2012.sub.math$math)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=math)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, math, data=student2012.sub.math, colour=math, ylab="Math Score", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)  			
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
 ```
+
+```{r pairs, fig.width=12, fig.height=12}
+library(YaleToolkit)
+gpairs(student2012.sub.math[,-1])
+# Should do a PCA, to look at what countries do better on what types of math skills
+```

### 2014-06-21 ijlyttle/factor_explorer.Rmd
gets student explorer going

@@
 
-# using pv1 for now
-var_names_student_subject <- c(
-  math = "PV1MATH",
-  science = "PV1SCIE",
-  reading = "PV1READ"
-)
+var_names_student_subject <- 
+  var_names_student[c(
+    seq(501, 505), # math
+    seq(541, 545), # reading
+    seq(546, 551)  # science
+  )]
 
@@
 
-tmp <- var_names_student_factor[var_names_student_factor == var_names_student] 
+var_names_factor <- var_names_student[var_names_student %in% var_names_student_factor] 
 ```
@@
   choices = levels(student2012$CNT),
-  selected = "Albania",
+  selected = c("United States of America", "Canada"),
   multiple = TRUE
@@
 
-# selectInput(
-#   "factor"
-#   label = "Factor",
-#   choices = 
-# )
+selectInput(
+  "x_factor",
+  label = "X-axis factor",
+  choices = var_names_factor,
+  selected = "ST04Q01"
+)
+
+selectInput(
+  "fill_factor",
+  label = "Fill factor",
+  choices = var_names_factor,
+  selected = "ST26Q03"
+)
 ```
@@
     filter(CNT %in% input$country) 
+})
 
+val_get_name <- function(x, lst){
+  names(lst[lst == x])[[1]]
+}
+
+rct_count <- reactive({
+  df <-
+    rct_data() %>%
+    group_by()
+})
+
+rct_labels <- reactive({
+  list(
+    x = input$x_factor %>% val_get_name(var_names_factor),
+    y = input$subject %>% val_get_name(var_names_student_subject),
+    fill = input$fill_factor %>% val_get_name(var_names_factor)
+  )
 })
@@
   geom_violin(
-    aes_string(x = "factor(ST06Q01)", 
+    aes_string(
+      x = input$x_factor, 
       y = input$subject),
       scale = "width"
+  ) + 
+  geom_boxplot(
+    aes_string(
+      x = input$x_factor,
+      y = input$subject,
+      fill = input$fill_factor
     )  
+  ) + 
+  facet_grid(CNT ~ .) + 
+  scale_x_discrete(name = rct_labels()[["x"]]) +  
+  scale_y_continuous(limits = c(0, 1000), name = rct_labels()[["y"]]) +
+  scale_fill_discrete(name = rct_labels()[["fill"]]) 
 })
-
 ```

### 2014-06-22 dicook/map.Rmd
tinkering with maps, dotplots and pca

--- 2014-06-20 dicook/map.Rmd
@@
 new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
-
 ```
 
-```{r maps}
+```{r summarize}
+sort(table(student2012$CNT))
+schools.sampled <- summarise(group_by(student2012[,c(1,6)], CNT), numschools=length(unique(SCHOOLID)))
+schools.sampled[order(schools.sampled$numschools, decreasing=T),]
+```
+
+```{r maps, fig.width=12, fig.height=7}
 # Extract map polygons for modern world
 world <- getMap(resolution = "low")
+library(plyr)
 world.polys <- extractPolygons(world)
@@
 # Subset data
-student2012.sub <- student2012[, c(1:7, 500:550)]
+student2012.sub <- student2012[, c(1:7, seq(501, 550, 5))]
 colnames(student2012.sub)[1] <- "name"
@@
 student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
 student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
 student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
+student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
+student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
 
-student2012.sub.map <- left_join(student2012.sub, world.polys)
-unique(student2012.sub$name)
-qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=PropBoys) + new_theme_empty
+# Only need one of the plausible values, checked they are effectively identical
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1MACC <- as.numeric(student2012.sub$PV1MACC)
+student2012.sub$PV1MACQ <- as.numeric(student2012.sub$PV1MACQ)
+student2012.sub$PV1MACS <- as.numeric(student2012.sub$PV1MACS)
+student2012.sub$PV1MACU <- as.numeric(student2012.sub$PV1MACU)
+student2012.sub$PV1MAPE <- as.numeric(student2012.sub$PV1MAPE)
+student2012.sub$PV1MAPF <- as.numeric(student2012.sub$PV1MAPF)
+student2012.sub$PV1MAPI <- as.numeric(student2012.sub$PV1MAPI)
+student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
+student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub.summary <- summarise(group_by(student2012.sub[,c(1,8:17)], name), math=mean(PV1MATH, na.rm=T), 
+                                  mCC=mean(PV1MACC, na.rm=T), mCQ=mean(PV1MACQ, na.rm=T), 
+                                  mCS=mean(PV1MACS, na.rm=T), mCU=mean(PV1MACU, na.rm=T), 
+                                  mPE=mean(PV1MAPE, na.rm=T), mPF=mean(PV1MAPF, na.rm=T), 
+                                  mPI=mean(PV1MAPI, na.rm=T), read=mean(PV1READ, na.rm=T), 
+                                  science=mean(PV1SCIE, na.rm=T), 
+                                     mathr=diff(range(PV1MATH, na.rm=T)), 
+                                  mCCr=diff(range(PV1MACC, na.rm=T)), mCQr=diff(range(PV1MACQ, na.rm=T)), 
+                                  mCSr=diff(range(PV1MACS, na.rm=T)), mCUr=diff(range(PV1MACU, na.rm=T)), 
+                                  mPEr=diff(range(PV1MAPE, na.rm=T)), mPFr=diff(range(PV1MAPF, na.rm=T)), 
+                                  mPIr=diff(range(PV1MAPI, na.rm=T)), readr=diff(range(PV1READ, na.rm=T)), 
+                                  sciencer=diff(range(PV1SCIE, na.rm=T)))
+colnames(student2012.sub.summary)[3:9] <- c("Change", "Quantity", "Spatial", "Data", "Employ", "Formulate", "Interpret")
 
+# Left join to only get countries that are measured
+student2012.sub.map <- left_join(student2012.sub.summary, world.polys)
+#qplot(X1, X2, order=order, group=group, data=student2012.sub.map, geom="polygon", fill=math) + coord_map() + new_theme_empty
+
+# Really need all boundaries, doesn't quite work to have boundaries with one data: coord_map is the problem
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=math)) +
+  new_theme_empty + theme(legend.position="none")
+
+# Now try to do an insert
+# Note China is represented just Shanghai and Macau
+# Individual states in the USA, Florida, Mass, Conn are included as US
+# Russian Federation has two groups intributinf
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$math)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=math)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, math, data=student2012.sub.summary, colour=math, ylab="Math Score", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)  			
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
+
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$mathr)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=mathr)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, mathr, data=student2012.sub.summary, colour=mathr, ylab="Range Math", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)    		
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
+
+# Reading
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$read)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=read)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, read, data=student2012.sub.summary, colour=read, ylab="Reading Score", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)    		
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
+
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$readr)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=readr)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, readr, data=student2012.sub.summary, colour=readr, ylab="Range Reading", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)    		
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
+
+# Science
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$science)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=science)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, science, data=student2012.sub.summary, colour=science, ylab="Science Score", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)      	
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
+
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$sciencer)])
+p1 <- ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=sciencer)) +
+  new_theme_empty + theme(legend.position="none")
+p2 <- qplot(name, sciencer, data=student2012.sub.summary, colour=sciencer, ylab="Range Science", xlab="") + coord_flip() + 
+  theme(legend.position="none")
+p2 = ggplotGrob(p2)    		
+p1 + annotation_custom(grob = p2, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
 ```
+
+```{r all, fig.width=8, fig.height=12}
+student2012.sub.iqr <- summarise(group_by(student2012.sub[,c(1,8:17)], name), q0=min(PV1MATH, na.rm=T),
+                                 q25=quantile(PV1MATH, 0.25, na.rm=T), q50=median(PV1MATH, na.rm=T), 
+                                 q75=quantile(PV1MATH, 0.75, na.rm=T), q100=max(PV1MATH, na.rm=T), 
+                                 count=length(PV1MATH))
+student2012.sub.iqr$name <- factor(student2012.sub.iqr$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$math)])
+ggplot(data=student2012.sub.iqr) + ylab("Math Score") + xlab("") + ylim(c(0,1000)) + 
+  geom_point(aes(x=name, y=q50, size=count)) + geom_segment(aes(x=name, xend=name, y=q0, yend=q25)) +
+  geom_segment(aes(x=name, xend=name, y=q75, yend=q100)) +
+  coord_flip() + theme(legend.position="none")
+student2012.sub.iqr <- summarise(group_by(student2012.sub[,c(1,8:17)], name), q0=min(PV1READ, na.rm=T),
+                                 q25=quantile(PV1READ, 0.25, na.rm=T), q50=median(PV1READ, na.rm=T), 
+                                 q75=quantile(PV1READ, 0.75, na.rm=T), q100=max(PV1READ, na.rm=T), 
+                                 count=length(PV1READ))
+student2012.sub.iqr$name <- factor(student2012.sub.iqr$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$read)])
+ggplot(data=student2012.sub.iqr) + ylab("Reading Score") + xlab("") + ylim(c(0,1000)) + 
+  geom_point(aes(x=name, y=q50, size=count)) + geom_segment(aes(x=name, xend=name, y=q0, yend=q25)) +
+  geom_segment(aes(x=name, xend=name, y=q75, yend=q100)) +
+  coord_flip() + theme(legend.position="none")
+student2012.sub.iqr <- summarise(group_by(student2012.sub[,c(1,8:17)], name), q0=min(PV1SCIE, na.rm=T),
+                                 q25=quantile(PV1SCIE, 0.25, na.rm=T), q50=median(PV1SCIE, na.rm=T), 
+                                 q75=quantile(PV1SCIE, 0.75, na.rm=T), q100=max(PV1SCIE, na.rm=T), 
+                                 count=length(PV1SCIE))
+student2012.sub.iqr$name <- factor(student2012.sub.iqr$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$science)])
+ggplot(data=student2012.sub.iqr) + ylab("Science Score") + xlab("") + ylim(c(0,1000)) + 
+  geom_point(aes(x=name, y=q50, size=count)) + geom_segment(aes(x=name, xend=name, y=q0, yend=q25)) +
+  geom_segment(aes(x=name, xend=name, y=q75, yend=q100)) +
+  coord_flip() + theme(legend.position="none")
+```
+
+```{r pairs, fig.width=12, fig.height=12}
+library(YaleToolkit)
+gpairs(student2012.sub.summary[, c(2, 10, 11)])
+# library(cranvas)
+# qstudent <- qdata(student2012.sub.summary)
+# qscatter(math, read, qstudent)
+# qscatter(math, science, qstudent)
+# qscatter(read, science, qstudent)
+# record_selector(name, qstudent)
+# Should do a PCA, to look at what countries do better on what types of math skills
+student2012.sub.summary.nomiss <- subset(student2012.sub.summary, !is.na(student2012.sub.summary$Change)&!is.na(student2012.sub.summary$Employ))
+rownames(student2012.sub.summary.nomiss) <- student2012.sub.summary.nomiss[,1]
+student2012.sub.math.pca <- prcomp(student2012.sub.summary.nomiss[,2:9], scale=T, retx=T)
+student2012.sub.math.pca
+qplot(PC2, PC3, data=data.frame(student2012.sub.math.pca$x)) + theme(aspect.ratio=1)
+# High values on PC2 correspond to high data and interpretation, low corresponds to high spatial 
+# Ireland, UK, Greece, US states, Netherland, Norway, NZ, Finland, France, Croatia, Australia come up as high 
+# on data and interp but less on spatial AND
+# Chine, Albania, Taiwan, Kazakhstan, Korea, Japan, Russia come up as being high on spatial, less on data and interp
+rownames(student2012.sub.math.pca$x)[order(student2012.sub.math.pca$x[,2], decreasing=T)]
+# High on PC3 correspond to high quantification, employ, low on data
+# Israel, Croatia, Estonia, Czech are high; Taiwan, Japan, Indonesia, Massachussetts, Jordan are low
+rownames(student2012.sub.math.pca$x)[order(student2012.sub.math.pca$x[,3], decreasing=T)]
+student2012.sub.summary.pca <- prcomp(student2012.sub.summary.nomiss[,c(2,10,11)], scale=T, retx=T)
+student2012.sub.summary.pca
+```

### 2014-06-22 ijlyttle/factor_explorer.Rmd

@@
   scale_y_continuous(limits = c(0, 1000), name = rct_labels()[["y"]]) +
-  scale_fill_discrete(name = rct_labels()[["fill"]]) 
+  scale_fill_brewer(type = "seq", name = rct_labels()[["fill"]]) 
 })

### 2014-06-22 dicook/possessions.Rmd
looking at possessions, age

Exploring possessions, and other demographics
========================================================

```{r load_libraries, message=FALSE}
library(stringr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
library(rworldmap)
library (grid)    
library (scales)	
```

```{r load_data, echo=FALSE}
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){

  template <- c("2012.rda", "2012dict.rda")

  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path(".", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
#rm(fn_build, sets)
```

```{r modify_dictionaries, echo=FALSE}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict)

# clean
#rm(fn_make_df)
#rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

```{r usefulfn, echo=FALSE}
extractPolygons <- function(shapes) {

  dframe <- ldply(1:length(shapes@polygons), function(i) {
    ob <- shapes@polygons[[i]]@Polygons
    dframe <- ldply(1:length(ob), function(j) {
      x <- ob[[j]]
      co <- x@coords
      data.frame(co, order=1:nrow(co), group=j)
    })
    dframe$region <- i
    dframe$name <- shapes@polygons[[i]]@ID
    dframe
  })
  # construct a group variable from both group and polygon:
  dframe$group <- interaction(dframe$region, dframe$group)

  dframe
}

# To get a blank background on map
new_theme_empty <- theme_bw()
new_theme_empty$line <- element_blank()
new_theme_empty$rect <- element_blank()
new_theme_empty$strip.text <- element_blank()
new_theme_empty$axis.text <- element_blank()
new_theme_empty$plot.title <- element_blank()
new_theme_empty$axis.title <- element_blank()
new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
```

```{r countpossessions, warning=FALSE, message=FALSE}
student2012.sub <- student2012[, c(1:7, 44:57, 61:66, seq(501, 550, 5))]
colnames(student2012.sub)[1] <- "name"
for (i in 8:21) {
  student2012.sub[,i] <- as.character(student2012.sub[,i])
  student2012.sub[is.na(student2012.sub[,i]), i] <- ""
  student2012.sub[student2012.sub[,i]=="Yes",i] <- "1"
  student2012.sub[student2012.sub[,i]=="No",i] <- "0"
  student2012.sub[,i] <- as.numeric(student2012.sub[,i])
}
for (i in 22:26) {
  student2012.sub[,i] <- as.character(student2012.sub[,i])
  student2012.sub[is.na(student2012.sub[,i]), i] <- ""
  student2012.sub[student2012.sub[,i]=="None",i] <- "0"
  student2012.sub[student2012.sub[,i]=="One",i] <- "1"
  student2012.sub[student2012.sub[,i]=="Two",i] <- "2"
  student2012.sub[student2012.sub[,i]=="Three or more",i] <- "3"
  student2012.sub[,i] <- as.numeric(student2012.sub[,i])
}
student2012.sub[,27] <- as.character(student2012.sub[,27])
student2012.sub[is.na(student2012.sub[,27]), 27] <- ""
student2012.sub[student2012.sub[,27]=="0-10 books ",27] <- "0"
student2012.sub[student2012.sub[,27]=="11-25 books ",27] <- "1"
student2012.sub[student2012.sub[,27]=="26-100 books ",27] <- "2"
student2012.sub[student2012.sub[,27]=="101-200 books ",27] <- "3"
student2012.sub[student2012.sub[,27]=="201-500 books ",27] <- "4"
student2012.sub[student2012.sub[,27]=="More than 500 books",27] <- "5"
student2012.sub[,27] <- as.numeric(student2012.sub[,27])
student2012.sub$numposs <- apply(student2012.sub[,8:21], 1, sum, na.rm=T)
student2012.sub$numedposs <- apply(student2012.sub[,22:27], 1, sum, na.rm=T)
```

```{r plotstuff, fig.width=8, fig.height=8, warning=FALSE, message=FALSE}
qplot(numposs, PV1MATH, data=student2012.sub, alpha=I(0.1), geom=c("point", "smooth"))
qplot(numedposs, PV1MATH, data=student2012.sub, alpha=I(0.1), geom=c("point", "smooth"))
student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
student2012.sub$PV1MACC <- as.numeric(student2012.sub$PV1MACC)
student2012.sub$PV1MACQ <- as.numeric(student2012.sub$PV1MACQ)
student2012.sub$PV1MACS <- as.numeric(student2012.sub$PV1MACS)
student2012.sub$PV1MACU <- as.numeric(student2012.sub$PV1MACU)
student2012.sub$PV1MAPE <- as.numeric(student2012.sub$PV1MAPE)
student2012.sub$PV1MAPF <- as.numeric(student2012.sub$PV1MAPF)
student2012.sub$PV1MAPI <- as.numeric(studen=t2012.sub$PV1MAPI)
student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
student2012.sub.summary <- summarise(group_by(student2012.sub[,c(1,8:39)], name), math=mean(PV1MATH, na.rm=T), 
          read=mean(PV1READ, na.rm=T), science=mean(PV1SCIE, na.rm=T),
          poss = mean(numposs, na.rm=T), edposs = mean(numedposs, na.rm=T))
qplot(poss, math, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
qplot(poss, read, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
qplot(poss, science, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
head(student2012.sub.summary[order(student2012.sub.summary$math, decreasing=T),], 20) 
# Vietnam is the outlier, few poss, high scores
lm(math~poss, data=student2012.sub.summary)
# Each increase in one possession tends to increase math score by 25 points (poss=7, 12)
qplot(edposs, math, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
qplot(edposs, read, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
qplot(edposs, science, data=student2012.sub.summary) + geom_smooth(method="lm", se=F)
lm(math~edposs, data=student2012.sub.summary)
# Each increase in one possession tends to increase math score by 13 points (edposs=5, 14)
qplot(edposs, poss, data=student2012.sub.summary)
```

```{r plotstuff2, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
student2012.sub.summary <- summarise(group_by(student2012.sub[,c(1,8,28,38)], name), 
              math05=mean(PV1MATH[numposs<6], na.rm=T), math06=mean(PV1MATH[numposs==6], na.rm=T), 
              math07=mean(PV1MATH[numposs==7], na.rm=T), math08=mean(PV1MATH[numposs==8], na.rm=T),
              math09=mean(PV1MATH[numposs==9], na.rm=T), math10=mean(PV1MATH[numposs==10], na.rm=T),
              math11=mean(PV1MATH[numposs==11], na.rm=T), math12=mean(PV1MATH[numposs==12], na.rm=T),
              math13=mean(PV1MATH[numposs==13], na.rm=T), math14=mean(PV1MATH[numposs==14], na.rm=T))
student2012.sub.summary.m <- melt(student2012.sub.summary)
student2012.sub.summary.m$variable <- substr(student2012.sub.summary.m$variable, 5, 6) 
student2012.sub.summary.m$variable <- as.numeric(student2012.sub.summary.m$variable)
qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + facet_wrap(~name, ncol=8)
```

```{r age, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
summary(student2012$ST06Q01)
student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
student2012.age.iqr <- summarise(group_by(student2012[,c(1,14)], CNT), q0=min(ST06Q01, na.rm=T),
                                 q25=quantile(ST06Q01, 0.25, na.rm=T), q50=median(ST06Q01, na.rm=T), 
                                 q75=quantile(ST06Q01, 0.75, na.rm=T), q100=max(ST06Q01, na.rm=T), 
                                 count=length(ST06Q01))
student2012.age.iqr$CNT <- factor(student2012.age.iqr$CNT, 
      levels=student2012.age.iqr$CNT[order(student2012.age.iqr$q50)])
ggplot(data=student2012.age.iqr) + ylab("Age") + xlab("") + 
  geom_point(aes(x=CNT, y=q50, size=count)) + geom_segment(aes(x=CNT, xend=CNT, y=q0, yend=q25)) +
  geom_segment(aes(x=CNT, xend=CNT, y=q75, yend=q100)) +
  coord_flip() + theme(legend.position="none")
```

```{r agemath, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
student2012.sub <- student2012[, c(1, 14, 501)]
colnames(student2012.sub)[1] <- "name"
student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
student2012.sub.summary <- summarise(group_by(student2012.sub, name), math=mean(PV1MATH, na.rm=T), 
          age=mean(ST06Q01, na.rm=T))
qplot(age, math, data=student2012.sub.summary)
# Age at start school does not seem to make a difference on country level
# qplot(ST06Q01, PV1MATH, data=student2012, alpha=I(0.1), xlab="Age", ylab="Math")
student2012.sub.summary <- summarise(group_by(student2012.sub, name), math4=mean(PV1MATH[ST06Q01==4], na.rm=T), 
          math5=mean(PV1MATH[ST06Q01==5], na.rm=T), math6=mean(PV1MATH[ST06Q01==6], na.rm=T), 
          math7=mean(PV1MATH[ST06Q01==7], na.rm=T), math8=mean(PV1MATH[ST06Q01>7], na.rm=T))
student2012.sub.summary.m <- melt(student2012.sub.summary)
student2012.sub.summary.m$variable <- substr(student2012.sub.summary.m$variable, 5, 5)
student2012.sub.summary.m$variable <- as.numeric(student2012.sub.summary.m$variable)
qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + facet_wrap(~name, ncol=8)
# Looking by country is important: generally starting school early leads to better scores
```

### 2014-06-22 dicook/possessions.Rmd
looking at possessions, age

@@
 student2012.sub$PV1MAPF <- as.numeric(student2012.sub$PV1MAPF)
-student2012.sub$PV1MAPI <- as.numeric(studen=t2012.sub$PV1MAPI)
+student2012.sub$PV1MAPI <- as.numeric(student2012.sub$PV1MAPI)
 student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
@@
 student2012.sub.summary.m$variable <- as.numeric(student2012.sub.summary.m$variable)
-qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + facet_wrap(~name, ncol=8)
+qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + 
+  facet_wrap(~name, ncol=8)
 ```

### 2014-06-22 ijlyttle/factor_explorer.Rmd
new version of explorer

--- 2014-06-21 ijlyttle/factor_explorer.Rmd
@@
 # for what it's worth, I'm running the latest versions of RStudio (0.98.932 )and 
-# github versions of "rstudio/shiny", "rstudio/rmarkdown", "yihui/knitr"
+# github versions of "rstudio/shiny", "rstudio/rmarkdown", "yihui/knitr", "hadley/tidyr"
 ```
@@
     seq(541, 545), # reading
-    seq(546, 551)  # science
+    seq(546, 550)  # science
   )]
@@
 rct_count <- reactive({
+
+  # hacky hacky hacky  
+  str_symbols <- c("CNT", input$x_factor, input$fill_factor) 
+  symbols <- str_symbols %>% lapply(as.symbol)
+
+  df_data <- rct_data()
+
+  if (nrow(df_data) == 0)
+    return(NULL)
+
+#   # get counts by category
    df <- 
+     df_data[, str_symbols] %>%
+     regroup(symbols) %>%
+     summarize(n = n())
+
+  # get counts by country  
+  df_cnt <-
     rct_data() %>%
-    group_by()
+    group_by(CNT) %>%
+    summarize(n_country = n(), y_placeholder = 0)
+
+  # join, get proportion by country for each category
+  df_new <- 
+    left_join(df, df_cnt, by = "CNT") %>%
+    mutate(prop = n / n_country)
+
+  df_new
+})
+
+observe({  
+  print(rct_count())
+  print("")
+#  print(str(rct_count()))
+#  print("")
+#  print(summary(rct_count()))
 })
@@
       fill = input$fill_factor
-    )  
+    ),
+    alpha = 0.75
   ) + 
+#   geom_dotplot(
+#     aes_string(
+#       x = input$x_factor,
+#       y = input$subject,
+#       fill = input$fill_factor
+#     ),
+#     binaxis = "y", 
+#     stackdir = "center", 
+#     position = "dodge") + 
   facet_grid(CNT ~ .) + 
@@
   scale_y_continuous(limits = c(0, 1000), name = rct_labels()[["y"]]) +
-  scale_fill_discrete(name = rct_labels()[["fill"]]) 
+  scale_fill_brewer(type = "seq", name = rct_labels()[["fill"]]) 
+})
+
+renderPlot({
+  ggplot(
+    aes_string(
+      x = input$x_factor,
+      fill = input$fill_factor
+    ),
+    data = rct_data()
+  ) +
+  geom_bar(position = "dodge", color = "black") +   
+  facet_grid(CNT ~ ., scales = "free_y") + 
+  scale_x_discrete(name = rct_labels()[["x"]]) +  
+  scale_y_continuous(name = "number of students") +
+  scale_fill_brewer(type = "seq", name = rct_labels()[["fill"]]) 
 })

### 2014-06-22 dicook/map.Rmd

@@
     dframe$name <- shapes@polygons[[i]]@ID
+    dframe$area <- shapes@polygons[[i]]@area
     dframe

### 2014-06-22 ijlyttle/factor_explorer.Rmd
converts some student2012 variables, like starting age, to factor

@@
 The thought is that we could use this (I wish we had another couple weeks) to discover the interesting relationships, then turn those into posters.
+
+```{r make_factors}
+fn_make_factor <- function(vec){  
+  vec <- factor(vec, levels = sort(unique(vec)))
+}
+
+# hacky - fix when time
+student2012$ST01Q01 <- fn_make_factor(student2012$ST01Q01)
+student2012$ST02Q01 <- fn_make_factor(student2012$ST02Q01)
+student2012$ST06Q01 <- fn_make_factor(student2012$ST06Q01)
+student2012$ST21Q01 <- fn_make_factor(student2012$ST21Q01)
+student2012$ST115Q01 <- fn_make_factor(student2012$ST115Q01)
+
+```
 

### 2014-06-22 dicook/map.Rmd
new explorations on school, age and gender

@@
 
-```{r load_libraries, message=FALSE}
+```{r load_libraries, message=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
 library(stringr)
@@
 
-```{r load_data}
+```{r load_data, echo=FALSE, warning=FALSE, message=FALSE}
 setwd("..")
@@
 
-```{r modify_dictionaries}
+```{r modify_dictionaries, echo=FALSE, warning=FALSE, message=FALSE}
 # function to convert to data-frames
@@
 
-```{r usefulfn}
+```{r usefulfn, echo=FALSE, warning=FALSE, message=FALSE}
 extractPolygons <- function(shapes) {
@@
 
-```{r summarize}
+```{r summarize, warning=FALSE, message=FALSE}
 sort(table(student2012$CNT))
@@
 
-```{r maps, fig.width=12, fig.height=7}
+```{r maps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}
 # Extract map polygons for modern world
@@
 
-```{r all, fig.width=8, fig.height=12}
+```{r all, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
 student2012.sub.iqr <- summarise(group_by(student2012.sub[,c(1,8:17)], name), q0=min(PV1MATH, na.rm=T),
@@
 
-```{r pairs, fig.width=12, fig.height=12}
+```{r oz, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
+# Why is australia so varied?
+student2012.sub.oz <- student2012.sub[student2012.sub$name=="Australia",]
+student2012.sub.oz$SCHOOLID <- as.character(student2012.sub.oz$SCHOOLID)
+#table(student2012.sub.oz$SCHOOLID)
+#length(unique(student2012.sub.oz$SCHOOLID))
+student2012.sub.oz.mean <- summarise(group_by(student2012.sub[,c(1,6, 8)], SCHOOLID), math=mean(PV1MATH, na.rm=T), 
+                                     q0=min(PV1MATH, na.rm=T), q25=quantile(PV1MATH, p=0.25, na.rm=T), 
+                                     q50=median(PV1MATH, na.rm=T),
+                                     q75=quantile(PV1MATH, p=0.75, na.rm=T), q100=max(PV1MATH, na.rm=T))
+student2012.sub.oz.mean$SCHOOLID <- factor(student2012.sub.oz.mean$SCHOOLID, levels=student2012.sub.oz.mean$SCHOOLID[order(student2012.sub.oz.mean$q50)])
+ggplot(data=student2012.sub.oz.mean) + ylab("Math Score") + xlab("") + 
+  geom_point(aes(x=SCHOOLID, y=q50)) + geom_segment(aes(x=SCHOOLID, xend=SCHOOLID, y=q0, yend=q25), alpha=0.1) +
+  geom_segment(aes(x=SCHOOLID, xend=SCHOOLID, y=q75, yend=q100), alpha=0.1) +
+  coord_flip() + theme(legend.position="none")
+# So mostly the variation is student to student, not school to school like a friend suggested
+# Not the best way to display this, I think
+# Really want to be able to show that the best student at a low scoring school does better 
+# than the worst student at a high-scoring school
+```
+
+```{r schoolmath, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
+# Wonder how other countries compare
+student2012.sub <- student2012[, c(1, 6, 501)]
+colnames(student2012.sub)[1] <- "name"
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub.summary <- summarise(group_by(student2012.sub, name, SCHOOLID), math=median(PV1MATH, na.rm=T))
+student2012.sub.summary$SCHOOLID <- as.numeric(as.character(student2012.sub.summary$SCHOOLID))
+order <- summarise(group_by(student2012.sub, name, SCHOOLID), math=median(PV1MATH, na.rm=T))
+bymedian <- with(student2012.sub.summary, reorder(name, math, median))
+qplot(bymedian, math, data=student2012.sub.summary, ylim=c(0,1000), xlab="Math", ylab="") + coord_flip()
+# Pretty similar overall, Australia still comes up as having a big variance, a few low-scoring schools 
+# and a few high-scoring schools, Spain has a really low-scoring school
+```
+
+```{r pairs, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
 library(YaleToolkit)

### 2014-06-22 dicook/possessions.Rmd
new explorations on school, age and gender

@@
 student2012.sub.summary.m$variable <- as.numeric(student2012.sub.summary.m$variable)
-qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + 
+qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F, 
+      xlab="Possessions", ylab="Math") + 
   facet_wrap(~name, ncol=8)
@@
 student2012.sub.summary.m$variable <- as.numeric(student2012.sub.summary.m$variable)
-qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F) + facet_wrap(~name, ncol=8)
+qplot(variable, value, data=student2012.sub.summary.m, group=name, geom=c("point","smooth"), se=F, 
+      xlab="Age", ylab="Math") + 
+  facet_wrap(~name, ncol=8)
 # Looking by country is important: generally starting school early leads to better scores
 ```
+
+```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
+student2012.sub <- student2012[, c(1, 12, 501)]
+colnames(student2012.sub)[1] <- "name"
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub.summary <- summarise(group_by(student2012.sub, name), male=mean(PV1MATH[ST04Q01=="Male"], na.rm=T), 
+          female=mean(PV1MATH[ST04Q01=="Female"], na.rm=T))
+student2012.sub.summary.m <- melt(student2012.sub.summary)
+colnames(student2012.sub.summary.m)[2] <- "gender"
+colnames(student2012.sub.summary.m)[3] <- "math"
+qplot(gender, math, data=student2012.sub.summary.m, geom="boxplot") + theme(aspect.ratio=1)
+# Overall there is a gender gap in math, perhaps 10 points, and the high scores are almost exclusively males
+student2012.sub.summary.m$name <- factor(student2012.sub.summary.m$name, 
+      levels=student2012.sub.summary$name[order(student2012.sub.summary$male)])
+student2012.sub.summary.m$gender <- factor(student2012.sub.summary.m$gender, levels=c("female", "male"))
+qplot(name, math, data=student2012.sub.summary.m, colour=gender) + coord_flip()
+# Some countries have no gender gap, a few surprises females better than males
+student2012.sub.summary.gap <- summarise(group_by(student2012.sub.summary, name), gap=male-female)
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$gap)])
+qplot(name, gap, data=student2012.sub.summary.gap, ylim=c(-30, 30)) + geom_hline(yintercept=0, colour="grey80") + coord_flip()
+# Gap is interesting
+```

### 2014-06-22 dicook/possessions.Rmd
reflect number of males/females measured

@@
 student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
-student2012.sub.summary <- summarise(group_by(student2012.sub, name), male=mean(PV1MATH[ST04Q01=="Male"], na.rm=T), 
+student2012.sub.summary1 <- summarise(group_by(student2012.sub, name), male=mean(PV1MATH[ST04Q01=="Male"], na.rm=T), 
           female=mean(PV1MATH[ST04Q01=="Female"], na.rm=T))
-student2012.sub.summary.m <- melt(student2012.sub.summary)
-colnames(student2012.sub.summary.m)[2] <- "gender"
-colnames(student2012.sub.summary.m)[3] <- "math"
-qplot(gender, math, data=student2012.sub.summary.m, geom="boxplot") + theme(aspect.ratio=1)
+student2012.sub.summary.m1 <- melt(student2012.sub.summary1)
+colnames(student2012.sub.summary.m1)[2] <- "gender"
+colnames(student2012.sub.summary.m1)[3] <- "math"
+student2012.sub.summary2 <- summarise(group_by(student2012.sub, name), 
+                                     male=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
+                                     female=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
+student2012.sub.summary.m2 <- melt(student2012.sub.summary2)
+colnames(student2012.sub.summary.m2)[2] <- "gender"
+colnames(student2012.sub.summary.m2)[3] <- "prop"
+student2012.sub.summary <- merge(student2012.sub.summary.m1, student2012.sub.summary.m2)
+qplot(gender, math, data=student2012.sub.summary.m1, geom="boxplot") + theme(aspect.ratio=1)
 # Overall there is a gender gap in math, perhaps 10 points, and the high scores are almost exclusively males
-student2012.sub.summary.m$name <- factor(student2012.sub.summary.m$name, 
-      levels=student2012.sub.summary$name[order(student2012.sub.summary$male)])
-student2012.sub.summary.m$gender <- factor(student2012.sub.summary.m$gender, levels=c("female", "male"))
-qplot(name, math, data=student2012.sub.summary.m, colour=gender) + coord_flip()
+student2012.sub.summary$name <- factor(student2012.sub.summary$name, 
+      levels=student2012.sub.summary1$name[order(student2012.sub.summary1$male)])
+student2012.sub.summary$gender <- factor(student2012.sub.summary$gender, levels=c("female", "male"))
+qplot(name, math, data=student2012.sub.summary, colour=gender, size=prop) + coord_flip()
 # Some countries have no gender gap, a few surprises females better than males
-student2012.sub.summary.gap <- summarise(group_by(student2012.sub.summary, name), gap=male-female)
+student2012.sub.summary.gap <- summarise(group_by(student2012.sub.summary1, name), gap=male-female)
+student2012.sub.summary.gapprop <- summarise(group_by(student2012.sub.summary2, name), prop=male-0.5)
+student2012.sub.summary.gap <- merge(student2012.sub.summary.gap, student2012.sub.summary.gapprop)
 student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
       levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$gap)])
-qplot(name, gap, data=student2012.sub.summary.gap, ylim=c(-30, 30)) + geom_hline(yintercept=0, colour="grey80") + coord_flip()
-# Gap is interesting
+qplot(name, gap, data=student2012.sub.summary.gap, ylim=c(-30, 30), size=prop) + 
+   geom_hline(yintercept=0, colour="grey80") + coord_flip()
+# Gap is interesting, muslim countries with no gap is not because only few top girls measured
 ```

### 2014-06-23 alex/maps.Rmd
simple map

Maps
========================================================
```{r load, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(plyr)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

```{r time, warning=FALSE, message=FALSE}
#Relevant variables for hours studied
hours = student2012 %>%
    select(CNT, LMINS, MMINS, OUTHOURS, SMINS, ST57Q01, ST57Q02, ST57Q03, ST57Q04, ST57Q05,
           PV1MATH, PV1SCIE, PV1READ)

hours$MMINS = as.numeric(hours$MMINS)
hours$LMINS = as.numeric(hours$LMINS)
hours$SMINS = as.numeric(hours$SMINS)
hours$ST57Q01 = as.numeric(hours$ST57Q01)
hours$ST57Q02 = as.numeric(hours$ST57Q02)
hours$ST57Q03 = as.numeric(hours$ST57Q03)
hours$ST57Q04 = as.numeric(hours$ST57Q04)
hours$ST57Q05 = as.numeric(hours$ST57Q05)
hours$PV1MATH = as.numeric(hours$PV1MATH)
hours$PV1READ = as.numeric(hours$PV1READ)
hours$PV1SCIE = as.numeric(hours$PV1SCIE)

#change country names
hours$CNT = as.character(hours$CNT)
hours$CNT[hours$CNT=="Serbia"] <- "Republic of Serbia"
hours$CNT[hours$CNT=="Korea"] <- "South Korea"
hours$CNT[hours$CNT=="Chinese Taipei"] <- "Taiwan"
hours$CNT[hours$CNT=="Slovak Republic"] <- "Slovakia"
hours$CNT[hours$CNT=="Russian Federation"] <- "Russia"
hours$CNT[hours$CNT=="Perm(Russian Federation)"] <- "Russia"
hours$CNT[hours$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
hours$CNT[hours$CNT=="China-Shanghai"] <- "China"
hours$CNT[hours$CNT=="China-Macau"] <- "China"
hours$CNT[hours$CNT=="Connecticut (USA)"] <- "United States of America"
hours$CNT[hours$CNT=="Florida (USA)"] <- "United States of America"
hours$CNT[hours$CNT=="Massachusetts (USA)"] <- "United States of America"

#calculate means by CNT
hours = hours %>%
    group_by(CNT) %>%
    summarise(m_OUTHOURS = mean(OUTHOURS, na.rm=T), math = mean(PV1MATH, na.rm=T),
              sci = mean(PV1SCIE, na.rm=T), read = mean(PV1READ, na.rm=T)) %>%
    arrange(m_OUTHOURS) %>% mutate(mscore = (math + sci + read)/3)

hours$CNT = factor(hours$CNT,
                      levels = hours$CNT[order(hours$m_OUTHOURS)])
names(hours)[1] = "name"
print(hours)
p1 = qplot(name, m_OUTHOURS, data = hours, color = m_OUTHOURS, size = mscore) + coord_flip()
```

```{r usefulfn, echo=FALSE, warning=FALSE, message=FALSE}
extractPolygons <- function(shapes) {

  dframe <- ldply(1:length(shapes@polygons), function(i) {
    ob <- shapes@polygons[[i]]@Polygons
    dframe <- ldply(1:length(ob), function(j) {
      x <- ob[[j]]
      co <- x@coords
      data.frame(co, order=1:nrow(co), group=j)
    })
    dframe$region <- i
    dframe$name <- shapes@polygons[[i]]@ID
    dframe$area <- shapes@polygons[[i]]@area
    dframe
  })
  # construct a group variable from both group and polygon:
  dframe$group <- interaction(dframe$region, dframe$group)

  dframe
}

# To get a blank background on map
new_theme_empty <- theme_bw()
new_theme_empty$line <- element_blank()
new_theme_empty$rect <- element_blank()
new_theme_empty$strip.text <- element_blank()
new_theme_empty$axis.text <- element_blank()
new_theme_empty$plot.title <- element_blank()
new_theme_empty$axis.title <- element_blank()
new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
```

```{r map, warning=FALSE, message=FALSE}
library(maps)
library(rworldmap)
world <- getMap(resolution = "low")
world.polys <- extractPolygons(world)
hours.map = left_join(hours, world.polys)

p2 = ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey70")) + 
  geom_polygon(data=hours.map, aes(x=X1, y=X2, order=order, group=group, fill=m_OUTHOURS)) +
  new_theme_empty + theme(legend.position="none")

p1 = ggplotGrob(p1)
p2 + annotation_custom(grob = p1, xmin = -40, xmax = 80, ymin = -110, ymax = 10)
```

### 2014-06-23 dicook/map.Rmd
trying to get maps inserted on other plots

@@
 ```
+
+```{r exploded-map}
+oz <- subset(world.polys, name=="Australia")
+oz.center <- c(mean(oz$X1), mean(oz$X2))
+nz <- subset(world.polys, name=="New Zealand")
+nz.center <- c(mean(nz$X1), mean(nz$X2))
+ind <- subset(world.polys, name=="Indonesia")
+ind.center <- c(mean(ind$X1), mean(ind$X2))
+student2012.sub <- student2012[student2012$CNT=="Australia"|student2012$CNT=="New Zealand"|student2012$CNT=="Indonesia", 
+                               c(1, 6, 501)]
+colnames(student2012.sub)[1] <- "name"
+colnames(student2012.sub)[3] <- "math"
+p1 <- qplot(math, name, data=student2012.sub, xlim=c(0,1000))
+summarise(group_by(student2012.sub, name), math=mean(math, na.rm=T))
+oz$X1 <- scale(oz$X1, center=T, scale=T)
+oz$X2 <- scale(oz$X2, center=T, scale=T)
+nz$X1 <- scale(nz$X1, center=T, scale=T)
+nz$X2 <- scale(nz$X2, center=T, scale=T)
+ind$X1 <- scale(ind$X1, center=T, scale=T)
+ind$X2 <- scale(ind$X2, center=T, scale=T)
+p2 <- qplot(X1, X2, data=oz, group=group, order=order, geom="polygon") + new_theme_empty
+p2 = ggplotGrob(p2)      	
+p3 <- qplot(X1, X2, data=ind, group=group, order=order, geom="polygon") + new_theme_empty
+p3 = ggplotGrob(p3)        
+p4 <- qplot(X1, X2, data=nz, group=group, order=order, geom="polygon") + new_theme_empty
+p4 = ggplotGrob(p4)        
+p1 + annotation_custom(grob = p2, xmin = 900, xmax = 1000, ymin = 0.8, ymax = 1.2) +
+  annotation_custom(grob = p3, xmin = 900, xmax = 1000, ymin = 1.5, ymax = 2.5) + 
+  annotation_custom(grob = p4, xmin = 900, xmax = 1000, ymin = 2.5, ymax = 3.5)
+```

### 2014-06-23 ijlyttle/factor_explorer.Rmd
adds color bands

@@
 ```{r load_data}
-sets <- c("item", "parent", "school", "scoredItem", "student")
+#sets <- c("item", "parent", "school", "scoredItem", "student")
+sets <- "student"
 
@@
 # there's a clever way to do this, but beyond me for naw
-dict_item2012 <- fn_make_df(item2012dict) 
-dict_parent2012 <- fn_make_df(parent2012dict) 
-dict_school2012 <- fn_make_df(school2012dict) 
-dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
+# dict_item2012 <- fn_make_df(item2012dict) 
+# dict_parent2012 <- fn_make_df(parent2012dict) 
+# dict_school2012 <- fn_make_df(school2012dict) 
+# dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
 dict_student2012 <- fn_make_df(student2012dict)
@@
   choices = levels(student2012$CNT),
-  selected = c("United States of America", "Canada"),
+  selected = c("United States of America", "Peru", "China-Shanghai"),
   multiple = TRUE
@@
   choices = var_names_factor,
-  selected = "ST04Q01"
+  selected = "ST27Q03"
 )
@@
   choices = var_names_factor,
-  selected = "ST26Q03"
+  selected = "ST28Q01"
 )
@@
   ) + 
+  geom_rect(
+    aes(fill = CNT),
+    xmin = -Inf, xmax = Inf, 
+    ymin = -Inf, ymax = Inf,
+    alpha = 0.5,
+    data = data.frame(CNT = unique(rct_data()$CNT))
+  ) + 
   geom_violin(
@@
   ) + 
+#   geom_boxplot(
+#     aes_string(
+#       x = input$x_factor,
+#       y = input$subject,
+#       group = input$fill_factor
+#     ),
+#     fill = "white",
+#     alpha = 1,
+#   ) + 
   geom_boxplot(
@@
       y = input$subject,
-      fill = input$fill_factor
+      alpha = input$fill_factor
     ),
-    alpha = 0.75
+    fill = "blue"
   ) + 
-#   geom_dotplot(
-#     aes_string(
-#       x = input$x_factor,
-#       y = input$subject,
-#       fill = input$fill_factor
-#     ),
-#     binaxis = "y", 
-#     stackdir = "center", 
-#     position = "dodge") + 
   facet_grid(CNT ~ .) + 
@@
   scale_y_continuous(limits = c(0, 1000), name = rct_labels()[["y"]]) +
-  scale_fill_brewer(type = "seq", name = rct_labels()[["fill"]]) 
+  scale_alpha_discrete(range = c(0.1, 0.9), na.value = 0, name = rct_labels()[["fill"]]) +  
+  scale_fill_brewer(type = "seq", palette = "BuGn", guide = FALSE) 
 })
@@
   ggplot(
+    data = rct_data()
+  ) +
+  geom_rect(
+    aes(fill = CNT),
+    xmin = -Inf, xmax = Inf, 
+    ymin = -Inf, ymax = Inf,
+    alpha = 0.5,
+    data = data.frame(CNT = unique(rct_data()$CNT))
+  ) + 
+  geom_bar(
     aes_string(
       x = input$x_factor,
-      fill = input$fill_factor
+      group = input$fill_factor
     ),
-    data = rct_data()
+    position = "dodge", 
+    fill = "white", 
+    alpha = 1,
+    color = "black"
   ) +   
-  geom_bar(position = "dodge", color = "black") +   
+  geom_bar(
+    aes_string(
+      x = input$x_factor,
+      alpha = input$fill_factor
+    ),
+    position = "dodge", 
+    fill = "blue"
+  ) +   
   facet_grid(CNT ~ ., scales = "free_y") + 
@@
   scale_y_continuous(name = "number of students") +
-  scale_fill_brewer(type = "seq", name = rct_labels()[["fill"]]) 
+  scale_alpha_discrete(range = c(0.1, 0.9), na.value = 0, name = rct_labels()[["fill"]]) +  
+  scale_fill_brewer(type = "seq", palette = "BuGn", guide = FALSE) 
 })

### 2014-06-23 dicook/map.Rmd

@@
 library(lubridate)
-library(ggvis)
+#library(ggvis)
 library(maps)
@@
 library(YaleToolkit)
+student2012.sub <- student2012[, c(1:7, seq(501, 550, 5))]
+colnames(student2012.sub)[1] <- "name"
+student2012.sub$name <- as.character(student2012.sub$name)
+# Check mismatches of names
+# unique(anti_join(student2012.sub, world.polys)[1])
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
+student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
+student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
+
+# Only need one of the plausible values, checked they are effectively identical
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1MACC <- as.numeric(student2012.sub$PV1MACC)
+student2012.sub$PV1MACQ <- as.numeric(student2012.sub$PV1MACQ)
+student2012.sub$PV1MACS <- as.numeric(student2012.sub$PV1MACS)
+student2012.sub$PV1MACU <- as.numeric(student2012.sub$PV1MACU)
+student2012.sub$PV1MAPE <- as.numeric(student2012.sub$PV1MAPE)
+student2012.sub$PV1MAPF <- as.numeric(student2012.sub$PV1MAPF)
+student2012.sub$PV1MAPI <- as.numeric(student2012.sub$PV1MAPI)
+student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
+student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub.summary <- summarise(group_by(student2012.sub[,c(1,8:17)], name), math=mean(PV1MATH, na.rm=T), 
+                                  mCC=mean(PV1MACC, na.rm=T), mCQ=mean(PV1MACQ, na.rm=T), 
+                                  mCS=mean(PV1MACS, na.rm=T), mCU=mean(PV1MACU, na.rm=T), 
+                                  mPE=mean(PV1MAPE, na.rm=T), mPF=mean(PV1MAPF, na.rm=T), 
+                                  mPI=mean(PV1MAPI, na.rm=T), read=mean(PV1READ, na.rm=T), 
+                                  science=mean(PV1SCIE, na.rm=T), 
+                                     mathr=diff(range(PV1MATH, na.rm=T)), 
+                                  mCCr=diff(range(PV1MACC, na.rm=T)), mCQr=diff(range(PV1MACQ, na.rm=T)), 
+                                  mCSr=diff(range(PV1MACS, na.rm=T)), mCUr=diff(range(PV1MACU, na.rm=T)), 
+                                  mPEr=diff(range(PV1MAPE, na.rm=T)), mPFr=diff(range(PV1MAPF, na.rm=T)), 
+                                  mPIr=diff(range(PV1MAPI, na.rm=T)), readr=diff(range(PV1READ, na.rm=T)), 
+                                  sciencer=diff(range(PV1SCIE, na.rm=T)))
+colnames(student2012.sub.summary)[3:9] <- c("Change", "Quantity", "Spatial", "Data", "Employ", "Formulate", "Interpret")
 gpairs(student2012.sub.summary[, c(2, 10, 11)])

### 2014-06-25 alex/reading_science_scores.Rmd
a bunch of stuff

looking at reading and science scores
==============

Since everyone has been primarily focusing on math scores I looked at science and reading scores a bit.
```{r load, echo=FALSE, results='hide'}
library(ggplot2)
library(plyr)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

```

```{r readplots}
student2012.sub = student2012 %>% select(CNT, SCHOOLID, STIDSTD, OECD, PV1SCIE, PV1READ, ST04Q01)
student2012.sub$PV1SCIE = as.numeric(student2012.sub$PV1SCIE)
student2012.sub$PV1READ = as.numeric(student2012.sub$PV1READ)
student2012.sub$CNT = as.character(student2012.sub$CNT)
student2012.sub$CNT[student2012.sub$CNT=="Serbia"] <- "Republic of Serbia"
student2012.sub$CNT[student2012.sub$CNT=="Korea"] <- "South Korea"
student2012.sub$CNT[student2012.sub$CNT=="Chinese Taipei"] <- "Taiwan"
student2012.sub$CNT[student2012.sub$CNT=="Slovak Republic"] <- "Slovakia"
student2012.sub$CNT[student2012.sub$CNT=="Russian Federation"] <- "Russia"
student2012.sub$CNT[student2012.sub$CNT=="Perm(Russian Federation)"] <- "Russia"
student2012.sub$CNT[student2012.sub$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
student2012.sub$CNT[student2012.sub$CNT=="China-Shanghai"] <- "China"
student2012.sub$CNT[student2012.sub$CNT=="China-Macau"] <- "China"
student2012.sub$CNT[student2012.sub$CNT=="Connecticut (USA)"] <- "United States of America"
student2012.sub$CNT[student2012.sub$CNT=="Florida (USA)"] <- "United States of America"
student2012.sub$CNT[student2012.sub$CNT=="Massachusetts (USA)"] <- "United States of America"

df = student2012.sub %>% group_by(CNT, ST04Q01) %>% summarise(msci = mean(PV1SCIE),
                                                         mread = mean(PV1READ)) 
male = df %>% filter(ST04Q01 == "Male")
female = df %>% filter(ST04Q01 == "Female")

male_female = data.frame(CNT = male$CNT, msci_m = male$msci, mread_m = male$mread,
                         msci_f = female$msci, mread_f = female$mread)
male_female$sci_diff = male_female$msci_m - male_female$msci_f
male_female$read_diff = male_female$mread_m - male_female$mread_f

read_diff = male_female %>% select(CNT, read_diff)
read_diff$CNT = factor(read_diff$CNT,
                       levels = read_diff$CNT[order(read_diff$read_diff)])
qplot(CNT, read_diff, data = read_diff) + coord_flip()

```

```{r sciplots}
sci_diff =male_female %>% select(CNT, sci_diff)
sci_diff$CNT = factor(sci_diff$CNT,
                       levels = sci_diff$CNT[order(sci_diff$sci_diff)])
qplot(CNT, sci_diff, data = sci_diff) + coord_flip()
```

### 2014-06-25 alex/school_poverty.Rmd
a bunch of stuff

Grouping schools by poverty
========================================================

```{r load, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(reshape2)
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

#choose which of the 5 PVs to use for each subject
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])
```

```{r usefulfn, echo=FALSE, message=FALSE, warning=FALSE}
fix_country = function(df) {
df$CNT = as.character(df$CNT)
df$CNT[df$CNT=="Serbia"] <- "Republic of Serbia"
df$CNT[df$CNT=="Korea"] <- "South Korea"
df$CNT[df$CNT=="Chinese Taipei"] <- "Taiwan"
df$CNT[df$CNT=="Slovak Republic"] <- "Slovakia"
df$CNT[df$CNT=="Russian Federation"] <- "Russia"
df$CNT[df$CNT=="Perm(Russian Federation)"] <- "Russia"
df$CNT[df$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
df$CNT[df$CNT=="China-Shanghai"] <- "China"
df$CNT[df$CNT=="China-Macau"] <- "China"
df$CNT[df$CNT=="Connecticut (USA)"] <- "United States of America"
df$CNT[df$CNT=="Florida (USA)"] <- "United States of America"
df$CNT[df$CNT=="Massachusetts (USA)"] <- "United States of America"

df
}

```

```{r school}
school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC22Q01,
                                       SC22Q02, SC22Q03, SC22Q04,
                                       SC22Q05, SC22Q06, SC22Q07,
                                       SC22Q08, SC22Q09, SC22Q10,
                                       SC22Q11, SC22Q12, SC22Q13)
student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, ST08Q01, ST09Q01)
school2012.sub = fix_country(school2012.sub)
student2012.sub = fix_country(student2012.sub)

truancy = student2012.sub %>% group_by(CNT, SCHOOLID) %>% left_join(school2012.sub, by = c("CNT", "SCHOOLID"))
```

```{r truancy}
#school perception of truancy vs math scores
#percentage of schools reporting high truancy issues
df = school2012.sub %>% group_by(CNT) %>% summarise( truancy = sum(SC22Q01 == "A lot", na.rm=T) / length(CNT) ) %>% arrange(desc(truancy)) %>% 
  left_join(
    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
  )
#russia and china are outliers
qplot(x = truancy, y = m_math, data = df) + stat_smooth(method="lm", se = FALSE)

#student self reported truancy within the last two weeeks
#percentage of high truancy students
df2 = student2012.sub %>% group_by(CNT) %>% summarise(truancy_student = sum(ST08Q01 == "Five or more times  " | ST08Q01 == "Three or four times  " | ST08Q01 == "One or two times  ", na.rm=T) / length(CNT)) %>% arrange(desc(truancy_student)) %>% 
  left_join(
    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
  )

qplot(truancy_student, m_math, data = df2)
```

```{r drugs}
#percentage of schools reporting drug problems
drugs = school2012.sub %>% group_by(CNT) %>% summarise(drug = sum(SC22Q07 == "A lot" | SC22Q07 == "To some extent", na.rm=T) / length(CNT)) %>% arrange(desc(drug)) %>%
   left_join(
    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
  )

school2012.sub %>% group_by(CNT) %>% summarise(drug = sum(SC22Q07 == "A lot", na.rm=T) / length(CNT)) %>% arrange(desc(drug))
```

```{r competition}
comp = school2012 %>% 
  select(CNT, SCHOOLID, SC04Q01) %>% 
  group_by(CNT) %>%
  summarise(other = 
              sum(SC04Q01 == "Two or More", na.rm=T) / length(CNT)) %>% 
  arrange(desc(other)) %>%
   left_join(
    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
  )

qplot(m_math, other, data = comp)
```

### 2014-06-25 alex/single_parents.Rmd
a bunch of stuff

Single parents
==============

```{r load, echo=FALSE, results='hide'}
library(ggplot2)
library(plyr)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

```

Grandparents at home doesn't have much of an effect on scores.  If we look at the average test scores (averaging math, science and reading), there is an obvious difference between have both parents versus have neither parent.  However for single parents, having only a father will generally mean a much lower score.
```{r singleparents}
athome = student2012 %>% select(CNT, SCHOOLID, STIDSTD, OECD, PV1MATH, PV1SCIE, PV1READ,
                                ST11Q01, ST11Q02, ST11Q03,
                                ST11Q04, ST11Q05, ST11Q06)
athome$PV1MATH = as.numeric(athome$PV1MATH)
athome$PV1SCIE = as.numeric(athome$PV1SCIE)
athome$PV1READ = as.numeric(athome$PV1READ)
athome$CNT = as.character(athome$CNT)
athome$CNT[athome$CNT=="Serbia"] <- "Republic of Serbia"
athome$CNT[athome$CNT=="Korea"] <- "South Korea"
athome$CNT[athome$CNT=="Chinese Taipei"] <- "Taiwan"
athome$CNT[athome$CNT=="Slovak Republic"] <- "Slovakia"
athome$CNT[athome$CNT=="Russian Federation"] <- "Russia"
athome$CNT[athome$CNT=="Perm(Russian Federation)"] <- "Russia"
athome$CNT[athome$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
athome$CNT[athome$CNT=="China-Shanghai"] <- "China"
athome$CNT[athome$CNT=="China-Macau"] <- "China"
athome$CNT[athome$CNT=="Connecticut (USA)"] <- "United States of America"
athome$CNT[athome$CNT=="Florida (USA)"] <- "United States of America"
athome$CNT[athome$CNT=="Massachusetts (USA)"] <- "United States of America"

#########################################
#overall
overall = athome %>% group_by(CNT) %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_overall = (mmath + mread + msci) / 3) %>% arrange(desc(overall_overall))

#both parents
both = athome %>% group_by(CNT) %>% filter(ST11Q01 == "Yes" & ST11Q02 == "Yes") %>%
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_both = (mmath + mread + msci) / 3) %>% arrange(desc(overall_both))

#single mothers
mother = athome %>% group_by(CNT) %>% filter(ST11Q01 == "Yes" & ST11Q02 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_mother = (mmath + mread + msci) / 3) %>% arrange(desc(overall_mother))

#single fathers
father = athome %>% group_by(CNT) %>% filter(ST11Q01 == "No" & ST11Q02 == "Yes") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_father = (mmath + mread + msci) / 3) %>% arrange(desc(overall_father))

#neither mother nor father
neither = athome %>% group_by(CNT) %>% filter(ST11Q01 == "No" & ST11Q02 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_neither = (mmath + mread + msci) / 3) %>% arrange(desc(overall_neither))

#grandparents
grand = athome %>% group_by(CNT) %>% filter(ST11Q05 == "Yes") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_grand = (mmath + mread + msci) / 3) %>% arrange(desc(overall_grand))

#no grandparents
nogrand= athome %>% group_by(CNT) %>% filter(ST11Q05 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_grand = (mmath + mread + msci) / 3) %>% arrange(desc(overall_grand))

#########################################
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "neither"
#qplot(x = parents, y = PV1MATH, geom = "boxplot", facets = ~CNT, data = athome) + coord_flip()

single_parents = merge(mother[,c(1,6)], father[,c(1,6)], by.x = c("CNT"), by.y = "CNT")
single_parents$diff = single_parents$overall_mother - single_parents$overall_father
single_parents = single_parents %>% arrange(desc(diff))
single_parents$CNT = factor(single_parents$CNT,
                            levels = single_parents$CNT[order(single_parents$diff)])
single_parents %>% left_join(overall %>% select(CNT, overall_overall))
df = merge(single_parents, overall %>% select(CNT, overall_overall), by.x = 'CNT', by.y = 'CNT')
df = merge(df, athome %>% group_by(CNT) %>% summarise(oecd = OECD[1]))
#single mother scores - single father scores
qplot(CNT, diff, data = df, color = oecd) + coord_flip()
```

Comparing having both parents to just having a mother, it appears there are many countries where there isn't a huge difference in test scores.
```{r single-both, eval=FALSE}
#both parents to single mothers only
df2 = merge(mother[,c(1,6)], both[, c(1,6)], by.x = c("CNT"), by.y = c("CNT"))
df2$diff = df2$overall_both - df2$overall_mother
df2 %>% arrange(desc(diff))
```

### 2014-06-25 alex/albania.Rmd
albania outhours - strange vertical line

--- 2014-06-25 alex/school_poverty.Rmd
@@
-Grouping schools by poverty
-========================================================
+albania is weird
+=============
 
-```{r load, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
-library(reshape2)
+```{r load, echo=FALSE, results='hide'}
 library(ggplot2)
@@
 
-#choose which of the 5 PVs to use for each subject
 student2012$math = as.numeric(student2012[, "PV1MATH"])
@@
 
-```{r usefulfn, echo=FALSE, message=FALSE, warning=FALSE}
-fix_country = function(df) {
-df$CNT = as.character(df$CNT)
-df$CNT[df$CNT=="Serbia"] <- "Republic of Serbia"
-df$CNT[df$CNT=="Korea"] <- "South Korea"
-df$CNT[df$CNT=="Chinese Taipei"] <- "Taiwan"
-df$CNT[df$CNT=="Slovak Republic"] <- "Slovakia"
-df$CNT[df$CNT=="Russian Federation"] <- "Russia"
-df$CNT[df$CNT=="Perm(Russian Federation)"] <- "Russia"
-df$CNT[df$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
-df$CNT[df$CNT=="China-Shanghai"] <- "China"
-df$CNT[df$CNT=="China-Macau"] <- "China"
-df$CNT[df$CNT=="Connecticut (USA)"] <- "United States of America"
-df$CNT[df$CNT=="Florida (USA)"] <- "United States of America"
-df$CNT[df$CNT=="Massachusetts (USA)"] <- "United States of America"
+```{r outhours}
+student.sub = student2012 %>% 
+  select(CNT, OUTHOURS, math) %>% filter(!is.na(OUTHOURS))
 
-df
-}
+qplot(x = OUTHOURS, y = math, 
+      data = student.sub, facets = ~CNT, 
+      alpha = I(0.05))
 
@@
 
-```{r school}
-school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC22Q01,
-                                       SC22Q02, SC22Q03, SC22Q04,
-                                       SC22Q05, SC22Q06, SC22Q07,
-                                       SC22Q08, SC22Q09, SC22Q10,
-                                       SC22Q11, SC22Q12, SC22Q13)
-student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, ST08Q01, ST09Q01)
-school2012.sub = fix_country(school2012.sub)
-student2012.sub = fix_country(student2012.sub)
+```{r albania}
+qplot(x = OUTHOURS, y = math, 
+      data = student.sub %>% filter(CNT == "Albania"), 
+      alpha = I(0.05))
 
-truancy = student2012.sub %>% group_by(CNT, SCHOOLID) %>% left_join(school2012.sub, by = c("CNT", "SCHOOLID"))
 ```
-
-```{r truancy}
-#school perception of truancy vs math scores
-#percentage of schools reporting high truancy issues
-df = school2012.sub %>% group_by(CNT) %>% summarise( truancy = sum(SC22Q01 == "A lot", na.rm=T) / length(CNT) ) %>% arrange(desc(truancy)) %>% 
-  left_join(
-    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
-  )
-#russia and china are outliers
-qplot(x = truancy, y = m_math, data = df) + stat_smooth(method="lm", se = FALSE)
-
-#student self reported truancy within the last two weeeks
-#percentage of high truancy students
-df2 = student2012.sub %>% group_by(CNT) %>% summarise(truancy_student = sum(ST08Q01 == "Five or more times  " | ST08Q01 == "Three or four times  " | ST08Q01 == "One or two times  ", na.rm=T) / length(CNT)) %>% arrange(desc(truancy_student)) %>%
-  left_join(
-    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
-  )
-
-qplot(truancy_student, m_math, data = df2)
-```
-
-```{r drugs}
-#percentage of schools reporting drug problems
-drugs = school2012.sub %>% group_by(CNT) %>% summarise(drug = sum(SC22Q07 == "A lot" | SC22Q07 == "To some extent", na.rm=T) / length(CNT)) %>% arrange(desc(drug)) %>%
-   left_join(
-    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
-  )
-
-school2012.sub %>% group_by(CNT) %>% summarise(drug = sum(SC22Q07 == "A lot", na.rm=T) / length(CNT)) %>% arrange(desc(drug))
-```
-
-```{r competition}
-comp = school2012 %>% 
-  select(CNT, SCHOOLID, SC04Q01) %>% 
-  group_by(CNT) %>%
-  summarise(other = 
-              sum(SC04Q01 == "Two or More", na.rm=T) / length(CNT)) %>% 
-  arrange(desc(other)) %>%
-   left_join(
-    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
-  )
-
-qplot(m_math, other, data = comp)
-```

### 2014-06-27 alex/school_poverty.Rmd

@@
 
+Competition between schools
 ```{r competition}
@@
 ```
+
+Math teacher shortages
+```{r shortages}
+school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
+student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci)
+school2012.sub = fix_country(school2012.sub)
+student2012.sub = fix_country(student2012.sub)
+
+scores_by_school = student2012.sub %>% group_by(CNT, SCHOOLID) %>% summarise(mmath = mean(math), mread = mean(read), msci = mean(sci))
+
+df = scores_by_school %>% left_join(school2012.sub)
+qplot(SC14Q02, mmath2, data = df %>% group_by(CNT, SC14Q02) %>% summarise(mmath2 = mean(mmath)), facets = ~CNT)
+qplot(SC14Q10, mmath2, data = df %>% group_by(CNT, SC14Q10) %>% summarise(mmath2 = mean(mmath)), facets = ~CNT)
+```
+
+looking at just the US
+```{r shortages1}
+library(ggplot2)
+qplot(SC14Q02, mmath, data = df %>% group_by(CNT, SC14Q02) %>% 
+        filter(CNT == "United States of America"), geom="boxplot")
+```
+
+```{r shortages2}
+library(ggplot2)
+df2 = df%>% select(CNT, SCHOOLID, mmath, mread, msci, SC14Q02,
+                  SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11,
+                  SC14Q12, SC14Q13)
+for(i in 6:13) {
+  df2[,i] = as.numeric(df2[,i])
+}
+
+df2$sum = apply(df2[,6:13], 1, sum, na.rm=T)
+qplot(x = sum, y = m2, data = df2 %>% group_by(CNT, sum) %>% summarise(m2 = mean(mmath)), facets=~CNT)
+
+qplot(x = pSC14Q07, mmath2, data = 
+df2 %>% group_by(CNT) %>% 
+  summarise(pSC14Q07 = sum(SC14Q07 == 4, na.rm=T) / length(CNT),
+            mmath2 = mean(mmath)) %>% 
+  arrange(pSC14Q07)
+)
+```

### 2014-06-27 alex/2020.Rmd

@@
         filter(!is.na(OUTHOURS)), 
-      alpha = I(0.1)) + stat_smooth()
+      alpha = I(0.05)) + stat_smooth()
 qplot(x = OUTHOURS, y = read, 
@@
         filter(!is.na(OUTHOURS)), 
-      alpha = I(0.1) ) + stat_smooth()
+      alpha = I(0.05) ) + stat_smooth()
 qplot(x = OUTHOURS, y = sci, 
@@
         filter(!is.na(OUTHOURS)), 
-      alpha = I(0.1) ) + stat_smooth()
+      alpha = I(0.05)) + stat_smooth()
 ```

### 2014-06-27 alex/schools.Rmd

Schools
========================================================
```{r load, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])
```

mean number of possessions grouped by CNT and SCHOOLID
```{r countpossessions, warning=FALSE, message=FALSE}
student2012.sub <- student2012[, c(1:7, 44:57, 61:66, seq(501, 550, 5))]
colnames(student2012.sub)[1] <- "name"
for (i in 8:21) {
  student2012.sub[,i] <- as.character(student2012.sub[,i])
  student2012.sub[is.na(student2012.sub[,i]), i] <- ""
  student2012.sub[student2012.sub[,i]=="Yes",i] <- "1"
  student2012.sub[student2012.sub[,i]=="No",i] <- "0"
  student2012.sub[,i] <- as.numeric(student2012.sub[,i])
}
for (i in 22:26) {
  student2012.sub[,i] <- as.character(student2012.sub[,i])
  student2012.sub[is.na(student2012.sub[,i]), i] <- ""
  student2012.sub[student2012.sub[,i]=="None",i] <- "0"
  student2012.sub[student2012.sub[,i]=="One",i] <- "1"
  student2012.sub[student2012.sub[,i]=="Two",i] <- "2"
  student2012.sub[student2012.sub[,i]=="Three or more",i] <- "3"
  student2012.sub[,i] <- as.numeric(student2012.sub[,i])
}
student2012.sub[,27] <- as.character(student2012.sub[,27])
student2012.sub[is.na(student2012.sub[,27]), 27] <- ""
student2012.sub[student2012.sub[,27]=="0-10 books ",27] <- "0"
student2012.sub[student2012.sub[,27]=="11-25 books ",27] <- "1"
student2012.sub[student2012.sub[,27]=="26-100 books ",27] <- "2"
student2012.sub[student2012.sub[,27]=="101-200 books ",27] <- "3"
student2012.sub[student2012.sub[,27]=="201-500 books ",27] <- "4"
student2012.sub[student2012.sub[,27]=="More than 500 books",27] <- "5"
student2012.sub[,27] <- as.numeric(student2012.sub[,27])
student2012.sub$numposs <- apply(student2012.sub[,8:21], 1, sum, na.rm=T)
student2012.sub$numedposs <- apply(student2012.sub[,22:27], 1, sum, na.rm=T)

library(ggplot2)
library(dplyr)
student2012.sub$PV1MATH = as.numeric(student2012.sub$PV1MATH)
qplot(mnumposs, mmath, data = student2012.sub %>% select(name, SCHOOLID, numposs, PV1MATH) %>% group_by(name, SCHOOLID) %>% summarise(mmath = mean(PV1MATH), mnumposs = mean(numposs)), facets = ~name)

```

Hard to see in this document but if you zoom in there's a pretty clear positive correlation.

### 2014-06-27 alex/albania.Rmd
albania

@@
-albania is weird
+Equality for all: Only in student2012.sub
 =============
-
-```{r load, echo=FALSE, results='hide'}
+```{r load, echo=FALSE, results='hide', warning = FALSE, message=FALSE}
 library(ggplot2)
@@
 setwd("..")
-sets <- c("item", "parent", "school", "scoredItem", "student")
+sets <- c("school", "student")
 
@@
 fn_build <- function(file_name) {
-
     template <- c("2012.rda", "2012dict.rda")
-
     file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
@@
 
-# there's a clever way to do this, but beyond me for naw
-dict_item2012 <- fn_make_df(item2012dict)
-dict_parent2012 <- fn_make_df(parent2012dict)
 dict_school2012 <- fn_make_df(school2012dict)
-dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
 dict_student2012 <- fn_make_df(student2012dict)
-
-rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
+rm(fn_make_df, fn_build, sets, school2012dict, student2012dict)
 
@@
 
-```{r outhours}
-student.sub = student2012 %>% 
-  select(CNT, OUTHOURS, math) %>% filter(!is.na(OUTHOURS))
+```{r vars}
+student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
 
-qplot(x = OUTHOURS, y = math, 
-      data = student.sub, facets = ~CNT, 
-      alpha = I(0.05))
+for (i in 13:26) {
+    student2012.sub[, i] <- as.character(student2012.sub[, i])
+    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
+    student2012.sub[student2012.sub[, i] == "Yes", i] <- "1"
+    student2012.sub[student2012.sub[, i] == "No", i] <- "0"
+    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
+}
+for (i in 27:31) {
+    student2012.sub[, i] <- as.character(student2012.sub[, i])
+    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
+    student2012.sub[student2012.sub[, i] == "None", i] <- "0"
+    student2012.sub[student2012.sub[, i] == "One", i] <- "1"
+    student2012.sub[student2012.sub[, i] == "Two", i] <- "2"
+    student2012.sub[student2012.sub[, i] == "Three or more", i] <- "3"
+    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
+}
+student2012.sub[, 32] <- as.character(student2012.sub[, 32])
+student2012.sub[is.na(student2012.sub[,32]), 32] <- ""
+student2012.sub[student2012.sub[, 32] == "0-10 books ", 32] <- "0"
+student2012.sub[student2012.sub[, 32] == "11-25 books ", 32] <- "1"
+student2012.sub[student2012.sub[, 32] == "26-100 books ", 32] <- "2"
+student2012.sub[student2012.sub[, 32] == "101-200 books ", 32] <- "3"
+student2012.sub[student2012.sub[, 32] == "201-500 books ", 32] <- "4"
+student2012.sub[student2012.sub[, 32] == "More than 500 books", 32] <- "5"
+student2012.sub[, 32] <- as.numeric(student2012.sub[, 32])
 
+#aggregate possessions
+student2012.sub$numposs <- apply(student2012.sub[, 13:26], 1, sum, na.rm = T)
+student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
+
+#separate albania
+student2012.sub$OECD = as.character(student2012.sub$OECD)
+student2012.sub[student2012.sub$CNT == "Albania", ]$OECD = "Albania"
 ```
 
-```{r albania}
-qplot(x = OUTHOURS, y = math, 
-      data = student.sub %>% filter(CNT == "Albania"), 
-      alpha = I(0.05))
-
+```{r possessions, echo=FALSE}
+qplot()
 ```

### 2014-06-27 alex/albania.Rmd
glory to albania

@@
 ```{r possessions, echo=FALSE}
-qplot()
+library(reshape2)
+df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+df$group = paste0(df$CNT, df$variable)
+qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group)
+
+qplot(x = numposs, y = mmath, data = student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), facets =~OECD, geom = "line", group = CNT) + ylab("Math Scores")
 ```
+
+```{r gender, echo=FALSE}
+
+```

### 2014-06-27 dicook/gender.Rmd
colorblind testing

Exploring gender and scores
========================================================

```{r load_libraries, message=FALSE}
library(stringr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
library(rworldmap)
library (grid)    
library (scales)  
```

```{r load_data, echo=FALSE}
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){

  template <- c("2012.rda", "2012dict.rda")

  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path(".", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
#rm(fn_build, sets)
```

```{r modify_dictionaries, echo=FALSE}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict)

# clean
#rm(fn_make_df)
#rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

```{r usefulfn, echo=FALSE}
extractPolygons <- function(shapes) {

  dframe <- ldply(1:length(shapes@polygons), function(i) {
    ob <- shapes@polygons[[i]]@Polygons
    dframe <- ldply(1:length(ob), function(j) {
      x <- ob[[j]]
      co <- x@coords
      data.frame(co, order=1:nrow(co), group=j)
    })
    dframe$region <- i
    dframe$name <- shapes@polygons[[i]]@ID
    dframe
  })
  # construct a group variable from both group and polygon:
  dframe$group <- interaction(dframe$region, dframe$group)

  dframe
}

# To get a blank background on map
new_theme_empty <- theme_bw()
new_theme_empty$line <- element_blank()
new_theme_empty$rect <- element_blank()
new_theme_empty$strip.text <- element_blank()
new_theme_empty$axis.text <- element_blank()
new_theme_empty$plot.title <- element_blank()
new_theme_empty$axis.title <- element_blank()
new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
```
```
```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
student2012.sub <- student2012[, c(1, 12, 501, 541, 546)]
colnames(student2012.sub)[1] <- "name"
student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
  scale_size("Prop male") +
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
# Test colors
#qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"=dichromat("skyblue"), "female"=dichromat("pink"), "none"=dichromat("lightgreen"))) +
#  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#ggsave("gendermathgap.pdf", width=7, height=14)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
  xlab("") +
  geom_point(aes(y=maxfemale), colour="pink") + 
  scale_y_continuous("High Math") + coord_flip() + theme_bw()
#ggsave("gendermathtop.pdf", width=3.5, height=7)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
  xlab("") +
  geom_point(aes(y=minfemale), colour="pink") + 
  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
#ggsave("gendermathbottom.pdf", width=3.5, height=7)
#
student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Reading Score Gap") + 
  scale_size("Prop male") +
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
#ggsave("genderreadgap.pdf", width=5, height=8)
student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Science Score Gap") + 
  scale_size("Prop male") +
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
#ggsave("gendergap.pdf", width=10, height=14)

````

```{r maps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}
# Extract map polygons for modern world
world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
# Now add data
student2012.sub.summary.gap$name <- as.character(student2012.sub.summary.gap$name)
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Serbia"] <- "Republic of Serbia"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Korea"] <- "South Korea"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Chinese Taipei"] <- "Taiwan"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Slovak Republic"] <- "Slovakia"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Russian Federation"] <- "Russia"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Perm(Russian Federation)"] <- "Russia"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="China-Shanghai"] <- "China"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="China-Macau"] <- "China"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Connecticut (USA)"] <- "United States of America"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Florida (USA)"] <- "United States of America"
student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Massachusetts (USA)"] <- "United States of America"

student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("gendermathmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("genderreadmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("gendermap.pdf", width=12, height=8)
#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
#  scale_fill_manual("Significant", values=c("male"=dichromat("skyblue"), "female"=dichromat("pink"), "none"=dichromat("lightgreen"))) +
#  new_theme_empty + theme(legend.position="none")
```

### 2014-06-27 alex/albania.Rmd
albania is awesome

--- 2014-06-27 alex/albania.Rmd
@@
 
-```{r possessions, echo=FALSE}
-qplot()
+```{r possessions, echo=FALSE, fig.width=15}
+library(reshape2)
+df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+df$group = paste0(df$CNT, df$variable)
+qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group)
+
+qplot(x = numposs, y = mmath, data = student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), facets =~OECD, geom = "line", group = CNT) + ylab("Math Scores")
 ```
+
+```{r gender, echo=FALSE}
+
+```

### 2014-06-27 alex/albania.Rmd

@@
 ```{r gender, echo=FALSE}
+df = student2012.sub %>% group_by(CNT, ST04Q01, OECD) %>% summarise(mmath = mean(math), mread = mean(read), msci = mean(sci))
 
 ```
+
+```{r OUTHOURS, echo=FALSE, fig.width=15}
+#qplot(x = OUTHOURS, data = student2012.sub, facets = ~OECD, group = CNT) 
+#ggplot(subset(student2012.sub, OECD == "Albania"), aes(x = OUTHOURS)) + geom_density(fill = "gray", alpha = I(0.7))
+#ggplot(subset(student2012.sub, OECD == "OECD"), aes(x = OUTHOURS, fill = CNT)) + 
+#  geom_density(alpha = I(0.1)) + scale_fill_manual(values = rep("gray", 34))
+ggplot(subset(student2012.sub, !is.na(OUTHOURS)), aes(x = OUTHOURS, fill = CNT)) + geom_density(alpha = I(0.4)) + facet_wrap(~OECD)
+```

### 2014-06-27 dicook/school_matters.Rmd
start on schools submission

Grouping schools by poverty
========================================================

```{r load, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(reshape2)
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

#choose which of the 5 PVs to use for each subject
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])
```

```{r usefulfn, echo=FALSE, message=FALSE, warning=FALSE}
fix_country = function(df) {
df$CNT = as.character(df$CNT)
df$CNT[df$CNT=="Serbia"] <- "Republic of Serbia"
df$CNT[df$CNT=="Korea"] <- "South Korea"
df$CNT[df$CNT=="Chinese Taipei"] <- "Taiwan"
df$CNT[df$CNT=="Slovak Republic"] <- "Slovakia"
df$CNT[df$CNT=="Russian Federation"] <- "Russia"
df$CNT[df$CNT=="Perm(Russian Federation)"] <- "Russia"
df$CNT[df$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
df$CNT[df$CNT=="China-Shanghai"] <- "China"
df$CNT[df$CNT=="China-Macau"] <- "China"
df$CNT[df$CNT=="Connecticut (USA)"] <- "United States of America"
df$CNT[df$CNT=="Florida (USA)"] <- "United States of America"
df$CNT[df$CNT=="Massachusetts (USA)"] <- "United States of America"

df
}

```

```{r school}
# Extract relevant variables from schools table
school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC22Q01,
                                       SC22Q02, SC22Q03, SC22Q04,
                                       SC22Q05, SC22Q06, SC22Q07,
                                       SC22Q08, SC22Q09, SC22Q10,
                                       SC22Q11, SC22Q12, SC22Q13, SC04Q01)
# Extract scores from students table
student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, ST08Q01, ST09Q01)
school2012.sub = fix_country(school2012.sub)
student2012.sub = fix_country(student2012.sub)

# Truancy data by student
truancy = student2012.sub %>% group_by(CNT, SCHOOLID) %>% left_join(school2012.sub, by = c("CNT", "SCHOOLID"))
```

```{r truancy, fig.width=6, fig.height=5}
# school perception of truancy vs math scores
# percentage of schools reporting high truancy issues
df = school2012.sub %>% group_by(CNT) %>% summarise(truancy = sum(SC22Q01 == "A lot", na.rm=T) / length(CNT) ) %>% arrange(desc(truancy)) %>% 
  left_join(
    student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math))
  )
# Russia and china are outliers, reporting a lot of truancy, but still having high scores
# A lot of schools with low truancy still have low scores
qplot(x = truancy, y = m_math, data = df, xlab="Proportion of schools reporting a lot of truancy", 
      ylab="Math Score") + stat_smooth(method="lm", se = FALSE)

# student self reported truancy within the last two weeks
# percentage of high truancy students
df2 = student2012.sub %>% group_by(CNT) %>% summarise(truancy_student = sum(ST08Q01 == "Five or more times  " |
            ST08Q01 == "Three or four times  " | 
            ST08Q01 == "One or two times  ", na.rm=T) / length(CNT)) %>% 
          arrange(desc(truancy_student)) %>% 
          left_join(student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math)))

# No schools stand out as outliers here
qplot(truancy_student, m_math, data = df2, 
      xlab="Proportion of students reporting missing at least one day in past two weeks", 
      ylab="Math Score") + stat_smooth(method="lm", se = FALSE)
```

```{r drugs, fig.width=6, fig.height=5}
# percentage of schools reporting drug problems
drugs = school2012.sub %>% group_by(CNT) %>% 
  summarise(drug = sum(SC22Q07 == "A lot" | SC22Q07 == "To some extent", na.rm=T) / length(CNT)) %>% 
  arrange(desc(drug)) %>%
   left_join(student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math)))
# Most countries report few drug problems, Kazakhstan reports a lot, China surprisingly also, Costa Rica, Colombia, Russia, Canada, USA
qplot(x = drug, y = m_math, data = drugs, xlab="Proportion of schools reporting some drug problems", 
      ylab="Math Score") + stat_smooth(method="lm", se = FALSE)

#school2012.sub %>% group_by(CNT) %>% summarise(drug = sum(SC22Q07 == "A lot", na.rm=T) / length(CNT)) %>% 
#  arrange(desc(drug))
```

Competition between schools, number of schools students could choose to attend

```{r competition, fig.width=6, fig.height=5}
comp = school2012.sub %>% 
  select(CNT, SCHOOLID, SC04Q01) %>% 
  group_by(CNT) %>%
  summarise(other = sum(SC04Q01 == "Two or More", na.rm=T) / length(CNT)) %>% 
  arrange(desc(other)) %>%
   left_join(student2012.sub %>% group_by(CNT) %>% summarise(m_math = mean(math)))

qplot(other, m_math, data = comp, xlab="Proportion of students who have school choice", ylab="Math Score") +
  stat_smooth(method="lm", se = FALSE)
```

Math teacher shortages

```{r shortages, fig.width=12, fig.height=10}
# SC11Q02, SC11Q03 Computers
# SC14Q02 - Math teacher shortage
# SC14Q07 - Computer shortage
# SC14Q08 - Internet shortage
# SC14Q09 - Software shortage
# SC14Q10 - Library material shortage
# SC14Q11 - Buildings/Grounds shortage
# SC14Q12 - Heating cooling shortage
# SC14Q13 - Instructional space shortage
#school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC14Q02, SC14Q07, SC14Q10)
student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci)
school2012.sub = fix_country(school2012.sub)
student2012.sub = fix_country(student2012.sub)

scores_by_school = student2012.sub %>% group_by(CNT, SCHOOLID) %>% 
  summarise(mmath = mean(math), mread = mean(read), msci = mean(sci))

df = scores_by_school %>% left_join(school2012.sub)
df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(CNT, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
qplot(SC14Q02, mmath2, data = df.mshort, facets = ~CNT, 
      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=CNT)
df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(CNT, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
qplot(SC14Q10, mmath2, data = df.lshort, facets = ~CNT, 
      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=CNT)
df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], CNT, SC14Q02), teachers = mean(mmath, na.rm=T))
colnames(df2)[2] <- "levels"
df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], CNT, SC14Q07), computers = mean(mmath, na.rm=T))
colnames(df3)[2] <- "levels"
df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], CNT, SC14Q10), libmaterials = mean(mmath, na.rm=T))
colnames(df4)[2] <- "levels"
df.all <- merge(df2, df3)
df.all <- merge(df.all, df4)
df.m <- melt(df.all, id=c("CNT", "levels"))
colnames(df.m)[4] <- "math"
qplot(levels, math, data = df.m, facets = ~CNT, 
      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
      theme(legend.position="bottom") 
#ggsave("shortages.pdf", width=14, height=14)
qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", colour=I("#1B9E77"))
#ggsave("teacher-shortages.pdf", width=5, height=5)
qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage")
#ggsave("computer-shortages.pdf", width=5, height=5)
qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage")
#ggsave("library-shortages.pdf", width=5, height=5)
```

looking at just the US
```{r shortages1}
library(ggplot2)
qplot(SC14Q02, mmath, data = df %>% group_by(CNT, SC14Q02) %>% 
        filter(CNT == "United States of America"), geom="boxplot")
```

```{r shortages2}
library(ggplot2)
df2 = df%>% select(CNT, SCHOOLID, mmath, mread, msci, SC14Q02,
                  SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11,
                  SC14Q12, SC14Q13)
for(i in 6:13) {
  df2[,i] = as.numeric(df2[,i])
}

df2$sum = apply(df2[,6:13], 1, sum, na.rm=T)
qplot(x = sum, y = m2, data = df2 %>% group_by(CNT, sum) %>% summarise(m2 = mean(mmath)), facets=~CNT)

qplot(x = pSC14Q07, mmath2, data = 
df2 %>% group_by(CNT) %>% 
  summarise(pSC14Q07 = sum(SC14Q07 == 4, na.rm=T) / length(CNT),
            mmath2 = mean(mmath)) %>% 
  arrange(pSC14Q07)
)
```

### 2014-06-28 dicook/gender.Rmd
checking weights, in gender.Rmd

@@
 ```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
-student2012.sub <- student2012[, c(1, 12, 501, 541, 546)]
+student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
 colnames(student2012.sub)[1] <- "name"
@@
 student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
 student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
-                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
+                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
+                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
+                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
+                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                   mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
@@
                   propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
+qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
+      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
 student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)

### 2014-06-28 dicook/school_matters.Rmd
checking weights, in gender.Rmd

@@
 #ggsave("shortages.pdf", width=14, height=14)
-qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", colour=I("#1B9E77"))
+qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
 #ggsave("teacher-shortages.pdf", width=5, height=5)
-qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage")
+qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
 #ggsave("computer-shortages.pdf", width=5, height=5)
-qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage")
+qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
 #ggsave("library-shortages.pdf", width=5, height=5)

### 2014-06-28 dicook/school_start_age.Rmd
starting age files

--- 2014-06-27 dicook/gender.Rmd
@@
 ```
-```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
-student2012.sub <- student2012[, c(1, 12, 501, 541, 546)]
+```{r age, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}
+student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
 colnames(student2012.sub)[1] <- "name"
+# Now add data
+student2012.sub$name <- as.character(student2012.sub$name)
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
+student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
+student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
+student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
 student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
@@
 student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
-student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
-                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
-                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
-                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
-                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
-                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
-                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
-                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
-                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
-                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
-                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
-                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
-                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
-                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
-                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
-                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
-student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
-student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
-qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
-# Test colors
-#qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
-#  xlab("") +
-#  scale_colour_manual("Significant", values=c("male"=dichromat("skyblue"), "female"=dichromat("pink"), "none"=dichromat("lightgreen"))) +
-#  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
-#  scale_size("Prop male") +
-#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
-#ggsave("gendermathgap.pdf", width=7, height=14)
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
-qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-  xlab("") +
-  geom_point(aes(y=maxfemale), colour="pink") + 
-  scale_y_continuous("High Math") + coord_flip() + theme_bw()
-#ggsave("gendermathtop.pdf", width=3.5, height=7)
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
-qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-  xlab("") +
-  geom_point(aes(y=minfemale), colour="pink") + 
-  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
-#ggsave("gendermathbottom.pdf", width=3.5, height=7)
-#
-student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
-student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
-qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Reading Score Gap") + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
-#ggsave("genderreadgap.pdf", width=5, height=8)
-student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
-student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
-qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Science Score Gap") + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
-#ggsave("gendergap.pdf", width=10, height=14)
-
+student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
+student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
+                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
+                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
+                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
+                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
+student2012.sub.age.m <- melt(student2012.sub.age)
+student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Math Score", 
+      geom=c("point","smooth"), se=F, group=name) + 
+ facet_wrap(~name, ncol=8) + theme_bw() 
+# ggsave("school-start.pdf", width=14, height=14)
+student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
+                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
+                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
+                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
+                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
+student2012.sub.age.m <- melt(student2012.sub.age)
+student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
+      geom=c("point","smooth"), se=F, group=name) + 
+ facet_wrap(~name, ncol=8) + theme_bw() 
+student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
+                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
+                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
+                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
+                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
+student2012.sub.age.m <- melt(student2012.sub.age)
+student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
+      geom=c("point","smooth"), se=F, group=name) + 
+ facet_wrap(~name, ncol=8) + theme_bw() 
 ````
@@
 detach("package:plyr")
-# Now add data
-student2012.sub.summary.gap$name <- as.character(student2012.sub.summary.gap$name)
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Serbia"] <- "Republic of Serbia"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Korea"] <- "South Korea"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Chinese Taipei"] <- "Taiwan"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Slovak Republic"] <- "Slovakia"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Russian Federation"] <- "Russia"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Perm(Russian Federation)"] <- "Russia"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="China-Shanghai"] <- "China"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="China-Macau"] <- "China"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Connecticut (USA)"] <- "United States of America"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Florida (USA)"] <- "United States of America"
-student2012.sub.summary.gap$name[student2012.sub.summary.gap$name=="Massachusetts (USA)"] <- "United States of America"
 
-student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
+student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
+student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
+student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
 ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
-#ggsave("gendermathmap.pdf", width=12, height=7)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
-#ggsave("genderreadmap.pdf", width=12, height=7)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
-#ggsave("gendermap.pdf", width=12, height=8)
-#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
-#  scale_fill_manual("Significant", values=c("male"=dichromat("skyblue"), "female"=dichromat("pink"), "none"=dichromat("lightgreen"))) +
-#  new_theme_empty + theme(legend.position="none")
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
+  scale_fill_discrete("Median Start Age") +
+  new_theme_empty + theme(legend.position="bottom")
+#ggsave("agestartmap.pdf", width=12, height=7)
 ```

### 2014-06-28 Luke/Graphs/Pisa viscomp final.R
Submission Code And pngs
I have a submission for both tracks

@@ -1,7 +1,7 @@
 setwd("/Users/lukefostvedt/Documents/PISA 2012/RDA files/")
 library(stringr)
 library(ggplot2)
-library(reshape)
+library(reshape2)
 library(dplyr)
 library(lubridate)
 library(ggvis)
@@ -9,6 +9,11 @@ library(doBy)
 library(maps)
 library(ggmap)
 library(rworldmap)
+library(grid)
+library(scales)
+
+
+
 sessionInfo()
 ################## IMPORTANT #######################
 # ************************************************
@@ -53,137 +58,8 @@ rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012
 load("student2012pvmeans.Rdata")
 student2012 <- cbind(student2012,a)
 
-student2012$ST15Q01 <- addNA(student2012$ST15Q01)
-student2012$ST19Q01 <- addNA(student2012$ST19Q01)
-
-student2012$name <- as.character(student2012$CNT)
-# unique(anti_join(student2012.sub, world.polys)[1])
-student2012$name[student2012$name=="United Arab Emirates"] <- "UAE"
-student2012$name[student2012$name=="United Kingdom"] <- "UK"
-student2012$name[student2012$name=="Serbia"] <- "Serbia"
-student2012$name[student2012$name=="Korea"] <- "South Korea"
-student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong"
-student2012$name[student2012$name=="China-Shanghai"] <- "China"
-student2012$name[student2012$name=="Macau"] <- "China"
-student2012$name[student2012$name=="Connecticut (USA)"] <- "USA"
-student2012$name[student2012$name=="Florida (USA)"] <- "USA"
-student2012$name[student2012$name=="Massachusetts (USA)"] <- "USA"
[…]

### 2014-06-28 Luke/parent submission code.R

@@ -1,0 +1,184 @@
+setwd("/Users/lukefostvedt/Documents/PISA 2012/RDA files/")
+library(stringr)
+library(ggplot2)
+library(reshape)
+library(dplyr)
+library(lubridate)
+library(ggvis)
+library(doBy)
+library(maps)
+library(ggmap)
+library(rworldmap)
+sessionInfo()
+################## IMPORTANT #######################
+# ************************************************
+# My file path has the rda files in the main folder.
+# They are not in a "data" folder.
+# ************************************************
+####################################################
+
+sets <- c("item", "parent", "school", "scoredItem", "student")
+
+# function to build the file names
+fn_build <- function(file_name) {
+
+    template <- c("2012.rda", "2012dict.rda")
+
+    file_name %>% vapply(str_join, template, template) %>% file.path(".", "", 
+        .)
+}
+
+# load the data
+sets %>% fn_build %>% lapply(load, .GlobalEnv)
+
+
+fn_make_df <- function(named_vector){
+  data.frame(
+    variable = attr(named_vector, "names"),
+    description = named_vector,
+    row.names = NULL
+  )
+}
+
+# there's a clever way to do this, but beyond me for naw
+dict_item2012 <- fn_make_df(item2012dict) 
+dict_parent2012 <- fn_make_df(parent2012dict) 
+dict_school2012 <- fn_make_df(school2012dict) 
+dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
+dict_student2012 <- fn_make_df(student2012dict) 
+
+# clean
+rm(fn_make_df)
+rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
+load("student2012pvmeans.Rdata")
+student2012 <- cbind(student2012,a)
+
+student2012$ST15Q01 <- addNA(student2012$ST15Q01)
+student2012$ST19Q01 <- addNA(student2012$ST19Q01)
+
+student2012$name <- as.character(student2012$CNT)
+# unique(anti_join(student2012.sub, world.polys)[1])
+student2012$name[student2012$name=="United Arab Emirates"] <- "UAE"
[…]

### 2014-06-28 Luke/Pisa viscomp final.R
Updated School Submission

@@ -139,7 +139,7 @@ student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01"
 study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
 names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
 levels(study$Subject) <- c("Math","Reading","Science")
-p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
+p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
 #ggsave(p,file="studytime.pdf",height=13,width=13)
 p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
 ggsave(p,file="HistHWschool.pdf",width=8,height=6)

### 2014-06-28 alex/albania.Rmd

--- 2014-06-27 alex/albania.Rmd
@@
-Equality for all: Only in student2012.sub
+Albania is awesome
 =============
@@
 student2012$sci = as.numeric(student2012[, "PV1SCIE"])
+
+#separate albania
+student2012$OECD = as.character(student2012$OECD)
+student2012[student2012$CNT == "Albania", ]$OECD = "Albania"
 ```
 
-```{r vars}
+```{r vars, echo=FALSE}
 student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
@@
 student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
-
-#separate albania
-student2012.sub$OECD = as.character(student2012.sub$OECD)
-student2012.sub[student2012.sub$CNT == "Albania", ]$OECD = "Albania"
 ```
 
-```{r possessions, echo=FALSE}
+```{r possessions, echo=FALSE, fig.width=15, warning=FALSE, message=FALSE}
 library(reshape2)
+
 df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+
 df$group = paste0(df$CNT, df$variable)
-qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group)
+qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions")
 
-qplot(x = numposs, y = mmath, data = student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), facets =~OECD, geom = "line", group = CNT) + ylab("Math Scores")
+qplot(x = numposs, y = mmath, data = student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), facets =~OECD, geom = "line", group = CNT) + ylab("math Scores")
 ```
 
-```{r gender, echo=FALSE}
+```{r OUTHOURS, echo=FALSE, fig.width=15, warning=FALSE, message=FALSE}
+ggplot(subset(student2012.sub, !is.na(OUTHOURS)), aes(x = OUTHOURS, fill = CNT)) + geom_density(alpha = I(0.4)) + facet_wrap(~OECD) + theme(legend.position = "none")
+qplot(x = OUTHOURS, data = student2012.sub, geom = "density", facets = ~CNT, fill = I("grey"))
+qplot(x = OUTHOURS, data = student2012.sub %>% filter(CNT == "Albania"), geom="density", fill = I("grey"))
+qplot(x = OUTHOURS, data = student2012.sub %>% filter(CNT == "Albania"))
+```
 
+```{r parentjobs, echo=FALSE, fig.width=15}
+father.df = melt(student2012 %>% select(CNT, ST15Q01, ST19Q01, math, sci, read, OECD) %>% group_by(ST15Q01, CNT, OECD) %>% summarise(math = mean(math), sci = mean(sci), read = mean(read)), id = c("CNT", "OECD", "ST15Q01"))
+father.df$group = paste0(father.df$CNT, father.df$variable)
+levels(father.df$ST15Q01) = c("full time", "part time", "unemployed", "retired")
+
+qplot(x = ST15Q01, y = value, color = variable, facets = ~OECD, data = father.df, group = group, geom = "line") + xlab("father's employment status") + ylab("test scores")
 ```
+
+```{r truancyschool, echo=FALSE, fig.width=15}
+school2012.sub = school2012 %>% select(CNT, SCHOOLID, SC22Q01)
+student2012.sub = student2012 %>% select(CNT, SCHOOLID, math, read, sci, OECD)
+truancy.df = melt( student2012.sub %>% 
+  group_by(CNT, SCHOOLID) %>% 
+  left_join(school2012.sub, by = c("CNT", "SCHOOLID")) %>% 
+  group_by(CNT, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
+  id = c("CNT", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
+truancy.df$group = paste0(truancy.df$CNT, truancy.df$variable)
+
+qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores")
+```
+
+```{r gendergap, echo=FALSE, warning = FALSE}
+scores = student2012 %>%
+  select(CNT, math, read, ST04Q01, OECD) %>% 
+  group_by(OECD, CNT, ST04Q01) %>%
+  summarise(math = mean(math), read = mean(read))
+male = scores %>% filter(ST04Q01 == "Male")
+female = scores %>% filter(ST04Q01 == "Female")
+scores.df = data.frame(OECD = male$OECD, CNT = male$CNT, math_difference = male$math - female$math, read_difference = male$read - female$read)
+
+scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$math_difference)])
+qplot(CNT, math_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
+  scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
+  geom_hline(yintercept=0, colour="grey80") + theme_bw() + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+
+scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$read_difference)])
+qplot(CNT, read_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
+  scale_y_continuous("read score gap: male - female scores") +
+  geom_hline(yintercept=0, colour="grey80") + theme_bw() + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+```

### 2014-06-28 alex/albania.Rmd
getting ready to put onto template

@@
 df$group = paste0(df$CNT, df$variable)
-qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions")
+qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
 
@@
 ```{r OUTHOURS, echo=FALSE, fig.width=15, warning=FALSE, message=FALSE}
-ggplot(subset(student2012.sub, !is.na(OUTHOURS)), aes(x = OUTHOURS, fill = CNT)) + geom_density(alpha = I(0.4)) + facet_wrap(~OECD) + theme(legend.position = "none")
+ggplot(subset(student2012.sub, !is.na(OUTHOURS)), aes(x = OUTHOURS, fill = CNT)) + geom_density(alpha = I(0.4)) + facet_wrap(~OECD) + theme(legend.position = "none") + scale_x_continuous("out of school study time (hours)", limits = c(0, 65))
+
 qplot(x = OUTHOURS, data = student2012.sub, geom = "density", facets = ~CNT, fill = I("grey")) 
@@
 
-qplot(x = ST15Q01, y = value, color = variable, facets = ~OECD, data = father.df, group = group, geom = "line") + xlab("father's employment status") + ylab("test scores")
+qplot(x = ST15Q01, y = value, color = variable, facets = ~OECD, data = father.df, group = group, geom = "line") + xlab("father's employment status") + ylab("test scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
 ```
@@
 
-qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores")
+qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
 ```
 
-```{r gendergap, echo=FALSE, warning = FALSE}
+```{r gendergap, echo=FALSE, warning = FALSE, fig.height = 10}
 scores = student2012 %>%
@@
   scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
-  geom_hline(yintercept=0, colour="grey80") + theme_bw() + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
 
@@
   scale_y_continuous("read score gap: male - female scores") +
-  geom_hline(yintercept=0, colour="grey80") + theme_bw() + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+
+#qplot(CNT, value, data = melt(scores.df, id = c("OECD", "CNT")), color = OECD, size = OECD, facets = ~variable) + 
+#  coord_flip() + xlab("") + ylab("Score difference: male - female") + scale_size_manual(values = c(4, 2.5, 2.5))
 ```

### 2014-06-28 alex/albania.Rmd
new poster, idea for alex

--- 2014-06-28 alex/albania.Rmd
@@
 ```
+
+```{r another idea}
+.width = 1
+
+data_small <- 
+  student2012 %>%
+  select(OECD, CNT, OUTHOURS) %>%
+  filter(!is.na(OUTHOURS))
+
+ggplot(mapping = aes(x = OUTHOURS)) + 
+  geom_histogram(
+    mapping = aes(y = ..density.., fill = OECD), 
+    binwidth = .width,
+    data = data_small %>% filter(OECD == "Albania")
+  ) + 
+  geom_line(
+    mapping = aes(group = CNT, color = OECD), 
+    stat = "density", 
+    alpha = 0.3,
+    data = data_small %>% filter(OECD != "Albania")
+  ) 
+```

### 2014-06-28 alex/albania.Rmd
plot for Alex

--- 2014-06-28 alex/albania.Rmd
@@
 ```
+
+```{r another idea}
+.width = 1
+
+data_small <- 
+  student2012 %>%
+  select(OECD, CNT, OUTHOURS) %>%
+  filter(!is.na(OUTHOURS))
+
+ggplot(mapping = aes(x = OUTHOURS)) + 
+  scale_x_continuous(name = "what are OUTHOURS?") +
+  scale_y_continuous(name = "is there a better description than density?") +
+  geom_histogram(
+    mapping = aes(y = ..density.., fill = OECD), 
+    binwidth = .width,
+    data = data_small %>% filter(OECD == "Albania")
+  ) + 
+  scale_fill_manual(
+    values = "black",
+    breaks = "Albania",
+    guide = guide_legend(title = "")
+  ) + 
+  geom_line(
+    mapping = aes(group = CNT, color = OECD), 
+    stat = "density", 
+    alpha = 0.3,
+    data = data_small %>% filter(OECD != "Albania")
+  ) +
+  scale_color_discrete(
+    guide = guide_legend(
+      title = "Other Countries",
+      override.aes = list(alpha = 1)
+    )
+  )
+```

### 2014-06-28 alex/albania.Rmd

--- 2014-06-28 alex/albania.Rmd
@@
-Albania is awesome
+Albania
 =============
@@
 scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$math_difference)])
+scores.df$OECD = factor(scores.df$OECD, levels = scores.df$OECD[])
+
 qplot(CNT, math_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
   scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
+  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
 
@@
   scale_y_continuous("read score gap: male - female scores") +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("")
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
+  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
 
@@
 
-```{r another idea}
+```{r another idea, fig.width=8, warning = FALSE, message=FALSE}
 .width = 1
@@
 ggplot(mapping = aes(x = OUTHOURS)) + 
+  scale_x_continuous(name = "out of school study time (hours)", limits = c(0,60)) +
+  scale_y_continuous(name = "density") +
   geom_histogram(
@@
     data = data_small %>% filter(OECD == "Albania")
+  ) + 
+  scale_fill_manual(
+    values = "black",
+    breaks = "Albania",
+    guide = guide_legend(title = "")
   ) + 
@@
     data = data_small %>% filter(OECD != "Albania")
+  ) +
+  scale_color_discrete(
+    guide = guide_legend(
+      title = "Other Countries",
+      override.aes = list(alpha = 1)
     )
+
+  ) +
+  scale_colour_manual(values = c("#f4a582", "#92c5de")) + guides(colour = guide_legend(override.aes = list(alpha = 1)))
 ```

### 2014-06-28 dicook/school_matters.Rmd

@@
 
-```{r shortages2}
+```{r shortages2, eval=FALSE}
 library(ggplot2)

### 2014-08-13 paper/paper.Rmd
paper directory

PISA 2012 It is weirder than you think
========================================================

Hello

### 2014-08-13 paper/paper.Rmd
title
first few paragraphs

@@
-PISA 2012 It is weirder than you think
+Seven Things That Affect Education Around the World
 ========================================================
 
-Hello
+Reducing inequities in education outcomes is of great interest to politicians.
+Dissecting the causes of inequality is a much tougher challenge with no panacea.
+Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints.
+The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating additional emphasis on one of mathematics, reading, and science.
+In 2012, the additional emphasis was on mathematics.
+All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey.
+This represents over 80% of the global economy.
+The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add a collective US$200 trillion to the world's collective GDP.
+
+The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p.
+3]{PISA2014a}).
+The goal of the PISA survey is to assess the workforce readiness of 15-year old students.
+Nearly 500,000 students were tested across the 65 countries and economies.
+Students are examined on how well they can apply the knowledge they learned in school to applications outside of school.
+Additional information about the students, parents, and schools is also collected.
+The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school.
+Schools principals responded to a questionnaire covering the school system and learning experiences for students.
+In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.
+
+This paper will evaluate the role of several factors in which we see inequalities in mathematics, reading, and science outcomes.
+In section XXXgender, we will evaluate in what subjects we observe gender inequalities.
+In section XXX we will evaluate inequalities based on the household structure and occupation of the parents.
+In section XXXpossessions we will look at what home possessions are good predictors of inequalities.
+Lastly, in sections XXXschool, we will look at the role of school influences.

### 2014-08-13 paper/paper.Rmd
topics
topics done

@@
 Lastly, in sections XXXschool, we will look at the role of school influences.
+
+### Teachers, computers, and libraries at school matter
+
+### Gender gap in math is not universal but the reading gap is
+
+### School start age affects performance
+
+### Time spent out of school studying is important, but only up to a point
+
+### More TVs yield higher math scores in the developing world, but lower in the developed countries
+
+### Parents matter
+
+### The weird and wacky republic of Albania

### 2014-08-15 ijlyttle/Untitled.R
cleanup from six weeks ago

@@ -1,0 +1,38 @@
+my_data <- student2012 %>% filter(CNT == "Albania")
+my_data$age <- factor(my_data$ST06Q01)
+
+ggplot(
+  data = my_data,
+  aes_string(
+    x = "age",
+    y = "PV1MATH"
+  )
+) + 
+  geom_violin(scale = "width")
+
+str_sym <- c("CNT", "ST04Q01", "ST28Q01")
+#str_sym <- c("CNT")
+
+symbols = str_sym %>% lapply(as.symbol)
+
+df <- 
+  student2012[, str_sym] %>%
+  regroup(symbols) %>%
+  summarize(n = n()) %>%
+
+df_cnt <-
+  student2012[, "CNT"] %>%
+  group_by(CNT) %>%
+  summarize(n_country = n())
+
+df_new <- 
+  left_join(df, df_cnt, by = "CNT") %>%
+  mutate(prop = n / n_country)
+  
+  
+call <- substitute(
+  select(student2012, sym),  
+  list(sym = lapply(str_sym, as.symbol))
+)
+
+tmp <- student2012[, str_sym]

### 2014-08-15 paper/paper.Rmd
adds data-loading scaffolding

@@
 ========================================================
+
+Luke Fostveldt, Alex Shum, Ian Lyttle, Di Cook
 
@@
 
+```{r read_me, echo = FALSE}
+# Some notes for we authours:
+# 
+# * If you want to run this code interactively, change your working directory to
+# `paper`.
+# 
+# * If there are packges we are all using, let's load them all once, in the
+# load_packages chunk.
+# 
+# * If there are packages particular to your section, please consider loading
+# them there, then detaching at the end of your section.
+# 
+# * I am being a bit presumptuous by leaving stringr loaded - maybe it makes
+# sense to detach it. I'm sure we'll figure out wich way to go with each of our
+# loaded packages.
+# 
+# - IJL
+```
+
+```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE}
+library(dplyr)
+library(stringr)
+```
+
+```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
+# This is the scaffolding to import the data
+sets <- c("item", "parent", "school", "scoredItem", "student")
+
+# function to build the file names
+fn_build <- function(file_name){
+
+  template <- c("2012.rda", "2012dict.rda")
+
+  file_name %>% 
+    vapply(str_join, template, template) %>% 
+    file.path("..", "data", .)
+}
+
+# load the data
+sets %>% fn_build %>% lapply(load, .GlobalEnv)
+
+# clean
+rm(fn_build, sets)
+
+# function to convert to data-frames
+fn_make_df <- function(named_vector){
+  data.frame(
+    variable = attr(named_vector, "names"),
+    description = named_vector,
+    row.names = NULL
+  )
+}
+
+# there's a clever way to do this, but beyond me for naw
+dict_item2012 <- fn_make_df(item2012dict) 
+dict_parent2012 <- fn_make_df(parent2012dict) 
+dict_school2012 <- fn_make_df(school2012dict) 
+dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
+dict_student2012 <- fn_make_df(student2012dict)
+
+# clean
+rm(fn_make_df)
+#rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
+```
+
 ### Teachers, computers, and libraries at school matter

### 2014-08-15 paper/paper.Rmd
hides output

@@
 
-```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE}
+```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
 library(dplyr)
@@
 
-```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
+```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
 # This is the scaffolding to import the data

### 2014-08-26 paper/paper.Rmd
code working for school age, needs some text description now

@@
 library(stringr)
+library(stringr)
+library(reshape2)
+library(ggplot2)
+library(lubridate)
+library(ggvis)
+library(maps)
+library(ggmap)
+library(rworldmap)
+library (grid)    
+library (scales) 
 ```
@@
 #rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
+dim(student2012)
+length(table(student2012$STIDSTD))
 ```
 
+```{r mapdata, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+# Produce the maps
+extractPolygons <- function(shapes) {
+
+  dframe <- ldply(1:length(shapes@polygons), function(i) {
+    ob <- shapes@polygons[[i]]@Polygons
+    dframe <- ldply(1:length(ob), function(j) {
+      x <- ob[[j]]
+      co <- x@coords
+      data.frame(co, order=1:nrow(co), group=j)
+    })
+    dframe$region <- i
+    dframe$name <- shapes@polygons[[i]]@ID
+    dframe
+  })
+  # construct a group variable from both group and polygon:
+  dframe$group <- interaction(dframe$region, dframe$group)
+
+  dframe
+}
+
+# To get a blank background on map
+new_theme_empty <- theme_bw()
+new_theme_empty$line <- element_blank()
+new_theme_empty$rect <- element_blank()
+new_theme_empty$strip.text <- element_blank()
+new_theme_empty$axis.text <- element_blank()
+new_theme_empty$plot.title <- element_blank()
+new_theme_empty$axis.title <- element_blank()
+new_theme_empty$plot.margin <- structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit")
+
+world <- getMap(resolution = "low")
+library(plyr)
+world.polys <- extractPolygons(world)
+detach("package:plyr")
+```
 ### Teachers, computers, and libraries at school matter
@@
 
+We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
+
+```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
+colnames(student2012.sub)[1] <- "name"
+# Now add data
+student2012.sub$name <- as.character(student2012.sub$name)
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
+student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
+student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
+student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
+student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
+student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
+                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
+                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
+                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
+                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
+student2012.sub.age.m <- melt(student2012.sub.age)
+student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
+orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
+student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
+student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
+student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
+qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Score", 
+      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
+ facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
+# ggsave("school-start.pdf", width=14, height=14)
+#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
+#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
+#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
+#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
+#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
+#student2012.sub.age.m <- melt(student2012.sub.age)
+#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
+#      geom=c("point","smooth"), se=F, group=name) + 
+# facet_wrap(~name, ncol=8) + theme_bw() 
+#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
+#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
+#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
+#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
+#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
+#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
+#student2012.sub.age.m <- melt(student2012.sub.age)
+#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
+#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
+#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
+#      geom=c("point","smooth"), se=F, group=name) + 
+# facet_wrap(~name, ncol=8) + theme_bw() 
+```
+
+```{r maps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
+# Extract map polygons for modern world
+world <- getMap(resolution = "low")
+library(plyr)
+world.polys <- extractPolygons(world)
+detach("package:plyr")
+
+student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
+student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
+  scale_fill_discrete("Median Start Age") +
+  new_theme_empty + theme(legend.position="bottom")
+#ggsave("agestartmap.pdf", width=12, height=7)
+```
 ### Time spent out of school studying is important, but only up to a point

### 2014-08-26 paper/paper.Rmd
Added some text to go along with the plots

@@
 
-We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
+***We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
+
+Broken out by country we have computed the average math test score by age that the children started school.
+In additon, for each country we computed the median age that children in the study started school.
+There are differences in typical school start age between countries.
+Ireland has the youngest median start age - children head to school at age 4.
+Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5.
+For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7.
+Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.
+
+Within a country, looking at the average math score for each start age, the score typically declines, especially after age 6.
+This is almost uniformly the pattern across all of the countries in the study.
+For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than another other starting age group.
+School start age matters, that children need to start school in the 4-6 age range to expect better math scores, on average.
 

### 2014-08-28 paper/paper.Rmd
revised intro

@@
 
-This paper will evaluate the role of several factors in which we see inequalities in mathematics, reading, and science outcomes.
-In section XXXgender, we will evaluate in what subjects we observe gender inequalities.
-In section XXX we will evaluate inequalities based on the household structure and occupation of the parents.
-In section XXXpossessions we will look at what home possessions are good predictors of inequalities.
-Lastly, in sections XXXschool, we will look at the role of school influences.
+The data was made available by the OECD as part of a data challenge for the useR!
+2014 conference.
+Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/.
+We learned a lot from this data, and won one of the prizes.
+This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.
 

### 2014-08-28 paper/paper.Rmd
setting up data

@@
 ```
+
+```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+colnames(student2012)[1] <- "name"
+# Now add data
+student2012$name <- as.character(student2012$name)
+student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
+student2012$name[student2012$name=="Korea"] <- "South Korea"
+student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
+student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
+student2012$name[student2012$name=="Russian Federation"] <- "Russia"
+student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
+student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012$name[student2012$name=="China-Shanghai"] <- "China"
+student2012$name[student2012$name=="China-Macau"] <- "China"
+student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
+student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
+student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
+student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
+student2012$PV1MATH <- as.numeric(student2012$PV1MATH)
+student2012$PV1READ <- as.numeric(student2012$PV1READ)
+student2012$PV1SCIE <- as.numeric(student2012$PV1SCIE)
+student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
+```
+
 ### Teachers, computers, and libraries at school matter
@@
 student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
-colnames(student2012.sub)[1] <- "name"
-# Now add data
-student2012.sub$name <- as.character(student2012.sub$name)
-student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
-student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
-student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
-student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
-student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
-student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
-student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
-student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
-student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
-student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
-student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
-student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
-student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
-student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
-student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
-student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
 student2012.sub.age <- summarise(group_by(student2012.sub, name), 

### 2014-08-28 paper/paper.Rmd
added gender exploration

@@
 
+*** We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
+
 ```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
@@
 
+```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
+student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
+colnames(student2012.sub)[1] <- "name"
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
+student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
+student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
+                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
+                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
+                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
+                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
+                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
+                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
+                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
+                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
+                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
+                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
+                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
+                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
+                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
+                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
+                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
+                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
+                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
+                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
+#qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
+#      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
+student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
+student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
+student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
+qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
+  xlab("") +
+  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
+  scale_size("Prop male") +
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
+qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
+  xlab("") +
+  geom_point(aes(y=maxfemale), colour="pink") + 
+  scale_y_continuous("High Math") + coord_flip() + theme_bw()
+#ggsave("gendermathtop.pdf", width=3.5, height=7)
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
+qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
+  xlab("") +
+  geom_point(aes(y=minfemale), colour="pink") + 
+  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
+#ggsave("gendermathbottom.pdf", width=3.5, height=7)
+#
+student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
+student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
+student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
+qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
+  xlab("") +
+  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  scale_y_continuous("Reading Score Gap") + 
+  scale_size("Prop male") +
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
+#ggsave("genderreadgap.pdf", width=5, height=8)
+student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
+student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
+student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
+student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
+qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
+  xlab("") +
+  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  scale_y_continuous("Science Score Gap") + 
+  scale_size("Prop male") +
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
+```
+
+```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}
+student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
+  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  new_theme_empty + theme(legend.position="none")
+#ggsave("gendermathmap.pdf", width=12, height=7)
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
+  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  new_theme_empty + theme(legend.position="none")
+#ggsave("genderreadmap.pdf", width=12, height=7)
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
+  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  new_theme_empty + theme(legend.position="none")
+```
+
 ### School start age affects performance
-
-***We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
 
@@
 
-```{r maps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
+```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
 # Extract map polygons for modern world

### 2014-08-28 paper/paper.Rmd
filled in the gender section

@@
 Students are examined on how well they can apply the knowledge they learned in school to applications outside of school.
+The scores reported range between 0-1000.
 Additional information about the students, parents, and schools is also collected.
@@
 
-### Gender gap in math is not universal but the reading gap is
+### The gender gap in math is not universal but the reading gap is, in favor of girls
+
+For each country, average scores on math and reading are calculated separately for boys and girls.
+These averages are differenced to compute the "gap" between boys and girls.
+T-statistics were calculated to determine the statistical significance of the difference.
+The results are plotted in the graphs below.
+Each dot corresponds to a difference between girls and boys scores, for each country.
+Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference).
+The size of the dot indicates the proportion of boys in the sample size.
+For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys.
+The maps color the countries by the presence of a gender gap.
+
+The countries are sorted from biggest gap to smallest gap.
+Most countries have a gender gap in math scores in favor of boys.
+Colombia has the biggest gender gap, about 30 points on average boys do better than girls.
+It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.
+Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.
+
+Conversely, though, when we look at reading scores the story is entirely different.
+Universally, girls score significantly better than boys.
+Again, the size of the gap is small, between 10 and 80 points.
+
+Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls.
+There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score.
+There is far more individual variability than the overall average differences.
 
@@
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
-qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-  xlab("") +
-  geom_point(aes(y=maxfemale), colour="pink") + 
-  scale_y_continuous("High Math") + coord_flip() + theme_bw()
+#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
+#qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
+#  xlab("") +
+#  geom_point(aes(y=maxfemale), colour="pink") + 
+#  scale_y_continuous("High Math") + coord_flip() + theme_bw()
 #ggsave("gendermathtop.pdf", width=3.5, height=7)
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
-qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-  xlab("") +
-  geom_point(aes(y=minfemale), colour="pink") + 
-  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
+#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
+#qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
+#  xlab("") +
+#  geom_point(aes(y=minfemale), colour="pink") + 
+#  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
 #ggsave("gendermathbottom.pdf", width=3.5, height=7)
@@
   scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
 #ggsave("genderreadgap.pdf", width=5, height=8)
-student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
-student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
-qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Science Score Gap") + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
+#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
+#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
+#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
+#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
+#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
+#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
+#  xlab("") +
+#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+#  scale_y_continuous("Science Score Gap") + 
+#  scale_size("Prop male") +
+#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
 ```
 
-```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE}
+```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
 student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
@@
 #ggsave("genderreadmap.pdf", width=12, height=7)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
+#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
+#  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+#  new_theme_empty + theme(legend.position="none")
 ```
 
-### School start age affects performance
+### Starting school at an older age generated lower average math score.
 
-Broken out by country we have computed the average math test score by age that the children started school.
+For each country we have computed the average math test score by age that the children started school.
 In additon, for each country we computed the median age that children in the study started school.
@@
 ### The weird and wacky republic of Albania
+
+### Play with the data yourself
+
+Shiny app
+
+### Acknowledgements
+
+Software packages

### 2014-08-29 paper/paper.Rmd
first copy/paste of text. more to come

--- 2014-08-28 paper/paper.Rmd
@@
 
-The data was made available by the OECD as part of a data challenge for the useR!
-2014 conference.
-Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/.
-We learned a lot from this data, and won one of the prizes.
-This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.
+This paper will evaluate the role of several factors in which we see inequalities in mathematics, reading, and science outcomes.
+In section XXXgender, we will evaluate in what subjects we observe gender inequalities.
+In section XXX we will evaluate inequalities based on the household structure and occupation of the parents.
+In section XXXpossessions we will look at what home possessions are good predictors of inequalities.
+Lastly, in sections XXXschool, we will look at the role of school influences.
 
@@
 ```
-
-```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-colnames(student2012)[1] <- "name"
-# Now add data
-student2012$name <- as.character(student2012$name)
-student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
-student2012$name[student2012$name=="Korea"] <- "South Korea"
-student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-student2012$name[student2012$name=="China-Shanghai"] <- "China"
-student2012$name[student2012$name=="China-Macau"] <- "China"
-student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
-student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
-student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
-student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
-student2012$PV1MATH <- as.numeric(student2012$PV1MATH)
-student2012$PV1READ <- as.numeric(student2012$PV1READ)
-student2012$PV1SCIE <- as.numeric(student2012$PV1SCIE)
-student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
-```
-
 ### Teachers, computers, and libraries at school matter
@@
 student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
+colnames(student2012.sub)[1] <- "name"
+# Now add data
+student2012.sub$name <- as.character(student2012.sub$name)
+student2012.sub$name[student2012.sub$name=="Serbia"] <- "Republic of Serbia"
+student2012.sub$name[student2012.sub$name=="Korea"] <- "South Korea"
+student2012.sub$name[student2012.sub$name=="Chinese Taipei"] <- "Taiwan"
+student2012.sub$name[student2012.sub$name=="Slovak Republic"] <- "Slovakia"
+student2012.sub$name[student2012.sub$name=="Russian Federation"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Perm(Russian Federation)"] <- "Russia"
+student2012.sub$name[student2012.sub$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012.sub$name[student2012.sub$name=="China-Shanghai"] <- "China"
+student2012.sub$name[student2012.sub$name=="China-Macau"] <- "China"
+student2012.sub$name[student2012.sub$name=="Connecticut (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Florida (USA)"] <- "United States of America"
+student2012.sub$name[student2012.sub$name=="Massachusetts (USA)"] <- "United States of America"
+student2012.sub$ST06Q01 <- as.numeric(student2012.sub$ST06Q01)
+student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
+student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
+student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
+student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
 student2012.sub.age <- summarise(group_by(student2012.sub, name), 
@@
 
+It seems that television is the great equalizer. As part of our investigation, we set out to explore how household factors, as indicators of inequality, affect student test scores.
+
+Math scores correlate with the number of household TVs, but how depends on where you live. Nevertheless, access to a quiet study space is generally a plus.
+
+For each country, students’ math scores are compared with number of TVs at home.
+Countries are grouped by direction and strength of the correlation; groups of countries are mapped (above) and identified (right).
+
+Students are considered by group, and categorized by household factors: number of TVs, and access to a quiet study space.
+At far left, number of students in each category; at near left, math scores are summarized for students in each category.
+
+Find your own correlations: http://bit.ly/pisa2012_explorer
+
 ### Parents matter

### 2014-08-29 paper/paper.Rmd
started working on schools matter, but need the full code matching the poster

--- 2014-08-28 paper/paper.Rmd
@@
 library (scales)
+library(doBy)
 ```
@@
 
-```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
 colnames(student2012)[1] <- "name"
+student2012$name <- as.character(student2012$name)
+colnames(school2012)[1] <- "name"
+school2012$name <- as.character(school2012$name)
+
+fix_country = function(df) {
+df$name = as.character(df$name)
+df$name[df$name=="Serbia"] <- "Republic of Serbia"
+df$name[df$name=="Korea"] <- "South Korea"
+df$name[df$name=="Chinese Taipei"] <- "Taiwan"
+df$name[df$name=="Slovak Republic"] <- "Slovakia"
+df$name[df$name=="Russian Federation"] <- "Russia"
+df$name[df$name=="Perm(Russian Federation)"] <- "Russia"
+df$name[df$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+df$name[df$name=="China-Shanghai"] <- "China"
+df$name[df$name=="China-Macau"] <- "China"
+df$name[df$name=="Connecticut (USA)"] <- "United States of America"
+df$name[df$name=="Florida (USA)"] <- "United States of America"
+df$name[df$name=="Massachusetts (USA)"] <- "United States of America"
+
+df
+}
+
 # Now add data
-student2012$name <- as.character(student2012$name)
-student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
-student2012$name[student2012$name=="Korea"] <- "South Korea"
-student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-student2012$name[student2012$name=="China-Shanghai"] <- "China"
-student2012$name[student2012$name=="China-Macau"] <- "China"
-student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
-student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
-student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
+#student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
+#student2012$name[student2012$name=="Korea"] <- "South Korea"
+#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
+#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
+#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
+#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
+#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+#student2012$name[student2012$name=="China-Shanghai"] <- "China"
+#student2012$name[student2012$name=="China-Macau"] <- "China"
+#student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
+#student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
+#student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
 student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
@@
 student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
+
+school2012 = fix_country(school2012)
+student2012 = fix_country(student2012)
 ```
 
-### Teachers, computers, and libraries at school matter
+### Teachers, computers, and libraries at school make a difference in math scores
+
+```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
+school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
+student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)
+
+scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
+                                      mread = mean(PV1READ), msci = mean(PV1SCIE))
+
+df = scores_by_school %>% left_join(school2012.sub)
+df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
+qplot(SC14Q02, mmath2, data = df2, facets = ~name)
+df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
+qplot(SC14Q10, mmath2, data = df3, facets = ~name)
+```
 
@@
 
+***Need to get this code working
+
+```
+# Out of School Study time
+b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
+names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
+# Out of school study time (total hours HW)
+b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
+b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
+b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
+qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
+head(b2)
+qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)
+
+student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
+study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
+names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
+levels(study$Subject) <- c("Math","Reading","Science")
+p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
+#ggsave(p,file="studytime.pdf",height=13,width=13)
+p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
+ggsave(p,file="HistHWschool.pdf",width=8,height=6)
+
+student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
+study <- student.sub
+names(study) <- c("Country","Math","HomeworkHours","Schoolplace")
+
+p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
+#ggsave(p,file="studytime.pdf",height=13,width=13)
+p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
+```
 ### More TVs yield higher math scores in the developing world, but lower in the developed countries

### 2014-08-29 paper/paper.Rmd
updated code

@@
 library(stringr)
-library(stringr)
 library(reshape2)
@@
 
-```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=TRUE}
 # This is the scaffolding to import the data
@@
 student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
-
 school2012 = fix_country(school2012)
@@
 
-```
+```{r, eval=FALSE}
 # Out of School Study time
@@
 
+```{r clean, echo=FALSE}
+student2012$CNT = student2012$name
+student2012$OECD <- as.character(student2012$OECD)
+student2012[student2012$CNT == "Albania", ]$OECD = "Albania"
+student2012$math = as.numeric(student2012[, "PV1MATH"])
+student2012$read = as.numeric(student2012[, "PV1READ"])
+student2012$sci = as.numeric(student2012[, "PV1SCIE"])
+
+student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
+
+for (i in 13:26) {
+    student2012.sub[, i] <- as.character(student2012.sub[, i])
+    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
+    student2012.sub[student2012.sub[, i] == "Yes", i] <- "1"
+    student2012.sub[student2012.sub[, i] == "No", i] <- "0"
+    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
+}
+for (i in 27:31) {
+    student2012.sub[, i] <- as.character(student2012.sub[, i])
+    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
+    student2012.sub[student2012.sub[, i] == "None", i] <- "0"
+    student2012.sub[student2012.sub[, i] == "One", i] <- "1"
+    student2012.sub[student2012.sub[, i] == "Two", i] <- "2"
+    student2012.sub[student2012.sub[, i] == "Three or more", i] <- "3"
+    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
+}
+student2012.sub[, 32] <- as.character(student2012.sub[, 32])
+student2012.sub[is.na(student2012.sub[,32]), 32] <- ""
+student2012.sub[student2012.sub[, 32] == "0-10 books ", 32] <- "0"
+student2012.sub[student2012.sub[, 32] == "11-25 books ", 32] <- "1"
+student2012.sub[student2012.sub[, 32] == "26-100 books ", 32] <- "2"
+student2012.sub[student2012.sub[, 32] == "101-200 books ", 32] <- "3"
+student2012.sub[student2012.sub[, 32] == "201-500 books ", 32] <- "4"
+student2012.sub[student2012.sub[, 32] == "More than 500 books", 32] <- "5"
+student2012.sub[, 32] <- as.numeric(student2012.sub[, 32])
+
+#aggregate possessions
+student2012.sub$numposs <- apply(student2012.sub[, 13:26], 1, sum, na.rm = T)
+student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
+```
+
+```{r albania-possessions, echo=FALSE}
+df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+
+df$group = paste0(df$CNT, df$variable)
+qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+```
+
+```{r truancyschool, echo=FALSE}
+school2012$CNT = school2012$name
+school2012.sub = school2012 %>% dplyr::select(CNT, SCHOOLID, SC22Q01)
+student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OECD)
+truancy.df = melt( student2012.sub %>% 
+  group_by(CNT, SCHOOLID) %>% 
+  left_join(school2012.sub, by = c("CNT", "SCHOOLID")) %>% 
+  group_by(CNT, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
+  id = c("CNT", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
+truancy.df$group = paste0(truancy.df$CNT, truancy.df$variable)
+
+qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+```
+
+```{r studytime, warning = FALSE, message=FALSE, echo=FALSE}
+.width = 1
+
+data_small <- 
+  student2012 %>%
+  dplyr::select(OECD, CNT, OUTHOURS) %>%
+  filter(!is.na(OUTHOURS))
+
+ggplot(mapping = aes(x = OUTHOURS)) + 
+  scale_x_continuous(name = "out of school study time (hours)", limits = c(0,60)) +
+  scale_y_continuous(name = "density") +
+  geom_histogram(
+    mapping = aes(y = ..density.., fill = OECD), 
+    binwidth = .width,
+    data = data_small %>% filter(OECD == "Albania")
+  ) + 
+  scale_fill_manual(
+    values = "black",
+    breaks = "Albania",
+    guide = guide_legend(title = "")
+  ) + 
+  geom_line(
+    mapping = aes(group = CNT, color = OECD), 
+    stat = "density", 
+    alpha = 0.3,
+    data = data_small %>% filter(OECD != "Albania")
+  ) +
+  scale_color_discrete(
+    guide = guide_legend(
+      title = "Other Countries",
+      override.aes = list(alpha = 1)
+    )
+
+  ) +
+  scale_colour_manual(values = c("#f4a582", "#92c5de")) + guides(colour = guide_legend(override.aes = list(alpha = 1)))
+```
+
+```{r gendergap, echo=FALSE, warning = FALSE, fig.height = 10}
+scores = student2012 %>%
+  dplyr::select(CNT, math, read, ST04Q01, OECD) %>% 
+  group_by(OECD, CNT, ST04Q01) %>%
+  summarise(math = mean(math), read = mean(read))
+male = scores %>% filter(ST04Q01 == "Male")
+female = scores %>% filter(ST04Q01 == "Female")
+scores.df = data.frame(OECD = male$OECD, CNT = male$CNT, math_difference = male$math - female$math, read_difference = male$read - female$read)
+
+scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$math_difference)])
+scores.df$OECD = factor(scores.df$OECD, levels = scores.df$OECD[])
+
+qplot(CNT, math_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
+  scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
+  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
+
+scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$read_difference)])
+qplot(CNT, read_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
+  scale_y_continuous("read score gap: male - female scores") +
+  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
+  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
+```
+
 ### Play with the data yourself

### 2014-08-29 paper/paper.Rmd
first pass at TV section

--- 2014-08-29 paper/paper.Rmd
@@
 library(ggmap)
+library(htmltools)
 library(rworldmap)
@@
 library (scales) 
-library(doBy)
 ```
@@
 
-```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
+```{r dataprep, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 colnames(student2012)[1] <- "name"
+# Now add data
 student2012$name <- as.character(student2012$name)
-colnames(school2012)[1] <- "name"
-school2012$name <- as.character(school2012$name)
-
-fix_country = function(df) {
-df$name = as.character(df$name)
-df$name[df$name=="Serbia"] <- "Republic of Serbia"
-df$name[df$name=="Korea"] <- "South Korea"
-df$name[df$name=="Chinese Taipei"] <- "Taiwan"
-df$name[df$name=="Slovak Republic"] <- "Slovakia"
-df$name[df$name=="Russian Federation"] <- "Russia"
-df$name[df$name=="Perm(Russian Federation)"] <- "Russia"
-df$name[df$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-df$name[df$name=="China-Shanghai"] <- "China"
-df$name[df$name=="China-Macau"] <- "China"
-df$name[df$name=="Connecticut (USA)"] <- "United States of America"
-df$name[df$name=="Florida (USA)"] <- "United States of America"
-df$name[df$name=="Massachusetts (USA)"] <- "United States of America"
-
-df
-}
-
-# Now add data
-#student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
-#student2012$name[student2012$name=="Korea"] <- "South Korea"
-#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-#student2012$name[student2012$name=="China-Shanghai"] <- "China"
-#student2012$name[student2012$name=="China-Macau"] <- "China"
-#student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
-#student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
-#student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
+student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
+student2012$name[student2012$name=="Korea"] <- "South Korea"
+student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
+student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
+student2012$name[student2012$name=="Russian Federation"] <- "Russia"
+student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
+student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
+student2012$name[student2012$name=="China-Shanghai"] <- "China"
+student2012$name[student2012$name=="China-Macau"] <- "China"
+student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
+student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
+student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
 student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
@@
 student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
-
-school2012 = fix_country(school2012)
-student2012 = fix_country(student2012)
 ```
 
-### Teachers, computers, and libraries at school make a difference in math scores
-
-```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
-school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
-student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)
-
-scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
-                                      mread = mean(PV1READ), msci = mean(PV1SCIE))
-
-df = scores_by_school %>% left_join(school2012.sub)
-df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
-qplot(SC14Q02, mmath2, data = df2, facets = ~name)
-df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
-qplot(SC14Q10, mmath2, data = df3, facets = ~name)
-```
+### Teachers, computers, and libraries at school matter
 
@@
 
-***Need to get this code working
+### More TVs yield higher math scores in the developing world, but lower in the developed countries
 
+One of the factors in the PISA 2012 dataset is the number of televisions in the student's household.
+In the dataset, possible values were: "zero", "one", "two", or "three or more".
+Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
+
+A linear regression as made for the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
+Using R's `lm()` function, the direction and significance of the correlation was determined for each country.
+Significance is indicated by the p-value for a given regression being less than the popular 0.05.
+
+For many countries, a significant relationship was be found between math scores and number of household televisions.
+The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
+
+Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
+
+It seems that television is a great equalizer.
+
+```{r place_map_countries, echo=FALSE}
+tags$div(
+  style = "height: 400px",
+  tags$p(
+    tags$img(
+      align = "right",
+      src = "img/ijl_map.png",
+      style="max-height: 100%"
+    ),
+    tags$img(
+      align = "left",
+      src = "img/ijl_countries.png",
+      style="max-height: 100%"
+    )
+  )
+)
 ```
-# Out of School Study time
-b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
-names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
-# Out of school study time (total hours HW)
-b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
-b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
-b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
-qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
-head(b2)
-qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)
 
-student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
-study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
-names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
-levels(study$Subject) <- c("Math","Reading","Science")
-p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
-ggsave(p,file="HistHWschool.pdf",width=8,height=6)
-
-student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
-study <- student.sub
-names(study) <- c("Country","Math","HomeworkHours","Schoolplace")
-
-p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
-```
-### More TVs yield higher math scores in the developing world, but lower in the developed countries
+Find your own correlations: http://bit.ly/pisa2012_explorer
 
@@
 
-Shiny app
+The PISA dataset contains dozens of household factors; we wanted a way to make our data-exploration more efficient.
+Moreover, we wanted a way that we could make explorations independent of each other, as during the course of the invesigation our team was in three cities over two hemispheres.
+
+To make our exploration easier, we developed a "Shiny" app.
+
+Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.
 
@@
 
-Software packages
+Shiny, by RStudio
+
+plyr, dplyr, and ggplot2, by Hadley Wickham

### 2014-08-29 paper/paper.Rmd

--- 2014-08-29 paper/paper.Rmd
@@
 
-It seems that television is the great equalizer. As part of our investigation, we set out to explore how household factors, as indicators of inequality, affect student test scores.
+One of the factors in the PISA 2012 dataset is the number of televisions in the student's household.
+In the dataset, possible values were: "zero", "one", "two", or "three or more".
+Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
 
-Math scores correlate with the number of household TVs, but how depends on where you live. Nevertheless, access to a quiet study space is generally a plus.
+A linear regression as made for the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
+Using R's `lm()` function, the direction and significance of the correlation was determined for each country.
+Significance is indicated by the p-value for a given regression being less than the popular 0.05.
 
-For each country, students’ math scores are compared with number of TVs at home.
-Countries are grouped by direction and strength of the correlation; groups of countries are mapped (above) and identified (right).
+For many countries, a significant relationship was be found between math scores and number of household televisions.
+The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
 
-Students are considered by group, and categorized by household factors: number of TVs, and access to a quiet study space.
-At far left, number of students in each category; at near left, math scores are summarized for students in each category.
+Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
+
+It seems that television is a great equalizer.
+
+```{r place_map_countries, echo=FALSE}
+tags$div(
+  style = "height: 400px",
+  tags$p(
+    tags$img(
+      align = "right",
+      src = "img/ijl_map.png",
+      style="max-height: 100%"
+    ),
+    tags$img(
+      align = "left",
+      src = "img/ijl_countries.png",
+      style="max-height: 100%"
+    )
+  )
+)
+```
 

### 2014-08-29 paper/paper.Rmd
*very* rough treatment of shiny app

@@
 
-Find your own correlations: http://bit.ly/pisa2012_explorer
-
 ### Parents matter
@@
 
-The PISA dataset contains dozens of household factors; we wanted a way to make our data-exploration more efficient.
-Moreover, we wanted a way that we could make explorations independent of each other, as during the course of the invesigation our team was in three cities over two hemispheres.
+The PISA dataset contains hundreds of factors; we wanted a way to make our data-exploration more efficient.
 
-To make our exploration easier, we developed a "Shiny" app.
+Moreover, for much of the time when the authors made their initial investigations, we (they?) were remote from each other: four team-members, in three cities across, two hemispheres.
+Thanks to a ten-hour dime difference, we would meet, virtually, only occasionally, and we had to make the most of our limited time.
+
+To make our exploration easier, we developed a "Shiny" app, so that we could show each other our ideas interactively.
+
+"Let's look at *this* factor, combined with *that* factor!"
+
+"Let's look at that other factor, instead."
 
 Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.
+
+We then decided to look at only household factors. Inner factor, using boxplots. Outer factor, using violin plots. Added correlation by country.
+
+Find your own correlations: http://bit.ly/pisa2012_explorer
 

### 2014-08-29 paper/paper.Rmd
revised TVs section, added screenshot of shiny app, and revised play section

--- 2014-08-29 paper/paper.Rmd
@@
 
-One of the factors in the PISA 2012 dataset is the number of televisions in the student's household.
-In the dataset, possible values were: "zero", "one", "two", or "three or more".
+One of the factors in the data is the number of televisions in the student's household.
+Possible values were: "zero", "one", "two", or "three or more".
 Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
 
-A linear regression as made for the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
-Using R's `lm()` function, the direction and significance of the correlation was determined for each country.
-Significance is indicated by the p-value for a given regression being less than the popular 0.05.
+A linear model is fitted the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
+The direction and significance of the correlation was determined for each country.
+Significance is indicated by the p-value for a given regression being less than the conventional 0.05.
 
@@
 The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
+
+The figure below illustrates the relationship, along with another factor, student has a place to study.
+It is a complex diagram, but of interest is the middle column of plots - the boxplots overlaid on other information show the math scores of students relative to the number of TVs in the household, separately plotted for countries falling into one of three categories, negative, zero or  positive correlation.
+The white regions are violin plots of math scores by presence or absence of a place to study.
+A place to study universally leads to better math scores on average.
+The map is colored by correlation between number of TVs and math scores.
+The bar charts indicates sample size for the categories.
+The dot plots show average math scores by country, separately for each correlation group.
 
@@
 
-```{r place_map_countries, echo=FALSE}
-tags$div(
-  style = "height: 400px",
-  tags$p(
-    tags$img(
-      align = "right",
-      src = "img/ijl_map.png",
-      style="max-height: 100%"
-    ),
-    tags$img(
-      align = "left",
-      src = "img/ijl_countries.png",
-      style="max-height: 100%"
-    )
-  )
-)
-```
-
-Find your own correlations: http://bit.ly/pisa2012_explorer
+!["TVs"](shiny-pisa.jpg)
 
@@
 
-The PISA dataset contains dozens of household factors; we wanted a way to make our data-exploration more efficient.
-Moreover, we wanted a way that we could make explorations independent of each other, as during the course of the invesigation our team was in three cities over two hemispheres.
-
-To make our exploration easier, we developed a "Shiny" app.
-
+The PISA dataset contains hundreds of factors, and this article can describe just a few findings.
+The reader is welcome to explore the data themselves using an <a href="http://bit.ly/pisa2012_explorer"> interactive web tool</a>, made with <a href="http://www.rstudio.com">RStudio</a> and Shiny.
 Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.

### 2014-08-29 paper/paper.Rmd
added first part of Luke’s parents matter

@@
 ```
+
 ### More TVs yield higher math scores in the developing world, but lower in the developed countries
@@
 
+The role of parents in a child's education and success later in life is of great interest to both educators and policy makers.
+The relationship between eduational achievement and both household structure and parent's occupational status were evaluated.
+To look at the relationship between educational success and family structure, boxplots were made of the student's scores conditioned whether the student had a two-parent household, only a mother at home, only a father at home, or no parents at home.
+The relationship was evaluated for every participating country.
+The majority of the students did live in a two-parent household, but there were still sizable numbers of students who did not.
+The information regarding a parent's occupational was provided in a way that can be broadly catergorized with five catergories: "Full-employment", "Part-time", "Unemployed-looking for work", "Other: Home duties, retired, etc.", and "Missing".
+The means achievement in mathematics for each of the categories was calculated and plotted for each country.
+
+```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
+athome = filter(athome, !is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
+athome$parents = NA
+athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
+athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
+athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
+athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
+#athome$name <- reorder(athome$name, athome$PV1MATH, mean, decreasing=F)
+#qplot(x = parents, data = athome) 
+```
+
+```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math Scores") + xlab("Family Structure") 
+```
 ### The weird and wacky republic of Albania

### 2014-08-29 paper/paper.Rmd
added second component of parents matter from Luke

@@
 ```
+
+One of the variables the survey addressed was the occupation of each of the parents.
+There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g.
+home duties, retired)”, and no response.
+The distribution of the number of parents who selected each of the categories can be seen in the bar chart.
+The chart shows that father’s primarily have full-time jobs.
+However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms.
+With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.
+
+```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+#student2012 <- student2012
+student2012$ST15Q01 <- addNA(student2012$ST15Q01)
+student2012$ST19Q01 <- addNA(student2012$ST19Q01)
+
+#student2012$name <- as.character(student2012$CNT)
+# unique(anti_join(student2012.sub, world.polys)[1])
+#student2012$name[student2012$name=="United Arab Emirates"] <- "UAE"
+#student2012$name[student2012$name=="United Kingdom"] <- "UK"
+#student2012$name[student2012$name=="Serbia"] <- "Serbia"
+#student2012$name[student2012$name=="Korea"] <- "South Korea"
+#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
+#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
+#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
+#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
+#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong"
+#student2012$name[student2012$name=="China-Shanghai"] <- "China"
+#student2012$name[student2012$name=="Macau"] <- "China"
+#student2012$name[student2012$name=="Connecticut (USA)"] <- "USA"
+#student2012$name[student2012$name=="Florida (USA)"] <- "USA"
+#student2012$name[student2012$name=="Massachusetts (USA)"] <- "USA"
+#student2012$name[student2012$name=="United States of America"] <- "USA"
+#unique(student2012$name)
+#student2012$name <- factor(student2012$name)
+
+a6 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST15Q01+name, FUN=mean, rm.na=T)
+a7 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST19Q01+name, FUN=mean, rm.na=T)
+colnames(a6) <- colnames(a7) <- c("Job.Status","name","Math","ESCS")
+levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")
+
+a8 <- rbind(cbind(a6, Parent="Mother"), cbind(a7, Parent="Father"))
+#a2 <- melt(a1)
+ind <- which(a8$Job.Status=="Other" & a8$Parent=="Mother")
+a8$name <- factor(a8$name, a8$name[order(a8$Math[ind])])
+qplot(name, Math, col=Job.Status, data = a8) + coord_flip() + facet_wrap(~Parent)
+```
+
 ### The weird and wacky republic of Albania

### 2014-08-29 paper/paper.Rmd
tweaked albania

@@
 student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
-qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Score", 
+qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Scores", 
       geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
@@
 
-### More TVs yield higher math scores in the developing world, but lower in the developed countries
-
-One of the factors in the data is the number of televisions in the student's household.
-Possible values were: "zero", "one", "two", or "three or more".
-Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
-
-A linear model is fitted the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
-The direction and significance of the correlation was determined for each country.
-Significance is indicated by the p-value for a given regression being less than the conventional 0.05.
-
-For many countries, a significant relationship was be found between math scores and number of household televisions.
-The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
-
-The figure below illustrates the relationship, along with another factor, student has a place to study.
-It is a complex diagram, but of interest is the middle column of plots - the boxplots overlaid on other information show the math scores of students relative to the number of TVs in the household, separately plotted for countries falling into one of three categories, negative, zero or  positive correlation.
-The white regions are violin plots of math scores by presence or absence of a place to study.
-A place to study universally leads to better math scores on average.
-The map is colored by correlation between number of TVs and math scores.
-The bar charts indicates sample size for the categories.
-The dot plots show average math scores by country, separately for each correlation group.
-
-Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
-
-It seems that television is a great equalizer.
-
-!["TVs"](shiny-pisa.jpg)
-
 ### Parents matter
@@
 ```{r clean, echo=FALSE}
-student2012$CNT = student2012$name
+#student2012$CNT = student2012$name
 student2012$OECD <- as.character(student2012$OECD)
-student2012[student2012$CNT == "Albania", ]$OECD = "Albania"
+#student2012[student2012$name == "Albania", ]$OECD = "Albania"
 student2012$math = as.numeric(student2012[, "PV1MATH"])
@@
 
-student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
+student2012.sub = student2012 %>% dplyr::select(name, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
 
@@
 
-```{r albania-possessions, echo=FALSE}
-df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+```{r albania-possessions, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+#df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("name", "numposs", "OECD"))
+df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math)), id=c("name", "numposs", "OECD"))
 
-df$group = paste0(df$CNT, df$variable)
-qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+#df$group = paste0(df$name, df$variable)
+#qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+qplot(x = numposs, y = value, data = df, facets = ~name, geom = "line") + ylab("Math Scores") + xlab("Number of Possessions") + theme_bw()
 ```
 
-```{r truancyschool, echo=FALSE}
-school2012$CNT = school2012$name
-school2012.sub = school2012 %>% dplyr::select(CNT, SCHOOLID, SC22Q01)
-student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OECD)
+```{r truancyschool, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+#school2012$name = school2012$name
+school2012.sub = school2012 %>% dplyr::select(name, SCHOOLID, SC22Q01)
+student2012.sub = student2012 %>% dplyr::select(name, SCHOOLID, math, read, sci, OECD)
+#truancy.df = melt( student2012.sub %>% 
+#  group_by(name, SCHOOLID) %>% 
+#  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
+#  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
+#  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
 truancy.df = melt( student2012.sub %>% 
-  group_by(CNT, SCHOOLID) %>% 
-  left_join(school2012.sub, by = c("CNT", "SCHOOLID")) %>% 
-  group_by(CNT, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
-  id = c("CNT", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
-truancy.df$group = paste0(truancy.df$CNT, truancy.df$variable)
+  group_by(name, SCHOOLID) %>% 
+  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
+  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math)), 
+  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
+#truancy.df$group = paste0(truancy.df$name, truancy.df$variable)
 
-qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+#qplot(x = SC22Q01, y = value, color = variable, facets = ~name, data = truancy.df, group = group, geom = "line") + xlab("Reported Truancy") + ylab("Test Scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+qplot(x = SC22Q01, y = value, facets = ~name, data = truancy.df, geom = "line", group=name) + xlab("Reported Truancy") + ylab("Math Scores") + theme_bw()
 ```
@@
   student2012 %>%
-  dplyr::select(OECD, CNT, OUTHOURS) %>%
+  dplyr::select(OECD, name, OUTHOURS) %>%
   filter(!is.na(OUTHOURS))
@@
   geom_line(
-    mapping = aes(group = CNT, color = OECD), 
+    mapping = aes(group = name, color = OECD), 
     stat = "density", 
@@
 
-```{r gendergap, echo=FALSE, warning = FALSE, fig.height = 10}
-scores = student2012 %>%
-  dplyr::select(CNT, math, read, ST04Q01, OECD) %>% 
-  group_by(OECD, CNT, ST04Q01) %>%
-  summarise(math = mean(math), read = mean(read))
-male = scores %>% filter(ST04Q01 == "Male")
-female = scores %>% filter(ST04Q01 == "Female")
-scores.df = data.frame(OECD = male$OECD, CNT = male$CNT, math_difference = male$math - female$math, read_difference = male$read - female$read)
+### More TVs yield higher math scores in the developing world, but lower in the developed countries
 
-scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$math_difference)])
-scores.df$OECD = factor(scores.df$OECD, levels = scores.df$OECD[])
+One of the factors in the data is the number of televisions in the student's household.
+Possible values were: "zero", "one", "two", or "three or more".
+Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
 
-qplot(CNT, math_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
-  scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
-  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
+A linear model is fitted the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
+The direction and significance of the correlation was determined for each country.
+Significance is indicated by the p-value for a given regression being less than the conventional 0.05.
 
-scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$read_difference)])
-qplot(CNT, read_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
-  scale_y_continuous("read score gap: male - female scores") +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
-  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
-```
+For many countries, a significant relationship was be found between math scores and number of household televisions.
+The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
+
+The figure below illustrates the relationship, along with another factor, student has a place to study.
+It is a complex diagram, but of interest is the middle column of plots - the boxplots overlaid on other information show the math scores of students relative to the number of TVs in the household, separately plotted for countries falling into one of three categories, negative, zero or  positive correlation.
+The white regions are violin plots of math scores by presence or absence of a place to study.
+A place to study universally leads to better math scores on average.
+The map is colored by correlation between number of TVs and math scores.
+The bar charts indicates sample size for the categories.
+The dot plots show average math scores by country, separately for each correlation group.
+
+Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
+
+It seems that television is a great equalizer.
+
+!["TVs"](shiny-pisa.jpg)
 

### 2014-08-29 paper/paperLuke.Rmd
test commit to pull/commit
tsting

--- 2014-08-29 paper/paper.Rmd
@@
+---
+output: pdf_document
+---
 Seven Things That Affect Education Around the World
@@
 Students are examined on how well they can apply the knowledge they learned in school to applications outside of school.
-The scores reported range between 0-1000.
 Additional information about the students, parents, and schools is also collected.
@@
 library(stringr)
+library(stringr)
 library(reshape2)
@@
 library(ggmap)
-library(htmltools)
 library(rworldmap)
@@
 library (scales) 
-library(doBy)
 ```
 
-```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=TRUE}
+```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
 # This is the scaffolding to import the data
@@
 ```
+### Teachers, computers, and libraries at school matter
 
-*** We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
+### Gender gap in math is not universal but the reading gap is
 
-```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
-colnames(student2012)[1] <- "name"
-student2012$name <- as.character(student2012$name)
-colnames(school2012)[1] <- "name"
-school2012$name <- as.character(school2012$name)
+### School start age affects performance
 
-fix_country = function(df) {
-df$name = as.character(df$name)
-df$name[df$name=="Serbia"] <- "Republic of Serbia"
-df$name[df$name=="Korea"] <- "South Korea"
-df$name[df$name=="Chinese Taipei"] <- "Taiwan"
-df$name[df$name=="Slovak Republic"] <- "Slovakia"
-df$name[df$name=="Russian Federation"] <- "Russia"
-df$name[df$name=="Perm(Russian Federation)"] <- "Russia"
-df$name[df$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-df$name[df$name=="China-Shanghai"] <- "China"
-df$name[df$name=="China-Macau"] <- "China"
-df$name[df$name=="Connecticut (USA)"] <- "United States of America"
-df$name[df$name=="Florida (USA)"] <- "United States of America"
-df$name[df$name=="Massachusetts (USA)"] <- "United States of America"
+***We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
 
-df
-}
-
-# Now add data
-#student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
-#student2012$name[student2012$name=="Korea"] <- "South Korea"
-#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
-#student2012$name[student2012$name=="China-Shanghai"] <- "China"
-#student2012$name[student2012$name=="China-Macau"] <- "China"
-#student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
-#student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
-#student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
-student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
-student2012$PV1MATH <- as.numeric(student2012$PV1MATH)
-student2012$PV1READ <- as.numeric(student2012$PV1READ)
-student2012$PV1SCIE <- as.numeric(student2012$PV1SCIE)
-student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
-school2012 = fix_country(school2012)
-student2012 = fix_country(student2012)
-```
-
-### Teachers, computers, and libraries at school make a difference in math scores
-
-```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
-school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
-student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)
-
-scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
-                                      mread = mean(PV1READ), msci = mean(PV1SCIE))
-
-df = scores_by_school %>% left_join(school2012.sub)
-df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
-qplot(SC14Q02, mmath2, data = df2, facets = ~name)
-df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
-qplot(SC14Q10, mmath2, data = df3, facets = ~name)
-```
-
-### The gender gap in math is not universal but the reading gap is, in favor of girls
-
-For each country, average scores on math and reading are calculated separately for boys and girls.
-These averages are differenced to compute the "gap" between boys and girls.
-T-statistics were calculated to determine the statistical significance of the difference.
-The results are plotted in the graphs below.
-Each dot corresponds to a difference between girls and boys scores, for each country.
-Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference).
-The size of the dot indicates the proportion of boys in the sample size.
-For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys.
-The maps color the countries by the presence of a gender gap.
-
-The countries are sorted from biggest gap to smallest gap.
-Most countries have a gender gap in math scores in favor of boys.
-Colombia has the biggest gender gap, about 30 points on average boys do better than girls.
-It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.
-Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.
-
-Conversely, though, when we look at reading scores the story is entirely different.
-Universally, girls score significantly better than boys.
-Again, the size of the gap is small, between 10 and 80 points.
-
-Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls.
-There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score.
-There is far more individual variability than the overall average differences.
-
-```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
-student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
-colnames(student2012.sub)[1] <- "name"
-student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
-student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
-student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
-student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
-student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
-                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
-                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
-                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
-                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
-                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
-                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
-                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
-                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
-                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
-                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
-                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
-                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
-                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
-                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
-                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
-                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
-                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
-                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
-#qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
-#      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
-student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
-student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
-qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
-#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
-#qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-#  xlab("") +
-#  geom_point(aes(y=maxfemale), colour="pink") + 
-#  scale_y_continuous("High Math") + coord_flip() + theme_bw()
-#ggsave("gendermathtop.pdf", width=3.5, height=7)
-#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
-#qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
-#  xlab("") +
-#  geom_point(aes(y=minfemale), colour="pink") + 
-#  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
-#ggsave("gendermathbottom.pdf", width=3.5, height=7)
-#
-student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
-student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
-student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
-student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
-qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
-  xlab("") +
-  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Reading Score Gap") + 
-  scale_size("Prop male") +
-   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
-#ggsave("genderreadgap.pdf", width=5, height=8)
-#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
-#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
-#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
-#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
-#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
-#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
-#  xlab("") +
-#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-#  scale_y_continuous("Science Score Gap") + 
-#  scale_size("Prop male") +
-#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
-```
-
-```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
-student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
-#ggsave("gendermathmap.pdf", width=12, height=7)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
-#ggsave("genderreadmap.pdf", width=12, height=7)
-#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
-#  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-#  new_theme_empty + theme(legend.position="none")
-```
-
-### Starting school at an older age generated lower average math score.
-
-For each country we have computed the average math test score by age that the children started school.
+Broken out by country we have computed the average math test score by age that the children started school.
 In additon, for each country we computed the median age that children in the study started school.
@@
 
-```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
+```{r maps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
 # Extract map polygons for modern world
@@
 ### Time spent out of school studying is important, but only up to a point
+For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school.
+The most prominent feature that we observe is the decreasing benefit of studying.
+For every country there is evidence of a substantial positive relationship for the students who spend less time studying.
+However, those students who continue to study for long hours see diminishing returns for their efforts.
+It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster.
+However, the relationship shows those spending
 
-***Need to get this code working
+```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
+study.luke <- melt(student.luke,id=c("CNT","ST57Q01"))
+studyplot <- ggplot(data = study.luke, aes(ST57Q01,value,colour=variable) )+facet_wrap(~CNT)+stat_smooth(se=F)
+studyplot
+```
 
-```{r, eval=FALSE}
-# Out of School Study time
-b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
-names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
-# Out of school study time (total hours HW)
-b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
-b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
-b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
-qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
-head(b2)
-qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)
-
-student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
-study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
-names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
-levels(study$Subject) <- c("Math","Reading","Science")
-p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
-ggsave(p,file="HistHWschool.pdf",width=8,height=6)
-
-student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
-study <- student.sub
-names(study) <- c("Country","Math","HomeworkHours","Schoolplace")
-
-p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
+```{r studyingbar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
+studybar
 ```
@@
 ### More TVs yield higher math scores in the developing world, but lower in the developed countries
-
-One of the factors in the data is the number of televisions in the student's household.
-Possible values were: "zero", "one", "two", or "three or more".
-Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
-
-A linear model is fitted the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
-The direction and significance of the correlation was determined for each country.
-Significance is indicated by the p-value for a given regression being less than the conventional 0.05.
-
-For many countries, a significant relationship was be found between math scores and number of household televisions.
-The interesting part was to notice that those countries with a significant *positive* relationship appear to include much of the developing world (with lower median-scores), while those countries with a significant *negative* relationship appear to include much of the developed world (with higher median-scores).
-
-The figure below illustrates the relationship, along with another factor, student has a place to study.
-It is a complex diagram, but of interest is the middle column of plots - the boxplots overlaid on other information show the math scores of students relative to the number of TVs in the household, separately plotted for countries falling into one of three categories, negative, zero or  positive correlation.
-The white regions are violin plots of math scores by presence or absence of a place to study.
-A place to study universally leads to better math scores on average.
-The map is colored by correlation between number of TVs and math scores.
-The bar charts indicates sample size for the categories.
-The dot plots show average math scores by country, separately for each correlation group.
-
-Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
-
-It seems that television is a great equalizer.
-
-!["TVs"](shiny-pisa.jpg)
 
@@
 
+### subsection: Number of parents in the home
+Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household.
+Students with just a mother at home do tend to perform better than students with only a father at home.
+Students with no parents at home are performing the worst, on average.
+
 ```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
-athome = filter(athome, !is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
+athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
+athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
 athome$parents = NA
@@
 athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
-#athome$name <- reorder(athome$name, athome$PV1MATH, mean, decreasing=F)
-#qplot(x = parents, data = athome) 
+athome$name <- reorder(athome$name,athome$PV1MATH,mean)
+pbarhome <- qplot(x = parents, data = athome) 
+pbarhome
+
 ```
@@
 ```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math Scores") + xlab("Family Structure") 
-```
-### The weird and wacky republic of Albania
-
-```{r clean, echo=FALSE}
-student2012$CNT = student2012$name
-student2012$OECD <- as.character(student2012$OECD)
-student2012[student2012$CNT == "Albania", ]$OECD = "Albania"
-student2012$math = as.numeric(student2012[, "PV1MATH"])
-student2012$read = as.numeric(student2012[, "PV1READ"])
-student2012$sci = as.numeric(student2012[, "PV1SCIE"])
-
-student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OUTHOURS, STIDSTD, ST04Q01, ST11Q01, ST11Q02, ST15Q01, ST19Q01, ST26Q01, ST26Q02, ST26Q03, ST26Q04, ST26Q05, ST26Q06, ST26Q07, ST26Q08, ST26Q09, ST26Q10, ST26Q11, ST26Q12, ST26Q13, ST26Q14, ST27Q01, ST27Q02, ST27Q03, ST27Q04, ST27Q05, ST28Q01, OECD)
-
-for (i in 13:26) {
-    student2012.sub[, i] <- as.character(student2012.sub[, i])
-    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
-    student2012.sub[student2012.sub[, i] == "Yes", i] <- "1"
-    student2012.sub[student2012.sub[, i] == "No", i] <- "0"
-    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
-}
-for (i in 27:31) {
-    student2012.sub[, i] <- as.character(student2012.sub[, i])
-    student2012.sub[is.na(student2012.sub[, i]), i] <- ""
-    student2012.sub[student2012.sub[, i] == "None", i] <- "0"
-    student2012.sub[student2012.sub[, i] == "One", i] <- "1"
-    student2012.sub[student2012.sub[, i] == "Two", i] <- "2"
-    student2012.sub[student2012.sub[, i] == "Three or more", i] <- "3"
-    student2012.sub[, i] <- as.numeric(student2012.sub[, i])
-}
-student2012.sub[, 32] <- as.character(student2012.sub[, 32])
-student2012.sub[is.na(student2012.sub[,32]), 32] <- ""
-student2012.sub[student2012.sub[, 32] == "0-10 books ", 32] <- "0"
-student2012.sub[student2012.sub[, 32] == "11-25 books ", 32] <- "1"
-student2012.sub[student2012.sub[, 32] == "26-100 books ", 32] <- "2"
-student2012.sub[student2012.sub[, 32] == "101-200 books ", 32] <- "3"
-student2012.sub[student2012.sub[, 32] == "201-500 books ", 32] <- "4"
-student2012.sub[student2012.sub[, 32] == "More than 500 books", 32] <- "5"
-student2012.sub[, 32] <- as.numeric(student2012.sub[, 32])
-
-#aggregate possessions
-student2012.sub$numposs <- apply(student2012.sub[, 13:26], 1, sum, na.rm = T)
-student2012.sub$numedposs <- apply(student2012.sub[, 27:31], 1, sum, na.rm = T)
+phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
+phome
 ```
 
-```{r albania-possessions, echo=FALSE}
-df = melt(student2012.sub %>% group_by(CNT, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("CNT", "numposs", "OECD"))
+### subsection: Parent's Occupation Status
+One of the variables the survey addressed was the occupation status of each of the parents.
+There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g.
+home duties, retired)”, and no response.
+The distribution of the number of parents who selected each of the categories can be seen in the bar chart.
+The chart shows that father’s primarily have full-time jobs.
+However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms.
+With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.
 
-df$group = paste0(df$CNT, df$variable)
-qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
+```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+studentLuke <- student2012
+studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
+studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)
+
+studentLuke$name <- as.character(studentLuke$CNT)
+# unique(anti_join(studentLuke.sub, world.polys)[1])
+studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
+studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
+studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
+studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
+studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
+studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
+studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
+studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
+studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
+studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
+studentLuke$name[studentLuke$name=="Macau"] <- "China"
+studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
+unique(studentLuke$name)
+studentLuke$name <- factor(studentLuke$name)
+
+a6 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST15Q01+ name,FUN=mean,rm.na=T)
+a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
+colnames(a6) <- colnames(a7) <- c("Job.Status","Country","Math","ESCS")
+levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")
+
+a8 <- rbind(cbind(a6,Parent="Mother"),cbind(a7,Parent="Father"))
+#a2 <- melt(a1)
+ind <- which(a8$Job.Status=="Other"& a8$Parent=="Mother")
+a8$Country <- factor(a8$Country,levels(a8$Country)[order(a8$Math[ind])])
+p <- qplot(Country, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
+p
 ```
 
-```{r truancyschool, echo=FALSE}
-school2012$CNT = school2012$name
-school2012.sub = school2012 %>% dplyr::select(CNT, SCHOOLID, SC22Q01)
-student2012.sub = student2012 %>% dplyr::select(CNT, SCHOOLID, math, read, sci, OECD)
-truancy.df = melt( student2012.sub %>% 
-  group_by(CNT, SCHOOLID) %>% 
-  left_join(school2012.sub, by = c("CNT", "SCHOOLID")) %>% 
-  group_by(CNT, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
-  id = c("CNT", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
-truancy.df$group = paste0(truancy.df$CNT, truancy.df$variable)
+The Figure ?*?
+shows the average performance of student’s performance based on the parent’s occupational status.
+The countries are sorted according to the average score for students with a "stay-at-home" mother.
+The averages for each parent in each of the work status categories are distinguished by color.
+As you can see in each of the plots below, students’ with a mother who works part-time or full-time tend outperform those students with a stay-at-home mother.
+This pattern holds with only a few exceptions.
+Meanwhile, when the father is employed part-time students tend to have much lower achievement that if the father stayed at home.
+When the father has a full-time jobs, the student’s have the highest average achievement among the possible work status’ for the father.
+The patterns described hold across mathematics, reading, and science.
 
-qplot(x = SC22Q01, y = value, color = variable, facets = ~OECD, data = truancy.df, group = group, geom = "line") + xlab("reported truancy") + ylab("test scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
-```
-
-```{r studytime, warning = FALSE, message=FALSE, echo=FALSE}
-.width = 1
-
-data_small <- 
-  student2012 %>%
-  dplyr::select(OECD, CNT, OUTHOURS) %>%
-  filter(!is.na(OUTHOURS))
-
-ggplot(mapping = aes(x = OUTHOURS)) + 
-  scale_x_continuous(name = "out of school study time (hours)", limits = c(0,60)) +
-  scale_y_continuous(name = "density") +
-  geom_histogram(
-    mapping = aes(y = ..density.., fill = OECD), 
-    binwidth = .width,
-    data = data_small %>% filter(OECD == "Albania")
-  ) + 
-  scale_fill_manual(
-    values = "black",
-    breaks = "Albania",
-    guide = guide_legend(title = "")
-  ) + 
-  geom_line(
-    mapping = aes(group = CNT, color = OECD), 
-    stat = "density", 
-    alpha = 0.3,
-    data = data_small %>% filter(OECD != "Albania")
-  ) +
-  scale_color_discrete(
-    guide = guide_legend(
-      title = "Other Countries",
-      override.aes = list(alpha = 1)
-    )
-
-  ) +
-  scale_colour_manual(values = c("#f4a582", "#92c5de")) + guides(colour = guide_legend(override.aes = list(alpha = 1)))
-```
-
-```{r gendergap, echo=FALSE, warning = FALSE, fig.height = 10}
-scores = student2012 %>%
-  dplyr::select(CNT, math, read, ST04Q01, OECD) %>% 
-  group_by(OECD, CNT, ST04Q01) %>%
-  summarise(math = mean(math), read = mean(read))
-male = scores %>% filter(ST04Q01 == "Male")
-female = scores %>% filter(ST04Q01 == "Female")
-scores.df = data.frame(OECD = male$OECD, CNT = male$CNT, math_difference = male$math - female$math, read_difference = male$read - female$read)
-
-scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$math_difference)])
-scores.df$OECD = factor(scores.df$OECD, levels = scores.df$OECD[])
-
-qplot(CNT, math_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
-  scale_y_continuous("math score gap: male - female scores", breaks=seq(-30, 30, 5)) +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
-  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
-
-scores.df$CNT = factor(scores.df$CNT, levels = scores.df$CNT[order(scores.df$read_difference)])
-qplot(CNT, read_difference, data = scores.df, color = OECD, size = OECD) + coord_flip() +
-  scale_y_continuous("read score gap: male - female scores") +
-  geom_hline(yintercept=0, colour="grey80") + scale_size_manual(values = c(4, 2.5, 2.5)) + xlab("") +  
-  scale_colour_manual(values = c("#ca0020", "#f4a582", "#92c5de"))
-```
-
-### Play with the data yourself
-
-The PISA dataset contains hundreds of factors, and this article can describe just a few findings.
-The reader is welcome to explore the data themselves using an <a href="http://bit.ly/pisa2012_explorer"> interactive web tool</a>, made with <a href="http://www.rstudio.com">RStudio</a> and Shiny.
-Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.
-
-### Acknowledgements
-
-Shiny, by RStudio
-
-plyr, dplyr, and ggplot2, by Hadley Wickham
+### The weird and wacky republic of Albania

### 2014-08-29 paper/paper.Rmd
updated with luke's code and comments

@@
 ### Time spent out of school studying is important, but only up to a point
+### Time spent out of school studying is important, but only up to a point
+For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school.
+The most prominent feature that we observe is the decreasing benefit of studying.
+For every country there is evidence of a substantial positive relationship for the students who spend less time studying.
+However, those students who continue to study for long hours see diminishing returns for their efforts.
+It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster.
+However, the relationship shows those spending
 
-***Need to get this code working
+```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
+study.luke <- melt(student.luke,id=c("CNT","ST57Q01"))
+studyplot <- ggplot(data = study.luke, aes(ST57Q01,value,colour=variable) )+facet_wrap(~CNT)+stat_smooth(se=F)
+studyplot
+```
 
-```{r, eval=FALSE}
-# Out of School Study time
-b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
-names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
-# Out of school study time (total hours HW)
-b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
-b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
-b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
-qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
-head(b2)
-qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)
-
-student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
-study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
-names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
-levels(study$Subject) <- c("Math","Reading","Science")
-p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
-ggsave(p,file="HistHWschool.pdf",width=8,height=6)
-
-student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
-study <- student.sub
-names(study) <- c("Country","Math","HomeworkHours","Schoolplace")
-
-p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
-#ggsave(p,file="studytime.pdf",height=13,width=13)
-p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
+```{r studyingbar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
+studybar
 ```
@@
 ### Parents matter
-
-The role of parents in a child's education and success later in life is of great interest to both educators and policy makers.
-The relationship between eduational achievement and both household structure and parent's occupational status were evaluated.
-To look at the relationship between educational success and family structure, boxplots were made of the student's scores conditioned whether the student had a two-parent household, only a mother at home, only a father at home, or no parents at home.
-The relationship was evaluated for every participating country.
-The majority of the students did live in a two-parent household, but there were still sizable numbers of students who did not.
-The information regarding a parent's occupational was provided in a way that can be broadly catergorized with five catergories: "Full-employment", "Part-time", "Unemployed-looking for work", "Other: Home duties, retired, etc.", and "Missing".
-The means achievement in mathematics for each of the categories was calculated and plotted for each country.
+### subsection: Number of parents in the home
+Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household.
+Students with just a mother at home do tend to perform better than students with only a father at home.
+Students with no parents at home are performing the worst, on average.
 
 ```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
-athome = filter(athome, !is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
+athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
+athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
 athome$parents = NA
@@
 ```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math Scores") + xlab("Family Structure") 
+phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
+phome
 ```
 
-One of the variables the survey addressed was the occupation of each of the parents.
+### subsection: Parent's Occupation Status
+One of the variables the survey addressed was the occupation status of each of the parents.
 There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g.
@@
 ```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-#student2012 <- student2012
-student2012$ST15Q01 <- addNA(student2012$ST15Q01)
-student2012$ST19Q01 <- addNA(student2012$ST19Q01)
+studentLuke <- student2012
+studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
+studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)
 
-#student2012$name <- as.character(student2012$CNT)
-# unique(anti_join(student2012.sub, world.polys)[1])
-#student2012$name[student2012$name=="United Arab Emirates"] <- "UAE"
-#student2012$name[student2012$name=="United Kingdom"] <- "UK"
-#student2012$name[student2012$name=="Serbia"] <- "Serbia"
-#student2012$name[student2012$name=="Korea"] <- "South Korea"
-#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
-#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
-#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
-#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
-#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong"
-#student2012$name[student2012$name=="China-Shanghai"] <- "China"
-#student2012$name[student2012$name=="Macau"] <- "China"
-#student2012$name[student2012$name=="Connecticut (USA)"] <- "USA"
-#student2012$name[student2012$name=="Florida (USA)"] <- "USA"
-#student2012$name[student2012$name=="Massachusetts (USA)"] <- "USA"
-#student2012$name[student2012$name=="United States of America"] <- "USA"
-#unique(student2012$name)
-#student2012$name <- factor(student2012$name)
+studentLuke$name <- as.character(studentLuke$CNT)
+# unique(anti_join(studentLuke.sub, world.polys)[1])
+studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
+studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
+studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
+studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
+studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
+studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
+studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
+studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
+studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
+studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
+studentLuke$name[studentLuke$name=="Macau"] <- "China"
+studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
+studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
+unique(studentLuke$name)
+studentLuke$name <- factor(studentLuke$name)
 
-a6 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST15Q01+name, FUN=mean, rm.na=T)
-a7 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST19Q01+name, FUN=mean, rm.na=T)
-colnames(a6) <- colnames(a7) <- c("Job.Status","name","Math","ESCS")
+a6 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST15Q01+ name,FUN=mean,rm.na=T)
+a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
+colnames(a6) <- colnames(a7) <- c("Job.Status","Country","Math","ESCS")
 levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")
 
-a8 <- rbind(cbind(a6, Parent="Mother"), cbind(a7, Parent="Father"))
+a8 <- rbind(cbind(a6,Parent="Mother"),cbind(a7,Parent="Father"))
 #a2 <- melt(a1)
-ind <- which(a8$Job.Status=="Other" & a8$Parent=="Mother")
-a8$name <- factor(a8$name, a8$name[order(a8$Math[ind])])
-qplot(name, Math, col=Job.Status, data = a8) + coord_flip() + facet_wrap(~Parent)
+ind <- which(a8$Job.Status=="Other"& a8$Parent=="Mother")
+a8$Country <- factor(a8$Country,levels(a8$Country)[order(a8$Math[ind])])
+p <- qplot(Country, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
+p
 ```
+
+The Figure ?*?
+shows the average performance of student’s performance based on the parent’s occupational status.
+The countries are sorted according to the average score for students with a "stay-at-home" mother.
+The averages for each parent in each of the work status categories are distinguished by color.
+As you can see in each of the plots below, students’ with a mother who works part-time or full-time tend outperform those students with a stay-at-home mother.
+This pattern holds with only a few exceptions.
+Meanwhile, when the father is employed part-time students tend to have much lower achievement that if the father stayed at home.
+When the father has a full-time jobs, the student’s have the highest average achievement among the possible work status’ for the father.
+The patterns described hold across mathematics, reading, and science.
 

### 2014-09-03 paper/paper.Rmd
albania is weird

@@
 
+Albania appears to be an outlier in many educational trends.
+Many socioeconomic and school factors that play a role in other nations'test scores do not appear to affect Albania's test scores.
+For example, Albania has one of the smallest gender gaps for math and reading scores.
+Additionally number of possessions and truancy rates affected test scores for every nation except Albania.
+
 ```{r clean, echo=FALSE}
@@
 
+Generally students with more possessions are expected to perform better in math, reading and science.
+The number of possessions is a stand in for a student's economic standing.
+The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.
+After aggregating total number of possessions together, the figure ?*?
+shows that increasing number of possessions is correlated with higher test scores in all nations except Albania.
+
 ```{r albania-possessions, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
@@
 ```
+
+A similar aberration is seen with school reported truancy rates.
+Each school reports the level of student truancy: not at all, very little, to some extent, or a lot.
+We group schools together by reported truancy level and aggregate the test scores for students at these schools.
+Generally students at schools with high truancy rates average lower test scores than students at schools with lower truancy rates except for schools in Albania.
 
@@
   id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
+levels(truancy.df$SC22Q01) = c("Not", "Little", "Some", "Lot")
+
 #truancy.df$group = paste0(truancy.df$name, truancy.df$variable)

### 2014-09-03 paper/paper.Rmd
small changes i think, but get a conflict with whats on git

--- 2014-08-29 paper/paper.Rmd
@@
 ```
+
 ### Time spent out of school studying is important, but only up to a point
-### Time spent out of school studying is important, but only up to a point
+
 For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school.

### 2014-09-10 paper/paper.Rmd
updated code for shortages, and for family structure

@@
 library(rworldmap)
-library (grid)    
-library (scales)
+library(grid)    
+library(scales)
 library(doBy)
@@
 
-*** We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!
-
 ```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
@@
 
-### Teachers, computers, and libraries at school make a difference in math scores
+# 1. Teachers, computers, and libraries at school make a difference in math scores
 
 ```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
-school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
-student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)
+school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
+student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
 
-scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
-                                      mread = mean(PV1READ), msci = mean(PV1SCIE))
+scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
+  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
 
 df = scores_by_school %>% left_join(school2012.sub)
-df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
-qplot(SC14Q02, mmath2, data = df2, facets = ~name)
-df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
-qplot(SC14Q10, mmath2, data = df3, facets = ~name)
+df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
+df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
+df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
+df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
+#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
+#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
+colnames(df2)[2] <- "levels"
+df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
+colnames(df3)[2] <- "levels"
+df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
+colnames(df4)[2] <- "levels"
+df.all <- merge(df2, df3)
+df.all <- merge(df.all, df4)
+df.m <- melt(df.all, id=c("name", "levels"))
+colnames(df.m)[4] <- "math"
+qplot(levels, math, data = df.m, facets = ~name, 
+      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
+      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
+      theme(legend.position="bottom") 
+#ggsave("shortages.pdf", width=14, height=14)
+qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
+#ggsave("teacher-shortages.pdf", width=5, height=5)
+qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
+#ggsave("computer-shortages.pdf", width=5, height=5)
+qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
+#ggsave("library-shortages.pdf", width=5, height=5)
 ```
 
-### The gender gap in math is not universal but the reading gap is, in favor of girls
+# 2. The gender gap in math is not universal but the reading gap is, in favor of girls
 
@@
 
-### Starting school at an older age generated lower average math score.
+# 3. Starting school at an older age generated lower average math score.
 
@@
 
-### Time spent out of school studying is important, but only up to a point
+# 4. Time spent out of school studying is important, but only up to a point
 
@@
 
-### Parents matter
-### subsection: Number of parents in the home
+# 5. Parents matter
+### a. Number of parents in the home
 Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household.
@@
 ```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
+athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
 athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
@@
 athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
-athome$name <- reorder(athome$name,athome$PV1MATH,mean)
-pbarhome <- qplot(x = parents, data = athome) 
+athome$parents <- factor(athome$parents, levels=c("Both", "Mother", "Father", "Neither"))
+#athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
+pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
 pbarhome
-
 ```
@@
 ```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
+phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family Structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
 phome
@@
 
-### subsection: Parent's Occupation Status
+### b. Parent's Occupation Status
 One of the variables the survey addressed was the occupation status of each of the parents.
@@
 
-studentLuke$name <- as.character(studentLuke$CNT)
+#studentLuke$name <- as.character(studentLuke$CNT)
 # unique(anti_join(studentLuke.sub, world.polys)[1])
-studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
-studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
-studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
-studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
-studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
-studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
-studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
-studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
-studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
-studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
-studentLuke$name[studentLuke$name=="Macau"] <- "China"
-studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
-studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
-studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
-studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
+#studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
+#studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
+#studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
+#studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
+#studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
+#studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
+#studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
+#studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
+#studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
+#studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
+#studentLuke$name[studentLuke$name=="Macau"] <- "China"
+#studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
+#studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
+#studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
+#studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
 unique(studentLuke$name)
@@
 a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
-colnames(a6) <- colnames(a7) <- c("Job.Status","Country","Math","ESCS")
+colnames(a6) <- colnames(a7) <- c("Job.Status","name","Math","ESCS")
 levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")
@@
 #a2 <- melt(a1)
-ind <- which(a8$Job.Status=="Other"& a8$Parent=="Mother")
-a8$Country <- factor(a8$Country,levels(a8$Country)[order(a8$Math[ind])])
-p <- qplot(Country, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
+ind <- which(a8$Job.Status=="Full-Time"& a8$Parent=="Mother")
+a8$name <- factor(a8$name,levels(a8$name)[order(a8$Math[ind])])
+p <- qplot(name, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
 p
@@
 
-### The weird and wacky republic of Albania
+# 6. Albania is not like anyone else!
 
@@
 
-### More TVs yield higher math scores in the developing world, but lower in the developed countries
+# 7. More TVs yield higher math scores in the developing world, but lower in the developed countries
 

### 2014-09-10 paper/paper.Rmd
more code cleaning

@@
 
-```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
+```{r shortages, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
 school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
@@
 #ggsave("shortages.pdf", width=14, height=14)
+```
+
+```{r shortagesbars, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
 qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
@@
 ```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
-study.luke <- melt(student.luke,id=c("CNT","ST57Q01"))
-studyplot <- ggplot(data = study.luke, aes(ST57Q01,value,colour=variable) )+facet_wrap(~CNT)+stat_smooth(se=F)
+student.luke <- student2012[,c("name","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
+study.luke <- melt(student.luke,id=c("name","ST57Q01"))
+studyplot <- ggplot(data = study.luke, aes(ST57Q01, value, colour=variable)) + facet_wrap(~name) + stat_smooth(se=F)
 studyplot
@@
 
-```{r studyingbar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r studyingbar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
 studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")

### 2014-09-10 paper/paper.Rmd
code almost done, time to move on to the language

@@
 student.luke <- student2012[,c("name","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
-study.luke <- melt(student.luke,id=c("name","ST57Q01"))
-studyplot <- ggplot(data = study.luke, aes(ST57Q01, value, colour=variable)) + facet_wrap(~name) + stat_smooth(se=F)
+#study.luke <- melt(student.luke,id=c("name","ST57Q01"))
+studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH)) + facet_wrap(~name) + stat_smooth(se=F)
 studyplot
@@
 ```{r studyingbar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
-studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
+studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
 studybar
@@
 
-```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r par_athomebar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
 athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]

### 2014-09-10 paper/paper.Rmd
fixed plot sizes, wrote a little bit on shortages

@@
 
-# 1. Teachers, computers, and libraries at school make a difference in math scores
-
-```{r shortages, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
-school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
-student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
-
-scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
-  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
-
-df = scores_by_school %>% left_join(school2012.sub)
-df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
-df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
-df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
-df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
-#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
-#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
-colnames(df2)[2] <- "levels"
-df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
-colnames(df3)[2] <- "levels"
-df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
-colnames(df4)[2] <- "levels"
-df.all <- merge(df2, df3)
-df.all <- merge(df.all, df4)
-df.m <- melt(df.all, id=c("name", "levels"))
-colnames(df.m)[4] <- "math"
-qplot(levels, math, data = df.m, facets = ~name, 
-      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
-      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
-      theme(legend.position="bottom") 
-#ggsave("shortages.pdf", width=14, height=14)
-```
-
-```{r shortagesbars, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
-qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
-#ggsave("teacher-shortages.pdf", width=5, height=5)
-qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
-#ggsave("computer-shortages.pdf", width=5, height=5)
-qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
-#ggsave("library-shortages.pdf", width=5, height=5)
-```
-
-# 2. The gender gap in math is not universal but the reading gap is, in favor of girls
+# 1. The gender gap in math is not universal but the reading gap is, in favor of girls
 
@@
 #  new_theme_empty + theme(legend.position="none")
+```
+
+# 2. Teachers, computers, and libraries at school make a difference in math scores
+
+The school table has information about shortages on items viewed as being important for good instruction.
+The plot below examines at three of these, lack of teachers, lack of computers and shortage of library materials, in relation to math, reading and science scores from the student table, across countries.
+For most countries scores decrease on average the greater the shortage.
+
+```{r shortages, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
+school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
+student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
+
+scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
+  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
+
+df = scores_by_school %>% left_join(school2012.sub)
+df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
+df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
+df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
+df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
+#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
+#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
+colnames(df2)[2] <- "levels"
+df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
+colnames(df3)[2] <- "levels"
+df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
+colnames(df4)[2] <- "levels"
+df.all <- merge(df2, df3)
+df.all <- merge(df.all, df4)
+df.m <- melt(df.all, id=c("name", "levels"))
+colnames(df.m)[4] <- "math"
+qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
+#ggsave("teacher-shortages.pdf", width=5, height=5)
+qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
+#ggsave("computer-shortages.pdf", width=5, height=5)
+qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
+#ggsave("library-shortages.pdf", width=5, height=5)
+#ggsave("shortages.pdf", width=14, height=14)
+```
+
+```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
+qplot(levels, math, data = df.m, facets = ~name, 
+      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
+      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
+      theme(legend.position="bottom") 
 ```

### 2014-09-10 paper/paper.Rmd
small tweaks - but think the countries are messed up for the gender plots

@@
 
-# 1. The gender gap in math is not universal but the reading gap is, in favor of girls
+# 1. The gender gap in math is not universal but the reading gap is, in favor of girls.
 
@@
 There is far more individual variability than the overall average differences.
+
+*** Looks like I messed up the colors for countries - need to check this again
 
@@
 
-# 3. Starting school at an older age generated lower average math score.
+# 3. Starting school at an older age generates lower average math scores.
 
@@
 
-# 4. Time spent out of school studying is important, but only up to a point
+# 4. Time spent out of school studying is important, but only up to a point.
 
@@
 
-# 5. Parents matter
+# 5. Parents matter!
 ### a. Number of parents in the home
@@
 
-# 7. More TVs yield higher math scores in the developing world, but lower in the developed countries
+# 7. More TVs yield higher math scores in the developing world, but lower in the developed countries.
 
@@
 
-### Play with the data yourself
+### Play with the data yourself.
 

### 2014-09-10 paper/paper.Rmd
trying to fix gender maps

@@
 student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
-colnames(student2012.sub)[1] <- "name"
-student2012.sub$PV1MATH <- as.numeric(student2012.sub$PV1MATH)
-student2012.sub$PV1READ <- as.numeric(student2012.sub$PV1READ)
-student2012.sub$PV1SCIE <- as.numeric(student2012.sub$PV1SCIE)
-student2012.sub$SENWGT_STU <- as.numeric(student2012.sub$SENWGT_STU)
 student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
@@
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
+```
+
+```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
+student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
+ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
+  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
+  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  new_theme_empty + theme(legend.position="none")
 #student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
@@
 #
+```
+
+```{r genderead, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
 student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
@@
 
-```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
-student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
-ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
-  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
-  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  new_theme_empty + theme(legend.position="none")
+```{r gendermaps2, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
 #ggsave("gendermathmap.pdf", width=12, height=7)

### 2014-09-10 paper/paper.Rmd
not sure if it committed

@@
 student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
+student2012.sub.map$msig <- factor(student2012.sub.map$msig)
 ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 

### 2014-09-11 paper/paper.Rmd
A few small fixes here and there

@@
 
-Reducing inequities in education outcomes is of great interest to politicians.
+Reducing inequities in education outcomes is of great interest to politicians and the public, alike.
 Dissecting the causes of inequality is a much tougher challenge with no panacea.
@@
 This represents over 80% of the global economy.
-The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add a collective US$200 trillion to the world's collective GDP.
+The OECD estimates that if all students could reach a level-2 proficiency in mathematics -- a level-2 proficiency means that a student can only handle the ``simplest and most obvious tasks" -- it would add US$200 trillion to the world's collective GDP.
 
@@
 We learned a lot from this data, and won one of the prizes.
-This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.
+This article is a description of the seven main things we learned about 15-year olds, their math, science and reading skills, and some of the factors that affect these skills.
 
@@
 Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls.
-There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score.
+There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score.
 There is far more individual variability than the overall average differences.
@@
 
-The school table has information about shortages on items viewed as being important for good instruction.
-The plot below examines at three of these, lack of teachers, lack of computers and shortage of library materials, in relation to math, reading and science scores from the student table, across countries.
-For most countries scores decrease on average the greater the shortage.
+Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction.
+The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials.
+These factors are compared with math, reading and science scores from the student table, across countries.
+For most countries, scores decrease on average the greater the shortage.
 
@@
 
-For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school.
+For each country we have provided loess-smoothed curves of the relationship between mathematics score and the time spent studying outside of school.
 The most prominent feature that we observe is the decreasing benefit of studying.
@@
 ### a. Number of parents in the home
-Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household.
+Looking at the number of parents in a household shows that students in two-parent households are on average performing better than students not in a two-parent households.
 Students with just a mother at home do tend to perform better than students with only a father at home.
@@
 The distribution of the number of parents who selected each of the categories can be seen in the bar chart.
-The chart shows that father’s primarily have full-time jobs.
+The chart shows that fathers primarily have full-time jobs.
 However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms.
-With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.
+With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in mathematics, reading, and science.
 
@@
 The Figure ?*?
-shows the average performance of student’s performance based on the parent’s occupational status.
+shows the average performance of students' performance based on the parents' occupational statuses.
 The countries are sorted according to the average score for students with a "stay-at-home" mother.
 The averages for each parent in each of the work status categories are distinguished by color.
-As you can see in each of the plots below, students’ with a mother who works part-time or full-time tend outperform those students with a stay-at-home mother.
+In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers.
 This pattern holds with only a few exceptions.
-Meanwhile, when the father is employed part-time students tend to have much lower achievement that if the father stayed at home.
-When the father has a full-time jobs, the student’s have the highest average achievement among the possible work status’ for the father.
+Meanwhile, when the father is employed part-time, students tend to have much lower achievement that if the father stayed at home.
+Those students whose fathers have full-time jobs have the highest average achievement among the possible work statuses for the father.
 The patterns described hold across mathematics, reading, and science.
@@
 Albania appears to be an outlier in many educational trends.
-Many socioeconomic and school factors that play a role in other nations'test scores do not appear to affect Albania's test scores.
+Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.
 For example, Albania has one of the smallest gender gaps for math and reading scores.
@@
 
-Generally students with more possessions are expected to perform better in math, reading and science.
+Generally, students with more possessions are expected to perform better in math, reading and science.
 The number of possessions is a stand in for a student's economic standing.
@@
 The PISA dataset contains hundreds of factors, and this article can describe just a few findings.
-The reader is welcome to explore the data themselves using an <a href="http://bit.ly/pisa2012_explorer"> interactive web tool</a>, made with <a href="http://www.rstudio.com">RStudio</a> and Shiny.
-Shiny is a open-source tool, developed by RStudio, that allows an investigator to create a web-based appication powered by R.
+The reader is welcome to explore the data themselves using an  [interactive web tool](http://bit.ly/pisa2012_explorer), made with RStudio's and [Shiny](http://shiny.rstudio.com).
+Shiny is a open-source tool that allows an investigator to create web-based appications powered by R.
 
@@
 
-Shiny, by RStudio
-
-plyr, dplyr, and ggplot2, by Hadley Wickham
+shiny, plyr, dplyr, ggplot2

### 2014-09-11 paper/paper.Rmd
fixed gender map problem

@@
 ```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+library(plyr)
 library(dplyr)
@@
 
-```{r mapdata, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+```{r mapdata, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=FALSE}
 # Produce the maps
@@
 world <- getMap(resolution = "low")
-library(plyr)
+#library(plyr)
 world.polys <- extractPolygons(world)
@@
 
-*** Looks like I messed up the colors for countries - need to check this again
-
-```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
+```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
 student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
@@
   scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
+  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5), limits=c(-30, 30)) + 
   scale_size("Prop male") +
@@
 
-```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
+```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
+student2012.sub.summary.gap$name <- as.character(student2012.sub.summary.gap$name)
 student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
@@
 
-```{r genderead, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
+```{r genderead, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
 student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
@@
 
-```{r gendermaps2, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
+```{r gendermaps2, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
 #ggsave("gendermathmap.pdf", width=12, height=7)
@@
 
-The Figure ?*?
-shows the average performance of students' performance based on the parents' occupational statuses.
+The plot below shows the average performance of students' performance based on the parents' occupational statuses.
 The countries are sorted according to the average score for students with a "stay-at-home" mother.
@@
 Generally, students with more possessions are expected to perform better in math, reading and science.
-The number of possessions is a stand in for a student's economic standing.
+The number of possessions is a surrogate for a student's economic standing.
 The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.

### 2014-09-11 paper/paper.Rmd
fixed gender reading code, next job to work on country and categorical variable labels so that they fit into the facet label windows better

@@
 #ggsave("gendermathmap.pdf", width=12, height=7)
+student2012.sub.summary.gap$name <- as.character(student2012.sub.summary.gap$name)
+student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
+student2012.sub.map$rsig <- factor(student2012.sub.map$rsig)
 ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 

### 2014-09-11 paper/paper.Rmd
moved another section with maps up
labels for truancy adjusted

@@
 
+shorten_countrynames = function(df) {
+df$name = as.character(df$name)
+df$name[df$name=="Hong Kong S.A.R."] <- "Hong Kong"
+df$name[df$name=="Republic of Serbia"] <- "Serbia"
+df$name[df$name=="United Arab Emirates"] <- "UAE"
+df$name[df$name=="United States of America"] <- "USA"
+df$name[df$name=="United Kingdom"] <- "UK"
+
+df
+}
+
 # Now add data
@@
 
-# 2. Teachers, computers, and libraries at school make a difference in math scores
-
-Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction.
-The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials.
-These factors are compared with math, reading and science scores from the student table, across countries.
-For most countries, scores decrease on average the greater the shortage.
-
-```{r shortages, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
-school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
-student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
-
-scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
-  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
-
-df = scores_by_school %>% left_join(school2012.sub)
-df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
-df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
-df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
-df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
-#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
-#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
-colnames(df2)[2] <- "levels"
-df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
-colnames(df3)[2] <- "levels"
-df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
-colnames(df4)[2] <- "levels"
-df.all <- merge(df2, df3)
-df.all <- merge(df.all, df4)
-df.m <- melt(df.all, id=c("name", "levels"))
-colnames(df.m)[4] <- "math"
-qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
-#ggsave("teacher-shortages.pdf", width=5, height=5)
-qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
-#ggsave("computer-shortages.pdf", width=5, height=5)
-qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
-#ggsave("library-shortages.pdf", width=5, height=5)
-#ggsave("shortages.pdf", width=14, height=14)
-```
-
-```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
-qplot(levels, math, data = df.m, facets = ~name, 
-      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
-      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
-      theme(legend.position="bottom") 
-```
-
-# 3. Starting school at an older age generates lower average math scores.
+# 2. Starting school at an older age generates lower average math scores.
 
@@
 ```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
-# Extract map polygons for modern world
-world <- getMap(resolution = "low")
-library(plyr)
-world.polys <- extractPolygons(world)
-detach("package:plyr")
-
 student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
@@
 #ggsave("agestartmap.pdf", width=12, height=7)
+```
+
+# 3. Teachers, computers, and libraries at school make a difference in math scores
+
+Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction.
+The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials.
+These factors are compared with math, reading and science scores from the student table, across countries.
+For most countries, scores decrease on average the greater the shortage.
+
+```{r shortencountrynames, warning=FALSE, message=FALSE, echo=FALSE, results='hide'}
+#school2012$name <- as.character(school2012$name)
+#student2012$name <- as.character(student2012$name)
+school2012 = shorten_countrynames(school2012)
+student2012 = shorten_countrynames(student2012)
+```
+
+```{r shortages, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
+school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
+student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
+
+scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
+  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
+
+df = scores_by_school %>% left_join(school2012.sub)
+df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
+df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
+df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
+df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
+#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
+#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
+colnames(df2)[2] <- "levels"
+df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
+colnames(df3)[2] <- "levels"
+df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
+colnames(df4)[2] <- "levels"
+df.all <- merge(df2, df3)
+df.all <- merge(df.all, df4)
+df.m <- melt(df.all, id=c("name", "levels"))
+colnames(df.m)[4] <- "math"
+qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
+#ggsave("teacher-shortages.pdf", width=5, height=5)
+qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
+#ggsave("computer-shortages.pdf", width=5, height=5)
+qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
+#ggsave("library-shortages.pdf", width=5, height=5)
+#ggsave("shortages.pdf", width=14, height=14)
+```
+
+```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
+qplot(levels, math, data = df.m, facets = ~name, 
+      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
+      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
+      theme(legend.position="bottom") 
 ```
@@
 
+The plot below shows the average performance of students' performance based on the parents' occupational statuses.
+The countries are sorted according to the average score for students with a "stay-at-home" mother.
+The averages for each parent in each of the work status categories are distinguished by color.
+In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers.
+This pattern holds with only a few exceptions.
+Meanwhile, when the father is employed part-time, students tend to have much lower achievement that if the father stayed at home.
+Those students whose fathers have full-time jobs have the highest average achievement among the possible work statuses for the father.
+The patterns described hold across mathematics, reading, and science.
+
 ```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
@@
 #studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
-unique(studentLuke$name)
+#unique(studentLuke$name)
 studentLuke$name <- factor(studentLuke$name)
@@
 
-The plot below shows the average performance of students' performance based on the parents' occupational statuses.
-The countries are sorted according to the average score for students with a "stay-at-home" mother.
-The averages for each parent in each of the work status categories are distinguished by color.
-In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers.
-This pattern holds with only a few exceptions.
-Meanwhile, when the father is employed part-time, students tend to have much lower achievement that if the father stayed at home.
-Those students whose fathers have full-time jobs have the highest average achievement among the possible work statuses for the father.
-The patterns described hold across mathematics, reading, and science.
-
 # 6. Albania is not like anyone else!
@@
 
-```{r clean, echo=FALSE}
+```{r clean, echo=FALSE, message = FALSE, warning = FALSE}
 #student2012$CNT = student2012$name
@@
 #qplot(x = SC22Q01, y = value, color = variable, facets = ~name, data = truancy.df, group = group, geom = "line") + xlab("Reported Truancy") + ylab("Test Scores") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
-qplot(x = SC22Q01, y = value, facets = ~name, data = truancy.df, geom = "line", group=name) + xlab("Reported Truancy") + ylab("Math Scores") + theme_bw()
+qplot(x = SC22Q01, y = value, facets = ~name, data = truancy.df, group=name, geom="smooth", se=F) + 
+  scale_x_discrete("Reported Truancy, 0=None, 1=Little, 2=Some, 3=Lot", labels=c("Not"="0", "Little"="1", "Some"="2", "Lot"="3")) + 
+  ylab("Math Scores") + theme_bw()
 ```

### 2014-09-11 paper/paper.Rmd
resized pisa image to fit better, tweaking code, language

@@
 #study.luke <- melt(student.luke,id=c("name","ST57Q01"))
-studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH)) + facet_wrap(~name) + stat_smooth(se=F)
+studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH), xlab="Time spent", ylab="Math scores") + facet_wrap(~name) + stat_smooth(se=F)
 studyplot
@@
 
-```{r studytime, warning = FALSE, message=FALSE, echo=FALSE}
-.width = 1
-
-data_small <- 
-  student2012 %>%
-  dplyr::select(OECD, name, OUTHOURS) %>%
-  filter(!is.na(OUTHOURS))
-
-ggplot(mapping = aes(x = OUTHOURS)) + 
-  scale_x_continuous(name = "out of school study time (hours)", limits = c(0,60)) +
-  scale_y_continuous(name = "density") +
-  geom_histogram(
-    mapping = aes(y = ..density.., fill = OECD), 
-    binwidth = .width,
-    data = data_small %>% filter(OECD == "Albania")
-  ) + 
-  scale_fill_manual(
-    values = "black",
-    breaks = "Albania",
-    guide = guide_legend(title = "")
-  ) + 
-  geom_line(
-    mapping = aes(group = name, color = OECD), 
-    stat = "density", 
-    alpha = 0.3,
-    data = data_small %>% filter(OECD != "Albania")
-  ) +
-  scale_color_discrete(
-    guide = guide_legend(
-      title = "Other Countries",
-      override.aes = list(alpha = 1)
-    )
-
-  ) +
-  scale_colour_manual(values = c("#f4a582", "#92c5de")) + guides(colour = guide_legend(override.aes = list(alpha = 1)))
-```
-
 # 7. More TVs yield higher math scores in the developing world, but lower in the developed countries.
@@
 
-shiny, plyr, dplyr, ggplot2
+The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy.

### 2014-09-11 paper/paper.Rmd
small edits

@@
 However, those students who continue to study for long hours see diminishing returns for their efforts.
-It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster.
-However, the relationship shows those spending
+It is possible that there may be a selection issue occuring, where more-talented students spend less time studying as they complete their homework faster.
+However, the relationship shows those spending === NEED THE REST OF THE SENTENCE ===
 
@@
 The PISA dataset contains hundreds of factors, and this article can describe just a few findings.
-The reader is welcome to explore the data themselves using an  [interactive web tool](http://bit.ly/pisa2012_explorer), made with RStudio's and [Shiny](http://shiny.rstudio.com).
+The reader is welcome to explore the data themselves using an  [interactive web tool](http://bit.ly/pisa2012_explorer), made with RStudio's [Shiny](http://shiny.rstudio.com).
 Shiny is a open-source tool that allows an investigator to create web-based appications powered by R.

### 2014-09-11 paper/paper.Rmd
out of school study time

--- 2014-09-11 paper/paper.Rmd
@@
 
-For each country we have provided loess-smoothed curves of the relationship between mathematics score and the time spent studying outside of school.
+For each country we have provided loess-smoothed curves of the relationship between the mathematics score and the time spent studying outside of school.
 The most prominent feature that we observe is the decreasing benefit of studying.
@@
 However, those students who continue to study for long hours see diminishing returns for their efforts.
+This pattern is consistent across most countries with a few notable exceptions.
+Most of the Asian countries still have gains for longer hours studied.
+
+The vast majority of students spend less than 10 hours studying per week.
+It is among the students that spend 5 or less hours per week that could see improvement in their scores if they put in more time.
+The pattern tends to level off at 10 hours.
+
 It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster.
-However, the relationship shows those spending
+Nonetheless the pattern suggest decreasing benefits to increased out of school study time after 10 hour.
 

### 2014-09-11 paper/paper.Rmd
intro and section 1

@@
 
-Reducing inequities in education outcomes is of great interest to politicians and the public, alike.
+Reducing inequities in educational outcomes is of great interest to politicians and the public, alike.
 Dissecting the causes of inequality is a much tougher challenge with no panacea.
 Nonetheless, tackling inequality in educational outcomes is essential from both moral and economic viewpoints.
-The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating additional emphasis on one of mathematics, reading, and science.
-In 2012, the additional emphasis was on mathematics.
+The Programme for International Student Assessment (PISA) is a triennial survey conducted by the Organization for Economic Cooperation and Development (OECD) with a rotating emphasis on one of mathematics, reading, or science.
+In 2012, the emphasis was on mathematics.
 All 34 member countries of the OECD and 31 partnering countries and economies participated in the survey.
@@
 
-The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p.
-3]{PISA2014a}).
+The OECD PISA Results in Focus report describes the survey as ``the world's global metric for quality, equity and efficiency in school education".
 The goal of the PISA survey is to assess the workforce readiness of 15-year old students.
-Nearly 500,000 students were tested across the 65 countries and economies.
-Students are examined on how well they can apply the knowledge they learned in school to applications outside of school.
-The scores reported range between 0-1000.
-Additional information about the students, parents, and schools is also collected.
-The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school.
-Schools principals responded to a questionnaire covering the school system and learning experiences for students.
+Nearly 500,000 students were tested across 65 countries and economies.
+Students were examined on how well they can apply the knowledge they learned in school to applications outside of school.
+The  reported scores range between 0-1000.
+Information about the students, parents, and schools is also collected.
+The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affect their performance in school.
+School principals responded to a questionnaire covering their school system and learning experiences for their students.
 In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.
@@
 We learned a lot from this data, and won one of the prizes.
-This article is a description of the seven main things we learned about 15-year olds, their math, science and reading skills, and some of the factors that affect these skills.
+This article is a description of the seven main things we learned about 15-year olds, their math, science, and reading skills, and some of the factors that affect these skills.
 
@@
 These averages are differenced to compute the "gap" between boys and girls.
-T-statistics were calculated to determine the statistical significance of the difference.
+Unadjusted t-statistics were calculated to determine the statistical significance of the difference.
 The results are plotted in the graphs below.
-Each dot corresponds to a difference between girls and boys scores, for each country.
-Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference).
+Each dot corresponds to a difference between girls' and boys' scores, for each country.
+Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no significant difference).
 The size of the dot indicates the proportion of boys in the sample size.
@@
 It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.
-Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.
+Five countries have gender gaps in favor of girls, and even more surprising, these are middle eastern or Muslim countries.
 
-Conversely, though, when we look at reading scores the story is entirely different.
-Universally, girls score significantly better than boys.
-Again, the size of the gap is small, between 10 and 80 points.
+Conversely, though, looking at reading scores, the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.
 
-Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls.
-There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score.
-There is far more individual variability than the overall average differences.
+Although not shown here, it also should be noted that this is the difference between the average scores for boys and girls.
+There is a slightly different story for each country if we look at the top math score for each gender only.
+We would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score.
+There is far more individual variability than the overall average differences: who you are is more important than your demographic.
 
@@
 The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy.
+
+### Reference
+
+OECD (2014) "Pisa 2012 results in focus: What 15-year-olds know and what they can do with what they know." Technical Report, Organization for Economic Development and Cooperation.

### 2014-09-11 paper/paper.Rmd
hyphen
section 3

@@
 We would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score.
-There is far more individual variability than the overall average differences: who you are is more important than your demographic.
+There is far more individual variability than the overall average differences - who you are is more important than your demographic.
 
@@
 
-For each country we have computed the average math test score by age that the children started school.
-In additon, for each country we computed the median age that children in the study started school.
+For each country, we have computed the average math score by age that the children started school.
+In addition, for each country we computed the median age that children in the study started school.
 There are differences in typical school start age between countries.
@@
 Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5.
-For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7.
-Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.
+For many other countries the median start age is 6, and for northern Europe, much of Asia, and Brazil median start age is as late as 7.
+Some countries appear to have rigid starting ages, e.g.
+in Japan all children start at age 6.
 
@@
 This is almost uniformly the pattern across all of the countries in the study.
-For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than another other starting age group.
-School start age matters, that children need to start school in the 4-6 age range to expect better math scores, on average.
+For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than any other age group.
+School start age matters.
+Children need to start school in the 4-6 age range to expect better math scores, on average.
 
@@
 
-# 3. Teachers, computers, and libraries at school make a difference in math scores
+# 3. Shortages in teachers, computers, and libraries at school make a difference in math scores in a few countries
 
 Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction.
-The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials.
-These factors are compared with math, reading and science scores from the student table, across countries.
-For most countries, scores decrease on average the greater the shortage.
+The plot below examines three of these: lack of teachers, lack of computers, and shortage of library materials.
+The barcharts show the overall counts of shortages in schools.
+Teacher shortage is less comman than material shortages.
+
+These factors are compared with math scores from the student table, across countries.
+In most countries, scores don't change mcuh as shortages get worse.
+Scores decrease by 10-20 points onaverage in countries like Australia, China, Uruguay, Turkey, UAE, USA.
+Teacher shortages appear to have impact in a few countries like Greece, Slovakia, and Serbia.
 
@@
 ```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
-qplot(levels, math, data = df.m, facets = ~name, 
+df$SC14Q02 <- as.character(df$SC14Q02)
+df$SC14Q07 <- as.character(df$SC14Q07)
+df$SC14Q10 <- as.character(df$SC14Q10)
+df$SC14Q02[df$SC14Q02=="A lot"] <- "To some extent"
+df$SC14Q07[df$SC14Q07=="A lot"] <- "To some extent"
+df$SC14Q10[df$SC14Q10=="A lot"] <- "To some extent"
+df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2))
+df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2))
+df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2))
+df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
+#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
+#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
+#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+#      ylab="Math Score", geom=c("point","line"), group=name)
+df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
+colnames(df2)[2] <- "levels"
+df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
+colnames(df3)[2] <- "levels"
+df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
+colnames(df4)[2] <- "levels"
+df.all <- merge(df2, df3)
+df.all <- merge(df.all, df4)
+df.m <- melt(df.all, id=c("name", "levels"))
+colnames(df.m)[4] <- "math"
+dqplot(levels, math, data = df.m, facets = ~name, 
       xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 

### 2014-09-11 paper/paper.Rmd
fixed code error, hopefully

@@
 colnames(df.m)[4] <- "math"
-dqplot(levels, math, data = df.m, facets = ~name, 
-      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
+qplot(levels, math, data = df.m, facets = ~name, 
+      xlab="Shortages (0=None, 1=Very little, 2=To some extent, or A lot)", 
       ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
@@
 
-For each country we have provided loess-smoothed curves of the relationship between the mathematics score and the time spent studying outside of school.
-The most prominent feature that we observe is the decreasing benefit of studying.
-For every country there is evidence of a substantial positive relationship for the students who spend less time studying.
-However, those students who continue to study for long hours see diminishing returns for their efforts.
-This pattern is consistent across most countries with a few notable exceptions.
-Most of the Asian countries still have gains for longer hours studied.
+For each country we have provided loess-smoothed curves of the relationship between the math score and the time spent studying outside of school.
+Math scores increase on average the more time spent studying, up to about 10 hours.
+Beyond that scores stay flat, with the exception being several Asian countries: Vietnam, China, South Korea, Taiwan.
+For China, the gain tails off around 20 hours of study.
 

### 2014-09-11 paper/paper.Rmd
out of school

@@
 
-The vast majority of students spend less than 10 hours studying per week.
-It is among the students that spend 5 or less hours per week that could see improvement in their scores if they put in more time.
-The pattern tends to level off at 10 hours.
-
-It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster.
-Nonetheless the pattern suggest decreasing benefits to increased out of school study time after 10 hour.
+The vast majority of students spend less than 10 hours studying per week. The pattern  suggests decreasing benefits to increased out of school study time after 10 hours.
 

### 2014-09-11 paper/paper.Rmd
parents

@@
 ### a. Number of parents in the home
-Looking at the number of parents in a household shows that students in two-parent households are on average performing better than students not in a two-parent households.
-Students with just a mother at home do tend to perform better than students with only a father at home.
-Students with no parents at home are performing the worst, on average.
+The role of parents in educational achievement is the subject of much research.
+Looking at the number of parents in a household shows that students in a two-parent household are, on average, performing better in mathematics than students not in a two-parent household.
+The majority of students live in a two-parent household.
+There are substantially more single parent household with a mother than a father.
+This may be a result of custody laws that tend to assign children to their mother's care.
+Regardless, the students with just a mother at home tend to do better than those household with just a father.
+In many cases, mostly Asian countries, the single mother household achieve comparibly to two-parent households.
+The students in a household with no parents tend to perform the worst.
+This pattern is consistent across all countries.
 

### 2014-09-11 paper/paper.Rmd
occupation

--- 2014-09-11 paper/paper.Rmd
@@
 ### a. Number of parents in the home
-Looking at the number of parents in a household shows that students in two-parent households are on average performing better than students not in a two-parent households.
-Students with just a mother at home do tend to perform better than students with only a father at home.
-Students with no parents at home are performing the worst, on average.
+The role of parents in educational achievement is the subject of much research.
+Looking at the number of parents in a household shows that students in two-parent households are, on average, performing better in math than students not in a two-parent households.
+The majority of students live in two-parent households.
+Of the single-parent households, a single mother is more common than a single father.
+Scores of students with just a mother at home tend to be higher than those with just a father at home.
+In many countries, mostly Asian, students from single-mother households achieve comparably to those from two-parent households.
+Students in households with no parents tend to perform the worst.
+This pattern is consistent across all countries.
 
@@
 home duties, retired)”, and no response.
-The distribution of the number of parents who selected each of the categories can be seen in the bar chart.
+The frequencies for these categories can be seen in the bar chart.
 The chart shows that fathers primarily have full-time jobs.
 However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms.
-With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in mathematics, reading, and science.
+With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in math.
 
-The plot below shows the average performance of students' performance based on the parents' occupational statuses.
-The countries are sorted according to the average score for students with a "stay-at-home" mother.
-The averages for each parent in each of the work status categories are distinguished by color.
+The plot below shows the average student math score by the parents' occupational status.
+The countries are sorted according to the average score for students with a "full-time" mother.
+Occupational status are distinguished by color.
 In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers.
@@
 a8$name <- factor(a8$name,levels(a8$name)[order(a8$Math[ind])])
-p <- qplot(name, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
+p <- qplot(name, Math, col=Job.Status, data = a8, xlab="") + coord_flip() + facet_wrap(~Parent) + theme(legend.position="bottom")
 p

### 2014-09-11 paper/paper.Rmd
occupation

@@
 There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g.
-home duties, retired)”, and no response.
+stay-at-home, retired)”, and no response.
 The frequencies for these categories can be seen in the bar chart.
 The chart shows that fathers primarily have full-time jobs.
-However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms.
+However, there are large numbers of mothers who work full-time, part-time, or are categorized as "Other".
 With women making up a larger portion of the work-force, it is interesting to look at how the parents' occupations relate to students' performance in math.
@@
 The plot below shows the average student math score by the parents' occupational status.
-The countries are sorted according to the average score for students with a "full-time" mother.
-Occupational status are distinguished by color.
-In each of the plots below, students with mothers who work part-time or full-time tend outperform those students with stay-at-home mothers.
-This pattern holds with only a few exceptions.
-Meanwhile, when the father is employed part-time, students tend to have much lower achievement that if the father stayed at home.
-Those students whose fathers have full-time jobs have the highest average achievement among the possible work statuses for the father.
-The patterns described hold across mathematics, reading, and science.
+The countries are sorted according to the average score for students with a mother who is "full-time" in the work force.
+Occupational status is distinguished by color.
+Students with mothers who work part-time or full-time tend outperform those students with mothers labelled as "Other".
+This pattern holds with only a few exceptions, Switzerland, Netherlands, USA, Japan.
+Meanwhile, when the father is employed part-time, students tend to have much lower scores that if the father labelled as "Other".
+Those students whose fathers have full-time jobs have the highest average scores among the possible work statuses for the father.
+The patterns described also hold across reading and science.
 

### 2014-09-11 paper/paper.Rmd
done!

@@
 Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.
-For example, Albania has one of the smallest gender gaps for math and reading scores.
-Additionally number of possessions and truancy rates affected test scores for every nation except Albania.
+For example, Albania has one of the smallest gender gaps for math scores.
+Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.
 
@@
 
-Generally, students with more possessions are expected to perform better in math, reading and science.
+Generally, students with more possessions score higher on average.
 The number of possessions is a surrogate for a student's economic standing.
 The survey records whether students have desks, computers, internet, books, a place to study and dozens of other possessions.
-After aggregating total number of possessions together, the figure ?*?
-shows that increasing number of possessions is correlated with higher test scores in all nations except Albania.
+After aggregating these number of possessions together, to get a total, we can see that scores increase on average as possessions increase in all nations, except Albania.
 
@@
 We group schools together by reported truancy level and aggregate the test scores for students at these schools.
-Generally students at schools with high truancy rates average lower test scores than students at schools with lower truancy rates except for schools in Albania.
+Generally students at schools with high truancy rates average lower test scores than students at schools with lower truancy rates, except for schools in Albania.
 
@@
 Possible values were: "zero", "one", "two", or "three or more".
-Because there were so (relatively) few observations of "zero", for this invesitgation, the "zero" and "one" categories were combined to form a "one or less" category.
+Because there were so (relatively) few observations of "zero", for this investigation, the "zero" and "one" categories were combined to form a "one or less" category.
 
-A linear model is fitted the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
+A linear model was fitted for the students of each country, using the math score as the dependent variable and the number of televisions as the independent variable (as an ordered factor).
 The direction and significance of the correlation was determined for each country.
-Significance is indicated by the p-value for a given regression being less than the conventional 0.05.
+Significance is indicated by the p-value for a given regression being less than 0.05.
 
@@
 
-The figure below illustrates the relationship, along with another factor, student has a place to study.
+The figure below illustrates the relationship, along with another factor, the student has a place to study.
 It is a complex diagram, but of interest is the middle column of plots - the boxplots overlaid on other information show the math scores of students relative to the number of TVs in the household, separately plotted for countries falling into one of three categories, negative, zero or  positive correlation.
@@
 
+It is interesing that the more TVs in the household in developed countries leads to lower math scores on average.
 Although results shown here are for math only, a similar set of relationships is found using the reading and science scores.
-
 It seems that television is a great equalizer.
@@
 
+# Summary
+
+Overall, we have seen that there are many factors that affect the average math scores, but the major source of variation in the data is from individual to individual.
+You as a person are more than your demographic -  you count, your individual efforts matter.
+
 ### Acknowledgements

### 2014-09-11 paper/paper.Rmd
fixed code error, and axis labels

@@
 student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
-qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Scores", 
+qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math score", 
       geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
@@
   geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
-  scale_fill_discrete("Median Start Age") +
+  scale_fill_discrete("Median start age") +
   new_theme_empty + theme(legend.position="bottom")
@@
 df$SC14Q10 <- as.character(df$SC14Q10)
-df$SC14Q02[df$SC14Q02=="A lot"] <- "To some extent"
-df$SC14Q07[df$SC14Q07=="A lot"] <- "To some extent"
-df$SC14Q10[df$SC14Q10=="A lot"] <- "To some extent"
+df$SC14Q02[df$SC14Q02==3] <- 2
+df$SC14Q07[df$SC14Q07==3] <- 2
+df$SC14Q10[df$SC14Q10==3] <- 2
 df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2))
@@
 qplot(levels, math, data = df.m, facets = ~name, 
-      xlab="Shortages (0=None, 1=Very little, 2=To some extent, or A lot)", 
-      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
+      xlab="Shortages (0=None, 1=Very little, 2=To some extent, or a lot)", 
+      ylab="Math score", geom=c("point","line"), group=variable, colour=variable) +
       scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
@@
 #study.luke <- melt(student.luke,id=c("name","ST57Q01"))
-studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH), xlab="Time spent", ylab="Math scores") + facet_wrap(~name) + stat_smooth(se=F)
+studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH)) + xlab("Hours spent out of school studying per week") + ylab("Math score") + facet_wrap(~name) + stat_smooth(se=F)
 studyplot
@@
 ```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family Structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
+phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family Structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
 phome
@@
 a8$name <- factor(a8$name,levels(a8$name)[order(a8$Math[ind])])
-p <- qplot(name, Math, col=Job.Status, data = a8, xlab="") + coord_flip() + facet_wrap(~Parent) + theme(legend.position="bottom")
+p <- qplot(name, Math, col=Job.Status, data = a8, xlab="Math score") + coord_flip() + facet_wrap(~Parent) + theme(legend.position="bottom")
 p
@@
 #qplot(x = numposs, y = value, data = df, facets = ~OECD, geom = "line", color = variable, group = group) + ylab("test scores") + xlab("number of possessions") + scale_colour_discrete(name = "Subject", labels=c("Math", "Science", "Reading"))
-qplot(x = numposs, y = value, data = df, facets = ~name, geom = "line") + ylab("Math Scores") + xlab("Number of Possessions") + theme_bw()
+qplot(x = numposs, y = value, data = df, facets = ~name, geom = "line") + ylab("Math score") + xlab("Number of possessions") + theme_bw()
 ```
@@
 qplot(x = SC22Q01, y = value, facets = ~name, data = truancy.df, group=name, geom="smooth", se=F) + 
-  scale_x_discrete("Reported Truancy, 0=None, 1=Little, 2=Some, 3=Lot", labels=c("Not"="0", "Little"="1", "Some"="2", "Lot"="3")) + 
-  ylab("Math Scores") + theme_bw()
+  scale_x_discrete("Reported truancy, 0=None, 1=Little, 2=Some, 3=Lot", labels=c("Not"="0", "Little"="1", "Some"="2", "Lot"="3")) + 
+  ylab("Math score") + theme_bw()
 ```

### 2014-09-11 paper/paper.Rmd
Final version, and word copy

@@
 ```{r studyingbar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
-studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
+studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=5) + xlab("Hours spent studying")
 studybar
@@
 
-```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
-phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family Structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
+```{r par_athome, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
 phome
@@
 a8$name <- factor(a8$name,levels(a8$name)[order(a8$Math[ind])])
-p <- qplot(name, Math, col=Job.Status, data = a8, xlab="Math score") + coord_flip() + facet_wrap(~Parent) + theme(legend.position="bottom")
+p <- qplot(name, Math, col=Job.Status, data = a8, ylab="Math score", xlab="") + coord_flip() + facet_wrap(~Parent) + theme(legend.position="bottom")
 p

### 2014-09-18 dicook/school_start_age.Rmd
slides of material in paper

@@
 ```
-```
+
 ```{r age, fig.width=8, fig.height=12, warning=FALSE, message=FALSE}

### 2015-05-08 dicook/gender.Rmd
added the review that we need to address

@@
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
+qplot(name, mathgap, data=student2012.sub.summary.gap, size=I(5), color=msig) + 
+  xlab("") +
+  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  scale_y_continuous("Math Score Gap", limits=c(-80, 30), breaks=seq(-80, 30, 5), labels=c(seq(80, 0, -5), seq(5, 30, 5))) + 
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
 # Test colors
@@
   scale_size("Prop male") +
+   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")
+qplot(name, readgap, data=student2012.sub.summary.gap, size=I(5), color=rsig) + 
+  xlab("") +
+  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
+  scale_y_continuous("Reading Score Gap", limits=c(-80, 30), breaks=seq(-80, 30, 5), labels=c(seq(80, 0, -5), seq(5, 30, 5))) + 
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="none")

### 2015-05-08 paper/review-progress.Rmd
added a file tht will contain the steps to do in revision

# Revision progress

The data is described very completely in the OECD site that is linked from the paper, and the data set is linked form that site so that the reader can pull a copy and examine it themselves. The $ figure comes from that site too, but I'll check it.

I'd be happy to change the plots as suggested by the first reviewer. And as you suggested the paper is too long, so some things need to be cut.

The idea of using clinical significance, or effect size, is a good one.

The second reviewer has a lot of criticism.

- What we did with the data was definitely not a fishing trip. Pretty much what we looked at we have reported here. There are many, many aspects of the data that could be examined, and so it is conceivable that we could be cherry picking the results. This wasn't the case.

- We don't mean to imply causality, this is just association. So we would want to change the language to ensure this is clear.

- Albania's data in all of the plots we made makes us suspicious about the collection happening there. It cannot be purely by chance to have no factors associated with math scores, completely different to every other country. We didn't specifically seek out Albania.

- I think that showing the top scores for each gender and country is an important point to make.

- We also certainly do not want to be considered offensive in the language about the reverse gender gap in some countries. We would want to change that language to make sure that it is more reasonable.

- Some of the comments made would push the article into being specifically for an advanced statistician, rather than a general reader. I'm not sure that is the purpose of Significance.

I am guessing that the second reviewer makes you less confident about the paper. I think that it is unfortunate that doing an exploratory graphical analysis of data often evokes this kind of response from some statisticians. The same criticisms can be levelled at model selection but typically isn't. But it does raise important points that probably could be addressed in the article: data fishing/cherry picking, data snooping, association is not causation. These are really important especially as more and more people have access to open data and analysis resources, to look at data themselves. One of the purposes of this paper, is to say, hey, here's a data set available that has some interesting things to say about education across. Looking at the data yourself can help digest what others report, and helps us all to be informed consumers of information.  These issues could be discussed as part of this article, and may be useful for a novice statistician to learn about. I think that we'd be happy to add a few paragraphs along these lines.

### 2015-05-08 paper/review-progress.Rmd
added luke's comments to revision progress file

@@
 I think that we'd be happy to add a few paragraphs along these lines.
+
+From Luke
+
+- The $200T figure only makes since as a cumulative figure (current world GDP is ~$50T) over many, many years.
+I suspect the OECD extrapolated out to infinity to get this number and the confidence interval is very wide.
+- The association between the number of books in the home has been well documented and is well known.
+Our results simply corroborate the known association between ownership of certain household items and test scores.
+- Also, the most critical reviewer seems very skeptical of survey statistics.
+The weighting scheme used accounts for non-response as is the case for any well-designed surveys.
+Does he/she want just want approximate percentages?
+- The OECD is only able to survey students.
+So women deprived of educational opportunities are not in the sampling frame.
+Our analysis indicates, that for Jordan and Qatar, there is evidence of girls/women outperforming boys in math and reading.
+- The visual plots seem like they would be much more informative regarding "scale" and "clinical differences" than raw numbers.

### 2015-05-08 paper/paper.Rmd
adjusts the studying-hours section
- adds standard error
- removed histogram

@@
 library(doBy)
+library("ggsubplot")
 ```
@@
 #study.luke <- melt(student.luke,id=c("name","ST57Q01"))
-studyplot <- ggplot(data = student.luke, aes(ST57Q01, PV1MATH)) + xlab("Hours spent out of school studying per week") + ylab("Math score") + facet_wrap(~name) + stat_smooth(se=F)
+studyplot <- 
+  ggplot(
+    data = student.luke, 
+    aes(x = ST57Q01, y = PV1MATH)
+  ) + 
+  xlab("Hours spent out of school studying per week") + 
+  ylab("Math score") + 
+  stat_smooth(se=TRUE) +
+  facet_wrap(~name)
+
 studyplot
@@
 ```{r studyingbar, fig.width=4, fig.height=4, echo = FALSE, message = FALSE, warning = FALSE}
-studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=5) + xlab("Hours spent studying")
-studybar
+# studybar <- ggplot(data = student.luke, aes(ST57Q01) ) + geom_histogram(binwidth=5) + xlab("Hours spent studying")
+# studybar
 ```

### 2015-05-08 paper/paper.Rmd
removed the shortages section

--- 2014-09-11 paper/paper.Rmd
@@
-Seven Things That Affect Education Around the World
+Six Things That Affect Education Around the World
 ========================================================
@@
 
-# 3. Shortages in teachers, computers, and libraries at school make a difference in math scores in a few countries
-
-Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction.
-The plot below examines three of these: lack of teachers, lack of computers, and shortage of library materials.
-The barcharts show the overall counts of shortages in schools.
-Teacher shortage is less comman than material shortages.
-
-These factors are compared with math scores from the student table, across countries.
-In most countries, scores don't change mcuh as shortages get worse.
-Scores decrease by 10-20 points onaverage in countries like Australia, China, Uruguay, Turkey, UAE, USA.
-Teacher shortages appear to have impact in a few countries like Greece, Slovakia, and Serbia.
-
-```{r shortencountrynames, warning=FALSE, message=FALSE, echo=FALSE, results='hide'}
-#school2012$name <- as.character(school2012$name)
-#student2012$name <- as.character(student2012$name)
-school2012 = shorten_countrynames(school2012)
-student2012 = shorten_countrynames(student2012)
-```
-
-```{r shortages, fig.width=4, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
-school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
-student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]
-
-scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
-  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))
-
-df = scores_by_school %>% left_join(school2012.sub)
-df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
-df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
-df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
-df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
-#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
-#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
-colnames(df2)[2] <- "levels"
-df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
-colnames(df3)[2] <- "levels"
-df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
-colnames(df4)[2] <- "levels"
-df.all <- merge(df2, df3)
-df.all <- merge(df.all, df4)
-df.m <- melt(df.all, id=c("name", "levels"))
-colnames(df.m)[4] <- "math"
-qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
-#ggsave("teacher-shortages.pdf", width=5, height=5)
-qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
-#ggsave("computer-shortages.pdf", width=5, height=5)
-qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
-#ggsave("library-shortages.pdf", width=5, height=5)
-#ggsave("shortages.pdf", width=14, height=14)
-```
-
-```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
-df$SC14Q02 <- as.character(df$SC14Q02)
-df$SC14Q07 <- as.character(df$SC14Q07)
-df$SC14Q10 <- as.character(df$SC14Q10)
-df$SC14Q02[df$SC14Q02==3] <- 2
-df$SC14Q07[df$SC14Q07==3] <- 2
-df$SC14Q10[df$SC14Q10==3] <- 2
-df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2))
-df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2))
-df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2))
-df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
-#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
-#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
-#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
-#      ylab="Math Score", geom=c("point","line"), group=name)
-df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
-colnames(df2)[2] <- "levels"
-df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
-colnames(df3)[2] <- "levels"
-df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
-colnames(df4)[2] <- "levels"
-df.all <- merge(df2, df3)
-df.all <- merge(df.all, df4)
-df.m <- melt(df.all, id=c("name", "levels"))
-colnames(df.m)[4] <- "math"
-qplot(levels, math, data = df.m, facets = ~name, 
-      xlab="Shortages (0=None, 1=Very little, 2=To some extent, or a lot)", 
-      ylab="Math score", geom=c("point","line"), group=variable, colour=variable) +
-      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
-      theme(legend.position="bottom") 
-```
-
-# 4. Time spent out of school studying is important, but only up to a point.
+# 3. Time spent out of school studying is important, but only up to a point.
 
@@
 
-# 5. Parents matter!
+# 4. Parents matter!
 ### a. Number of parents in the home
@@
 
-# 6. Albania is not like anyone else!
+# 5. Albania is not like anyone else!
 
@@
 
-# 7. More TVs yield higher math scores in the developing world, but lower in the developed countries.
+# 6. More TVs yield higher math scores in the developing world, but lower in the developed countries.
 

### 2015-05-08 paper/review-progress.Rmd
update to do
removed the shortages section

@@
 # Revision progress
+
+### TODO
+
+- Drop a plot or two (dropped shortages plot)
+- Make circles same size in dotplots, add error bars
+- Add histograms as subplots to facets for time spent studying (Not possible, se bands have been added)
+- paragraph on fishing (Di)
+- Clean language on middle east surprise
+- Age started vs years in school
+
+### My email to Brian Tarran
 
@@
 
-From Luke
+### From Luke
 

### 2015-05-09 paper/paper.Rmd
adds some words on sample size for hours-studied
made point size the same in dotplots

@@
 Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no significant difference).
-The size of the dot indicates the proportion of boys in the sample size.
 For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys.
@@
 
-Conversely, though, looking at reading scores, the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.
+Conversely, though, looking at reading scores, the story is entirely different. Universally, girls score significantly better than boys.
 
@@
       levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
-qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
+qplot(name, mathgap, data=student2012.sub.summary.gap, color=msig) + 
   xlab("") +  
@@
   scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5), limits=c(-30, 30)) + 
-  scale_size("Prop male") +
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
@@
       levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
-qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
+qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig) + 
   xlab("") +
@@
   scale_y_continuous("Reading Score Gap") + 
-  scale_size("Prop male") +
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")

### 2015-06-09 paper/paper.Rmd
started paragraph on the purpose of EDA

@@
 
-The vast majority of students spend less than 10 hours studying per week. The pattern  suggests decreasing benefits to increased out of school study time after 10 hours.
+The vast majority of students spend less than 10 hours studying per week.
+The pattern  suggests decreasing benefits to increased out of school study time after 10 hours.
+The thickness of the standard-error bars in the plot are an indication of the sample size.
 
@@
 
-# Summary
+# Discussion
+
+We have conducted what would be considered to be an exploratory analysis of the PISA data.
+There is a big difference between exploratory analyses and confirmatory analyses.
+When many people think of statistics they have in mind confirmatory methods, which involve deciding on a hypothesis, collecting the data, and doing a statistical test of significance.
+Exploratory data analysis is different, it focuses on the data at hand, and the goal is to let the data inform, to use various methods, primarily visual to find relationships and patterns and anomalies that may not have been anticipated ahead of the data collection.
+Exploratory analyses often lead to designed experiments where conditions are controlled, or specific data collection with all the controls to allow for hypothesis testing.
+Exploratory analysis is particularly important for today's big data society.
 

### 2015-06-09 paper/paper.Rmd
started paragraph on the purpose of EDA

@@
 Exploratory analysis is particularly important for today's big data society.
+Data is available everywhere, and typically collected because it is possible, without thought to testing particular hypotheses.
+
+The PISA data is a very large data set, with 500,000 students measured on more than 600 variables, with additional information on schools and home life.
+The selection of students into the study follows a carefully designed survey instrument, which means that each student also has a sample weight which reflects how their demographic is represented in the sample.
+These allow the inference of findings made from this sample of students to the entire student population in the country.
+Strictly speaking, for the PISA data the survey weights primarily apply to the test score components, covering different types of schools and ensuring good representation of both genders.
+There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented.
+The sampling design focuse don school types and gender, only.
+
+To organize an exploratory analysis, the best thing to do is to formulate several questions to answer, based on the description of the data.
+These will form the starting point for the analysis, but they might also lead to more questions being raised as the exploration ensues.
+For our analysis, these are the questions that we started with:
+
+- Is there a gender gap, like we hear about in the media?
+- Do more possessions, essentially higher socioeconomic status, lead to higher scores?
+- Is time spent studying outside school instrumental in higher scores?
+- Is there a houewife effect - that students with stay-at-home mothers score better?
+- What is the relationship between books and televisions and scores?
+- Does age when starting school influence scores?
+- Do parents occupations affect scores?
+
+Most of these were addressed in the analysis, with the exception of occupations because it turned out that there was too much missing data.
+
+Too many pursuits, will find something significant....
 

### 2015-06-09 paper/paper.Rmd
finished paragraph on the purpose of EDA

@@
 
-# 6. More TVs yield higher math scores in the developing world, but lower in the developed countries.
+# 6. On average, more TVs yield higher math scores in the developing world, but lower in the developed countries.
 
@@
 There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented.
-The sampling design focuse don school types and gender, only.
+The sampling design focused on school types and gender, only.
 
-To organize an exploratory analysis, the best thing to do is to formulate several questions to answer, based on the description of the data.
+To organize an exploratory analysis, start by formulating several questions to answer, based on the description of the data.
 These will form the starting point for the analysis, but they might also lead to more questions being raised as the exploration ensues.
@@
 - Is time spent studying outside school instrumental in higher scores?
-- Is there a houewife effect - that students with stay-at-home mothers score better?
+- Is there a housewife effect - that students with stay-at-home mothers score better?
 - What is the relationship between books and televisions and scores?
@@
 
-Most of these were addressed in the analysis, with the exception of occupations because it turned out that there was too much missing data.
+Most of these were answered in the analysis, with the exception of occupations because it turned out that there was too much missing data.
 
-Too many pursuits, will find something significant....
+Exploratory analysis, like confirmatory analysis, relies on probability, to some extent.
+If the sample data being examined is not representative of the population of interest then no inference is valid.
+If you search for many, many different associations it is likely that you will find something that looks important.
+This is the interplay of probability that helps determine whether what you see in the sample infers structure in the population.
 
-Overall, we have seen that there are many factors that affect the average math scores, but the major source of variation in the data is from individual to individual.
+The PISA data is a carefully designed survey, but the student and household questionaires are more loosely administered, yielding many missing values, and perhaps some questionable data.
+When results are surprising, and contradict current thinking, it can be helpful to research the problem in more details.
+For example, in follow up exploration of the PISA data during an introductory statistics class, we found that South Koreans do not use the internet as much as students in other countries.
+This is surprising, given the high tech nature of the society, and leads us to be skeptical about the data.
+There are several plausible reasons: (1) the mobile phone network dominates internet access more than internet providers, (2) individual identification on the questionaires might lead students not to answer honestly.
+If it is the latter, the problem could be fixed by re-designing the survey instrument.
+
+Our findings about test scores and other factors, documented in this article, some results that are consistent with current thinking, e.g.
+parents matter, socioeconomic status matters, time spent studying matters, but also some results that surprised us.
+These findings are what this data suggests, but they are only as good as the data.
+If the collection was not well done, and the collection is the responsibility of the local committee in each country, then the findings can not be trusted.
+An example of this is the data from Albania, which our analysis suggests that the PISA committee needs to re-assess what was done by that local committee, because the results from the questionaire are at odds with all other countries.
+
+Overall, our analysis yielded several interesting insights into factors that affect the average test scores.
+The data is rich, and should reliably reflect the larger population of 15 year olds.
+It is reassuring, though, that repeatedly we found that the major source of variation in the data is from individual to individual.
+For example, even though, on average boys perform better on the math tests in the USA, and girls on the reading tests, the top math score was achieved by a girl, and the top reading score was achieved by a boy.
 You as a person are more than your demographic -  you count, your individual efforts matter.

### 2015-06-09 paper/review-progress.Rmd
finished paragraph on the purpose of EDA

--- 2015-05-08 paper/review-progress.Rmd
@@
 # Revision progress
+
+### TODO
+
+- Drop a plot or two (dropped shortages plot) DONE
+- Make circles same size in dotplots, add error bars
+- Add histograms as subplots to facets for time spent studying (Not possible, se bands have been added) DONE
+- paragraph on fishing (Di) DONE
+- Clean language on middle east surprise
+- Age started vs years in school (DON'T PLAN TO DO ANYTHING ON THIS)
+
+### My email to Brian Tarran
 

### 2015-06-10 paper/paper.Rmd
Proposed new verbiage for middle-eastern discussion

@@
 It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.
-Five countries have gender gaps in favor of girls, and even more surprising, these are middle eastern or Muslim countries.
+
+Five countries have gender gaps in favor of girls; these are middle-eastern countries.
+Given popular perception of gender issues in the region, this may be a surprising result.
+Perhaps this is an example of how data can be used to show that reality is more nuanced than popular perception.
 

### 2015-06-17 paper/review-progress.Rmd
adds some DONE's to to-do list

--- 2015-05-08 paper/review-progress.Rmd
@@
 # Revision progress
+
+### TODO
+
+- Drop a plot or two (dropped shortages plot) DONE
+- Make circles same size in dotplots (DONE), add error bars 
+- Add histograms as subplots to facets for time spent studying (Not possible, se bands have been added) DONE
+- paragraph on fishing (Di) DONE
+- Clean language on middle east surprise DONE
+- Age started vs years in school (DON'T PLAN TO DO ANYTHING ON THIS)
+
+### My email to Brian Tarran
 

### 2015-07-25 paper/paper.Rmd
added bootstrap cis

@@
 
-```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+```{r load_packages, cache=FALSE, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
+library(dplyr)
 library(plyr)
-library(dplyr)
 library(stringr)
@@
 library(doBy)
-library("ggsubplot")
+library(boot)
 ```
 
-```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=TRUE}
+```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=FALSE}
 # This is the scaffolding to import the data
@@
 world.polys <- extractPolygons(world)
-detach("package:plyr")
+#detach("package:plyr")
 ```
 
-```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
+```{r dataprep, cache=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
 colnames(student2012)[1] <- "name"
@@
 ```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
+#library(plyr)
 student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
@@
 student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
+cifn <- function(d, i) {
+  x <- d[i,]
+  ci <- mean(x$PV1MATH[x$ST04Q01=="Male"], na.rm=T)-
+     mean(x$PV1MATH[x$ST04Q01=="Female"], na.rm=T)
+  ci
+}
+bootfn <- function(d) {
+  r <- boot(d, statistic=cifn, R=1000)
+  l <- sort(r$t)[25]
+  u <- sort(r$t)[975]
+  ci <- c(l, u)
+  return(ci)
+}
+#ddply(student2012.sub, .(name), bootfn)
+#results <- boot(data=aus, statistic=cifn, R=1000)
+student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
+colnames(student2012.sub.summary.gap.boot)[2:3] <- c("ml", "mu")
+student2012.sub.summary.gap <- merge(student2012.sub.summary.gap, student2012.sub.summary.gap.boot)
 student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
       levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
-qplot(name, mathgap, data=student2012.sub.summary.gap, color=msig) + 
+qplot(name, mathgap, data=student2012.sub.summary.gap, color=msig, size=I(3)) + 
+  geom_segment(aes(x=name, xend=name, y=ml, yend=mu)) +
   xlab("") +  
   scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
-  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5), limits=c(-30, 30)) + 
+  scale_y_continuous("Math Score Gap", breaks=seq(-45, 45, 5), limits=c(-45, 45)) + 
    geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
@@
 student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
+cifn <- function(d, i) {
+  x <- d[i,]
+  ci <- mean(x$PV1READ[x$ST04Q01=="Male"], na.rm=T)-
+     mean(x$PV1READ[x$ST04Q01=="Female"], na.rm=T)
+  ci
+}
+bootfn <- function(d) {
+  r <- boot(d, statistic=cifn, R=1000)
+  l <- sort(r$t)[25]
+  u <- sort(r$t)[975]
+  ci <- c(l, u)
+  return(ci)
+}
+#ddply(student2012.sub, .(name), bootfn)
+#results <- boot(data=aus, statistic=cifn, R=1000)
+student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
+colnames(student2012.sub.summary.gap.boot)[2:3] <- c("rl", "ru")
+student2012.sub.summary.gap <- merge(student2012.sub.summary.gap, student2012.sub.summary.gap.boot)
 student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
       levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
-qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig) + 
+qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig, size=I(3)) + 
+  geom_segment(aes(x=name, xend=name, y=rl, yend=ru)) +
   xlab("") +
@@
 
-```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r age, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+#library(dplyr)
+detach("package:plyr")
 student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
@@
 orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
-student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
+student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian, by="name")
 student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
@@
 
-```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
+```{r agemaps, cache=FALSE, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
 student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
@@
 
-```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r studying, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 student.luke <- student2012[,c("name","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
@@
 
-```{r par_athome, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r par_athome, cache=FALSE, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N"))
@@
 
-```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r parentsjob, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 studentLuke <- student2012
@@
 Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.
+This most likely indicates arrors in the data collected for the country.
 
-```{r clean, echo=FALSE, message = FALSE, warning = FALSE}
+```{r clean, cache=FALSE, echo=FALSE, message = FALSE, warning = FALSE}
 #student2012$CNT = student2012$name
@@
 
-```{r albania-possessions, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r albania-possessions, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 #df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("name", "numposs", "OECD"))
@@
 
-```{r truancyschool, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
+```{r truancyschool, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
 #school2012$name = school2012$name

### 2015-07-25 paper/paper.Rmd
fixing package conflicts

@@
 world.polys <- extractPolygons(world)
-#detach("package:plyr")
+detach("package:plyr")
 ```
@@
 These averages are differenced to compute the "gap" between boys and girls.
-Unadjusted t-statistics were calculated to determine the statistical significance of the difference.
+To gauge the statistical significance of the difference, t-statistics and bootstrap confidence intervals were calculated.
 The results are plotted in the graphs below.
-Each dot corresponds to a difference between girls' and boys' scores, for each country.
+Each dot corresponds to a difference between girls' and boys' scores, for each country, with lines giving the confidence interval.
+Wider intervals indicate less data support the statistics.
 Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no significant difference).
-For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys.
-The maps color the countries by the presence of a gender gap.
+The maps color the countries by the presence (or absence) of a gender gap.
 
@@
 #results <- boot(data=aus, statistic=cifn, R=1000)
+library(plyr)
 student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
+detach("package:plyr")
 colnames(student2012.sub.summary.gap.boot)[2:3] <- c("ml", "mu")
@@
 #results <- boot(data=aus, statistic=cifn, R=1000)
+library(plyr)
 student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
+detach("package:plyr")
 colnames(student2012.sub.summary.gap.boot)[2:3] <- c("rl", "ru")
@@
 #library(dplyr)
-detach("package:plyr")
+#detach("package:plyr")
 student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]

### 2015-07-25 paper/paper.Rmd
small things cleanup

@@
-Six Things That Affect Education Around the World
+What Can We Learn About Education Around the World from the PISA Data?
 ========================================================
@@
 Each dot corresponds to a difference between girls' and boys' scores, for each country, with lines giving the confidence interval.
-Wider intervals indicate less data support the statistics.
-Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no significant difference).
+Wider intervals indicate less data support the statistics, for example, Liechtenstein tests few students so there is more uncertainty about the mean.
+Color indicates statistical significance (blue in favor of boys, pink in favor of girls, and green means there was no difference).
 The maps color the countries by the presence (or absence) of a gender gap.
@@
 
-# 2. Starting school at an older age generates lower average math scores.
+# 2. Starting school at an older age is associated with lower average math scores.
 
@@
 The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy.
+Principles of reproducible research were used for the analysis and the markdown script is available from https://github.com/ijlyttle/isu_pisa/blob/master/paper/paper.Rmd.
 

### 2015-07-31 paper/paper.Rmd
few small fixes in the paper

@@
-What Can We Learn About Education Around the World from the PISA Data?
+---
+output: word_document
+---
+What Does the Data Collected During PISA Testing of Teenagers Tell Us About Education Around the World?
 ========================================================
@@
 The goal of the PISA survey is to assess the workforce readiness of 15-year old students.
-Nearly 500,000 students were tested across 65 countries and economies.
+Nearly 500,000 students were tested across 65 countries and 18,000 schools.
 Students were examined on how well they can apply the knowledge they learned in school to applications outside of school.
@@
 These averages are differenced to compute the "gap" between boys and girls.
-To gauge the statistical significance of the difference, t-statistics and bootstrap confidence intervals were calculated.
+To gauge the statistical significance of the difference, t-statistics and bootstrapped 95% confidence intervals were calculated.
 The results are plotted in the graphs below.
@@
 Most countries have a gender gap in math scores in favor of boys.
-Colombia has the biggest gender gap, about 30 points on average boys do better than girls.
+Colombia has the biggest gender gap, about 30 points, on average boys do better than girls.
 It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.
 
-Five countries have gender gaps in favor of girls; these are middle-eastern countries.
-Given popular perception of gender issues in the region, this may be a surprising result.
+Five countries have gender gaps in favor of girls; these are middle-eastern countries, Malaysia and Thailand.
+Given popular perception of gender issues in the Middle East, this may be a surprising result.
 Perhaps this is an example of how data can be used to show that reality is more nuanced than popular perception.
@@
 
-For each country, we have computed the average math score by age that the children started school.
+For each country, we have computed the average math score by age that the students started school.
 In addition, for each country we computed the median age that children in the study started school.
@@
 For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than any other age group.
-School start age matters.
-Children need to start school in the 4-6 age range to expect better math scores, on average.
+School start age appears to matter: with starting school in the 4-6 age range being associated with higher average math scores.
+There could be many other factors confounding this observation, most notably being that it is the more mathematically gifted students that start earlier, leading to a difference in the population characteristics of the different age (at school start) groups.
 
@@
 The pattern  suggests decreasing benefits to increased out of school study time after 10 hours.
-The thickness of the standard-error bars in the plot are an indication of the sample size.
+The thickness of the standard-error bars in the plot are an indication of the sample size, with thicker line indicating less certainty about the math value.
 
@@
 Looking at the number of parents in a household shows that students in two-parent households are, on average, performing better in math than students not in a two-parent households.
-The majority of students live in two-parent households.
-Of the single-parent households, a single mother is more common than a single father.
+The majority of students (97%) live in two-parent households.
+Of the single-parent households, a single mother is five times more common than a single father.
 Scores of students with just a mother at home tend to be higher than those with just a father at home.
@@
 #athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
-pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
-pbarhome
+#pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
+#pbarhome
 ```
@@
 stay-at-home, retired)”, and no response.
-The frequencies for these categories can be seen in the bar chart.
-The chart shows that fathers primarily have full-time jobs.
+Fathers primarily have full-time jobs.
 However, there are large numbers of mothers who work full-time, part-time, or are categorized as "Other".
@@
 Albania appears to be an outlier in many educational trends.
-Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.
-For example, Albania has one of the smallest gender gaps for math scores.
-Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.
-This most likely indicates arrors in the data collected for the country.
+Many socioeconomic and school factors that are associated with test scores in other nations are unrelated to Albania's test scores.
+For example, there is no difference in math scores based on household structure in Albania, or in terms of hours spent studying.
 
@@
 
+The lack of association where we would expect association is also evident in number of possessions.
 Generally, students with more possessions score higher on average.
@@
 
+These patterns, especially all together, would most likely indicate that there are some problems with the way the data was collected for Albania.
+
 # 6. On average, more TVs yield higher math scores in the developing world, but lower in the developed countries.
@@
 
+The PISA data is a very large data set, with 500,000 students measured on more than 600 variables, and additional information collected about school and home life.
+The selection of students into the study follows a carefully designed survey instrument, which means that each student also has a sample weight which reflects how their demographic is represented in the sample.
+These allow the inference of findings made from this sample of students to the entire student population in the country.
+Strictly speaking, for the PISA data the survey weights primarily apply to the test score components, covering different types of schools and ensuring good representation of both genders.
+There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented.
+The sampling design focused on school types and gender, only.
+
 We have conducted what would be considered to be an exploratory analysis of the PISA data.
@@
 Exploratory analysis is particularly important for today's big data society.
-Data is available everywhere, and typically collected because it is possible, without thought to testing particular hypotheses.
-
-The PISA data is a very large data set, with 500,000 students measured on more than 600 variables, with additional information on schools and home life.
-The selection of students into the study follows a carefully designed survey instrument, which means that each student also has a sample weight which reflects how their demographic is represented in the sample.
-These allow the inference of findings made from this sample of students to the entire student population in the country.
-Strictly speaking, for the PISA data the survey weights primarily apply to the test score components, covering different types of schools and ensuring good representation of both genders.
-There is no guarantee that all socioeconomic groups are represented equally, or that households with all sorts of possible possessions are represented.
-The sampling design focused on school types and gender, only.
+Data is available everywhere, and in many situations it is collected because it is easy to.
+Mining data, like we have done for the PISA survey data, can lead to constructing hypotheses that can be rigorously tested with new data.
 
@@
 For example, in follow up exploration of the PISA data during an introductory statistics class, we found that South Koreans do not use the internet as much as students in other countries.
-This is surprising, given the high tech nature of the society, and leads us to be skeptical about the data.
-There are several plausible reasons: (1) the mobile phone network dominates internet access more than internet providers, (2) individual identification on the questionaires might lead students not to answer honestly.
-If it is the latter, the problem could be fixed by re-designing the survey instrument.
+This is surprising, given the high tech nature of the society, and leads us to be skeptical about that part of the data.
+There are several plausible reasons: (1) the mobile phone network dominates internet access more than internet providers, and so internet usage is masked by phone usage, (2) individual identification on the questionaires might lead students not to answer honestly.
+Both  problems could be fixed by re-designing the survey instrument.
 
-Our findings about test scores and other factors, documented in this article, some results that are consistent with current thinking, e.g.
-parents matter, socioeconomic status matters, time spent studying matters, but also some results that surprised us.
+Our findings about test scores and other factors, documented in this article, some of the results are consistent with current thinking, e.g.
+parents matter, socioeconomic status matters, time spent studying matters, but also contained some results that surprised us.
 These findings are what this data suggests, but they are only as good as the data.
-If the collection was not well done, and the collection is the responsibility of the local committee in each country, then the findings can not be trusted.
+If the collection was not well done, and the collection is the responsibility of the local committee in each country, then the findings cannot be trusted.
 An example of this is the data from Albania, which our analysis suggests that the PISA committee needs to re-assess what was done by that local committee, because the results from the questionaire are at odds with all other countries.
@@
 The data is rich, and should reliably reflect the larger population of 15 year olds.
-It is reassuring, though, that repeatedly we found that the major source of variation in the data is from individual to individual.
+Average scores are not the individual experience.
+We found that the major source of variation in the data is from individual to individual.
 For example, even though, on average boys perform better on the math tests in the USA, and girls on the reading tests, the top math score was achieved by a girl, and the top reading score was achieved by a boy.
@@
 
-The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy.
+The following R packages were used to conduct the data exploration: shiny, plyr, dplyr, ggplot2, stringr, reshape2, lubridate, maps, ggmap, htmltools, rworldmap, grid, scales, doBy, boot.
 Principles of reproducible research were used for the analysis and the markdown script is available from https://github.com/ijlyttle/isu_pisa/blob/master/paper/paper.Rmd.
+By downloading the data from the OECD site, all of our results can be reproduced in the free, and open source, software R, particularly using the RStudio interface.
 
